### 2022-02-28 omicron-generation.tex
update

\documentclass[12pt]{article}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

\usepackage{graphicx}
\usepackage{xspace}
%\usepackage{adjustbox}

\newcommand{\comment}{\showcomment}
%% \newcommand{\comment}{\nocomment}

\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\bmb}[1]{\comment{blue}{BMB}{#1}}
\newcommand{\djde}[1]{\comment{red}{DJDE}{#1}}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc} % make sure fancy dashes etc. don't get dropped

\usepackage{lineno}
\linenumbers

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{unsrt}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\mathrm{c}}}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rres}{\Rx{\mathrm{res}}}
\newcommand{\Rinv}{\Rx{\mathrm{inv}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\Rintrinsic}{\ensuremath{{\mathcal R}_{\textrm{\tiny intrinsic}}}\xspace}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
\newcommand{\rres}{\rx{\mathrm{res}}}
\newcommand{\rinv}{\rx{\mathrm{inv}}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\usepackage{lettrine}

\newcommand{\dropcapfont}{\fontfamily{lmss}\bfseries\fontsize{26pt}{28pt}\selectfont}
\newcommand{\dropcap}[1]{\lettrine[lines=2,lraise=0.05,findent=0.1em, nindent=0em]{{\dropcapfont{#1}}}{}}

\begin{document}

\begin{flushleft}{
	\Large
	\textbf\newline{
		Epidemic dynamics (and network structures) affect differences in incubation-period and generation-interval distributions of SARS-CoV-2 variants
	}
}
\bigskip

\section*{Summary}

\begin{itemize}
  \item Neglecting growth rate differences can exaggerate or diminish the observed differences in epidemiological distributions (e.g., incubation period distribution and generation-interval distribution) of SARS-CoV-2 variants
  \item Accounting for growth rate differences yield similar incubation period distribution estimates for Delta and Omicron variants
  \item Despite similaries in the incubation period distributions, we generally estimate shorter mean generation intervals for the Omicron variant (2.7--3.9 days) than for the Delta variant (3.9--4.5 days).
  \item Higher reproduction number of the Omicron variant can lead to a faster depletion of susceptible contacts, esepcially within a household, resulting in shorter realized generation intervals. Simulations suggest that the estimated differences in the mean generation interval are consistent with differences driven by susceptible depletion effects in small networks.
\end{itemize}

\end{flushleft}

\pagebreak

\section{Introduction}

Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the ongoing COVID-19 pandemic.
Transmission advantages of invading variants are typically characterized by the ratios of reproduction numbers, $\Rinv/\Rres$, and the differences in growth rates, $\rinv-\rres$.
These two quantities are linked by the generation-interval distributions of the resident and invading variants.
For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical  $\Rinv = \Rres$.

There is currently a lack of generation-interval estimates of SARS-CoV-2 varaints, in part, due to difficulties in observing actual infection events.
Instead, many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or testing \citep{abbott2022test} of the infector and the infectee.
However, transmission-interval distributions are subject to dynamical biases and may not be an accurate proxy for the generation-interval distribution.
For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
Therefore, the observed differences (or the lack thereof) in transmission-interval distributions of the invading and resident variants can exaggerate or undermine true differences in their generation-interval distributions, depending on their growth rate differences.

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron varaints from Netherlands between 13 and 26 December 2021.
The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with and without S gene target failures, but they neglected growth rate differences in their inference.
We first discuss epidemiological context in Netherlands during the study period and provide corrected estimates for the incubation period.
We combine incubation-period estimates with serial-interval data to infer generation-interval distributions of the Delta and Omicron variants.

\section{Methods}

\subsection{Data}

We analyze time series of reported COVID-19 cases (\url{https://data.rivm.nl/covid-19/}) and proportions of SARS-CoV-2 variants detected (\url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}) from Netherlands between 29 November 2021 and 30 January 2022.
Both data are publicly available on the National Institute for Public Health and the Environment (RIVM) website.

Serial interval data are taken from the supplementary material of \cite{backer2021omicron}.
The data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates or symptom onset dates.
The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure, week of infectors' symptom onset date (week 50 and 51), and the type of transmission (within or between households).
For simplicity, we refer to transmission pairs with and without S gene target failure as Omicron and Delta transmission pairs, respectively.
See original article for details of data collection.

\subsection{Estimating epidemic growth rates}

To estimate the differences in growth rates of the Delta and Omicron variants, we first calculate the number of COVID-19 cases caused by each variant by multiplying the weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly, instead of daily, time series to remove changes in the numbers of cases caused testing patterns.
We then fit a generalized linear model to obtain smooth trajectories for case time series \citep{wood2001mgcv}.
More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood.
Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.

To obtain confidence intervals on the estimated time-varying growth rates, we generate 1000 parameter sets using estimated spline coefficients and the corresponding variance-covariance matrices as means and variance-covariance matrices of multivariate normal distributions.
We calculate time-varying growth rates from each parameter set and take 95\% equi-tailed quantiles to generate 95\% confidence limits.

\subsection{Estimating forward incubation-period distributions from backward incubation-period distributions}

\cite{backer2021omicron} estimated the incubation periods from 513 individuals (258 Omicron and 255 Delta cases), who showed symptom onsets between 1 December 2021 and 2 January 2022.
They used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring probability distributions of time of infection for each individual from their known exposure dates (and then taking the difference between symptom onset time and potential infection times). 
This method further assumes a uniform distribution over exposure intervals.

In practice, incubation periods (and any other epidemiological delays) can be measured in two ways.
The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
We expect this forward incubation-period distribution $f_I(\tau)$ to remain constant over the course of an epidemic and therefore provide accurate descriptions of incubation periods.
On the other hand, the backward incubation periods, measured from a cohort of individuals, who developed symptoms at the same time, are sensitive to epidemic growth.
In particular, when incidence of infection is growing exponentially, we are more likely to observe shorter backward incubation periods because there will be relatively more individuals who were infected recently.
Specifically, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
\begin{equation}
b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r\sigma) f_I(\sigma)  \dsigma},
\label{eq:backward}
\end{equation}
where $r$ represents the epidemic growth rate.
The method of \cite{backer2020incubation} effectively estimates the backward incubation-period distribution as it does not take growth rates into account.

To account for growth rate differences between the Delta and Omicron variants, we take the backward incubation-period distributions, estimated by \cite{backer2021omicron}, and calculate the corresponding forward incubation-period distributions by inverting \eref{backward}:
\begin{equation}
f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(r\sigma)  b_I(\sigma)\dsigma},
\label{eq:forward}
\end{equation}
where $b_I(\tau)$ is assumed to follow a Weibull distribution.
To further account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
First, we approximate the previously-inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parametrize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified Gamma distributions and estimate the corresponding forward distribution using \eref{forward}.
We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.

\subsection{Estimating forward generation-interval distributions from forward serial-interval distributions}

Dynamical biases in the serial-interval distributions are more complex because the serial interval depends on the incubation periods of the infector and the infectee as well as the generation interval between them.
For example, \cite{backer2021omicron} measured the forward serial-interval distributions from cohorts of infectors who developed symptoms during the same week.
In this case, the forward serial interval $\tau_s$ can be expressed in the form \citep{park2021forward}:
\begin{equation}
\tau_s =  - \tau_{i, 1} + \tau_{g} + \tau_{i, 2},
\label{eq:serial}
\end{equation}
where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), $\tau_{g|i,1}$ represents the forward generation interval, and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
When incidence of infection is growing exponentially, infectors are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean forward generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g}]$).

We further note that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} is also likely shorter than the mean forward generation interval $\mathbb E[\hat{\tau}_{g}]$ that we wish to estimate because infectors with shorter incubation periods (due to exponential growth) also transmit faster.
To illustrate this point, we write $f_{G|I}(\tau_g|\tau_{i,1})$ to represent forward generation-interval distribution conditional on the incubation period.
Then, we can see that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} depends on the exponential growth rate:
\begin{align}
\mathbb E[\tau_{g}] &= \int_0^\infty \int_0^\infty \sigma_g f_{G|I}(\sigma_g|\sigma_{i,1}) b_I(\sigma_{i,1}) \dd \sigma_g \sigma_{i, 1}\\
&= \frac{\int_0^\infty \int_0^\infty \sigma_g \exp(-r\sigma_{i,1})  f_{G|I}(\sigma_g|\sigma_{i,1}) f_I(\sigma_{i,1}) \dd \sigma_g \sigma_{i, 1}}{\int_0^\infty  \exp(-r\sigma) f_I(\sigma) \dsigma}
\end{align}
On the other hand, the mean forward generation interval of our interest does not depend on the exponential growth rate:
\begin{equation}
\mathbb E[\hat{\tau}_{g}] = \int_0^\infty \int_0^\infty \sigma_g f_{G|I}(\sigma_g|\sigma_{i,1}) f_I(\sigma_{i,1}) \dd \sigma_g \sigma_{i, 1}.\\
\end{equation}

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_epidemic.pdf}
\caption{
\textbf{Epidemic dynamics on the Delta and Omicron variants in Netherlands.}
(A) Daily numbers of reported COVID-19 cases in Netherlands (points).
The solid line represents the 7-day moving average.
Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
(B) Proportion of SARS-CoV-2 variants detected from Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
(C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth rate differences (purple).
Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
Growth rates are estimated by taking the derivative of the generalized additive model estimates.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_incubation.pdf}
\caption{
\textbf{Observed and corrected differences in incubation period distributions of Delta and Omicron variants.}
(A) Maximum likelihood estimates of the observed (backward) incubation periods of the Delta (black) and Omicron (orange) variants by \cite{backer2021omicron}.
(B) Corrected estimates of the forward incubation periods for different assumptions about the growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare.pdf}
\caption{
\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
(A) Observed forward serial intervals from household transmission pairs in Netherlands for the Delta (black points) and Omicron (orange trianlges) variants \citep{backer2021omicron}.
Serial intervals are calculated from infectors who developed symptoms between 13 and 19 December 2021.
The black sold line and the orange dashed lines represent the maximum likelihood fits for forward serial-interval distributions of Delta ($r=-0.05/\textrm{day}$) and Omicron ($r=0.15/\textrm{day}$) variants, respectively. 
Bivariate lognormal distributions with correlation coefficient of 0.75 are used to model relationships between incubation and generation intervals.
(B) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
(C--D) Maximum likelihood estimaes for the bivariate relationship between incubation and generation intervals of the Omicron (C) and Delta (D) variants.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_reproduction_advantage.pdf}
\caption{
\textbf{Estimated time-varying reproduction number advantages of the Omicron variant.}
The effective reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (Figure 1C) and taking the maximum likelihood estimates of the generation-interval distributions based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
Purple lines illustrate the ratio between the effective reproduction numbers of the Delta and Omicron variants.
Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_household.pdf}
\caption{
\textbf{Effects of network structures on realized generation intervals.}
We simulate transmission from 2000 fully connected networks with varying sizes (4--20), reproduction number ratios (1.5--2.5), and over-dispersion parameters ($k=0.1$--$10$) and compare mean generation intervals.
Egocentric simulations (solid lines) only consider onward transmission from a single, primary infector in each network.
Local simulations (dashed lines) allow infection to continue to spread within a network until infections can no longer spread (due to herd immunity, stochastic fadeouts, and/or a lack of susceptible individuals).
Lines represents mean across 2000 networks, and shaded regions represent corresponding 95\% confidence intervals.
The Delta variant (black) is assumed to have effective reproduction number of 0.8.
The Omicron variant (orange) is assumed to have effective reproduction number between 1.2 and 2.
Dotted horizontal lines represent the maximum likelihood estimates of the generation-interval distributions based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant.
}
\end{figure}

\pagebreak

\section*{Supplementary Materials}

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_he.pdf}
\caption{
\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
We model relationships between incubation and generation intervals using a shifted gamma distribution as suggested by \cite{he2020temporal} as a sensitivity analysis.
We obtain similar mean generation-interval estimates, but the predited bivariate relationships (C--D) are considerably different.
This model also gives overall worse fits to the observed serial-interval distribution.
See Figure 3 in the main text for details.
}
\end{figure}

\bibliography{omicron.bib}

\end{document}

### 2022-02-28 figure_epidemic.R

@@ -177,6 +177,8 @@ g4 <- ggplot(all_growth) +
     legend.title = element_blank()
   )
 
+all_growth %>% filter(type=="Delta")
+
 gtot <- ggarrange(g1, g2, g3, g4, nrow=2, labels=c("A", "B", "C", "D"), draw=FALSE)
 
 ggsave("figure_epidemic.pdf", gtot, width=12, height=8)

### 2022-02-28 figure_household.R

@@ -14,7 +14,8 @@ mean_sgtf <- exp(logmean_sgtf + logsd_sgtf^2/2)
 simulate_generation_household2 <- simulate_generation_household %>%
   mutate(
     k=paste0("k=", k),
-    ratio=paste0("Reproduction number ratio: ", ratio)
+    ratio=paste0("Reproduction number ratio: ", ratio),
+    type=factor(type, labels=c("Egocentric", "Local"))
   )
 
 g1 <- ggplot(simulate_generation_household2) +
@@ -22,7 +23,7 @@ g1 <- ggplot(simulate_generation_household2) +
   geom_hline(yintercept=mean_sgtf, lty = "dotted", col="orange") +
   geom_ribbon(aes(household, ymin=lwr, ymax=upr, fill=strain, col=strain, lty=type), alpha=0.1) +
   geom_line(aes(household, mean, col=strain, lty=type)) +
-  scale_x_continuous("Household sizes", breaks=1:5*4) +
+  scale_x_continuous("Network sizes", breaks=1:5*4) +
   scale_y_continuous("Mean generation interval (days)") +
   scale_color_manual("Variant", values=c("black", "orange")) +
   scale_fill_manual("Variant", values=c("black", "orange")) +

### 2022-02-28 figure_reproduction_advantage.R

@@ -75,13 +75,22 @@ gfit1 <- gam(log(delta_cases)~s(time, bs="cs"), data=variant4, method="REML")
 gfit2 <- gam(log(omicron_cases)~s(time, bs="cs"), data=variant4, method="REML")
 
 gfit1_p1 <- predict(gfit1, newdata = data.frame(time=seq(1, 10, by=1/14)), type = "lpmatrix")
+gfit1_p2 <- predict(gfit1, newdata = data.frame(time=seq(1, 10, by=1/14)+0.01), type = "lpmatrix")
 gfit2_p1 <- predict(gfit2, newdata = data.frame(time=seq(1, 10, by=1/14)), type = "lpmatrix")
+gfit2_p2 <- predict(gfit2, newdata = data.frame(time=seq(1, 10, by=1/14)+0.01), type = "lpmatrix")
+
+gfit1_Xp <- (gfit1_p2 - gfit1_p1) / 0.01/7
+gfit2_Xp <- (gfit2_p2 - gfit2_p1) / 0.01/7
 
 set.seed(101)
 gfit1_sim <- rmvnorm(nsample, mean = coef(gfit1), sigma = vcov(gfit1))
 gfit2_sim <- rmvnorm(nsample, mean = coef(gfit2), sigma = vcov(gfit2))
 lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_nsgtf_within)[1:2], sigma=vcov(fit_lognormal_base_50_nsgtf_within))
 lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_sgtf_within)[1:2], sigma=vcov(fit_lognormal_base_50_sgtf_within))
+
+gfit1_post <- apply(gfit1_sim, 1, function(x) {gfit1_Xp %*% x})
+gfit2_post <- apply(gfit2_sim, 1, function(x) {gfit2_Xp %*% x})
+advantage_post <- gfit2_post - gfit1_post
 
 advantagedata <- lapply(1:nsample, function(x) {
   i1 <- c(exp(gfit1_p1 %*% c(gfit1_sim[x,]))/14)
@@ -100,6 +109,7 @@ advantagedata <- lapply(1:nsample, function(x) {
     R_delta=R_delta,
     R_omicron=R_omicron,
     R_advantage=R_omicron/R_delta,
+    r_advantage=tail(advantage_post[,x], -n),
     sim=x,
     time=tail(seq(1, 10, by=1/14), -n)
   )
[…]

### 2022-02-28 fit_he_r.R

@@ -3,6 +3,7 @@ library(readxl)
 source("serialfun.R")
 source("fitfun.R")
 source("sample_incubation.R")
+load("fit_he_base.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -23,15 +24,15 @@ data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsg
 fit_he_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
-    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
+    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
   }, 0, 1000)[[1]]
   
   moment_1_nsgtf <- integrate(function(z) {
-    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
+    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
   }, 0, 1000)[[1]]
   
   moment_2_nsgtf <- integrate(function(z) {
-    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
+    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
   }, 0, 1000)[[1]]
   
   mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
@@ -41,8 +42,8 @@ fit_he_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
   
   ff <- fitfun_he(data=data_50_nsgtf_within, 
-                  inf_shape=5, 
-                  inf_mean=5, 
+                  inf_shape=coef(fit_he_base_50_nsgtf_within)[[1]], 
+                  inf_mean=coef(fit_he_base_50_nsgtf_within)[[2]], 
                   inf_shift=5,
                   logmean_inc=logmean_inc_nsgtf, 
[…]

### 2022-02-28 fit_lognormal_base.R

@@ -35,10 +35,4 @@ fit_lognormal_base_50_nsgtf_within <- fitfun_lognormal(data=data_50_nsgtf_within
                                                        rho=rho, 
                                                        r=r_nsgtf)
 
-#fit_lognormal_base_50_sgtf_within 
-
-#confint(fit_lognormal_base_50_sgtf_within, method="quad")
-
-#fit_lognormal_base_50_nsgtf_within
-
 save("fit_lognormal_base_50_sgtf_within", "fit_lognormal_base_50_nsgtf_within", file="fit_lognormal_base.rda")

### 2022-02-28 fit_lognormal_r_50_between.R

@@ -12,16 +12,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
+serialdata_50_sgtf_between <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="between")
 
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
+serialdata_50_nsgtf_between <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="between")
 
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+data_50_sgtf_between <- rep(serialdata_50_sgtf_between$serial, serialdata_50_sgtf_between$n)
+data_50_nsgtf_between <- rep(serialdata_50_nsgtf_between$serial, serialdata_50_nsgtf_between$n)
 
-fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_50_nsgtf_between <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
@@ -41,13 +41,13 @@ fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
   logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
   
-  ff <- fitfun_lognormal(data=data_50_nsgtf_within, 
+  ff <- fitfun_lognormal(data=data_50_nsgtf_between, 
                          logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
                          logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
-                         logmean_inc=logmean_inc_nsgtf, 
[…]

### 2022-02-28 fit_lognormal_r_51.R

@@ -1,3 +1,4 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
 source("serialfun.R")
@@ -12,16 +13,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
+serialdata_51_sgtf_within <- serialdata %>%
+  filter(week==51, strain=="SGTF", household=="within")
 
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
+serialdata_51_nsgtf_within <- serialdata %>%
+  filter(week==51, strain=="non-SGTF", household=="within")
 
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+data_51_sgtf_within <- rep(serialdata_51_sgtf_within$serial, serialdata_51_sgtf_within$n)
+data_51_nsgtf_within <- rep(serialdata_51_nsgtf_within$serial, serialdata_51_nsgtf_within$n)
 
-fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_51_nsgtf_within <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
@@ -41,13 +42,13 @@ fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
   logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
   
-  ff <- fitfun_lognormal(data=data_50_nsgtf_within, 
+  ff <- fitfun_lognormal(data=data_51_nsgtf_within, 
                          logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
                          logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
[…]

### 2022-02-28 fit_lognormal_r_51_between.R

@@ -12,16 +12,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
+serialdata_51_sgtf_between <- serialdata %>%
+  filter(week==51, strain=="SGTF", household=="between")
 
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
+serialdata_51_nsgtf_between <- serialdata %>%
+  filter(week==51, strain=="non-SGTF", household=="between")
 
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+data_51_sgtf_between <- rep(serialdata_51_sgtf_between$serial, serialdata_51_sgtf_between$n)
+data_51_nsgtf_between <- rep(serialdata_51_nsgtf_between$serial, serialdata_51_nsgtf_between$n)
 
-fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_51_nsgtf_between <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
@@ -41,13 +41,13 @@ fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
   logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
   
-  ff <- fitfun_lognormal(data=data_50_nsgtf_within, 
+  ff <- fitfun_lognormal(data=data_51_nsgtf_between, 
                          logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
                          logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
                          logmean_inc=logmean_inc_nsgtf, 
-                         logsd_inc=logsd_inc_nsgtf, 
-                         rho=rho, 
[…]

### 2022-02-28 fitfun.R

@@ -11,9 +11,11 @@ fitfun_lognormal_nll <- function(logmean_gen=1,
     return(Inf)
   }
   
-  ss <- serialfun_lognormal(data, logmean_gen=logmean_gen, logsd_gen=logsd_gen, 
+  ss <- try(serialfun_lognormal(data, logmean_gen=logmean_gen, logsd_gen=logsd_gen, 
                             logmean_inc=logmean_inc, 
-                            logsd_inc=logsd_inc, rho=rho, r=r, alphan=alphan)
+                            logsd_inc=logsd_inc, rho=rho, r=r, alphan=alphan))
+  
+  if (inherits(ss, "try-error")) return(Inf)
   
   if (debug) print(-sum(log(ss)))
   

### 2022-02-28 serialfun.R

@@ -48,24 +48,24 @@ serialfun_lognormal_numer <- function(x,
                                       logsd_inc=0.5,
                                       rho=0.5,
                                       r=0.1,
-                                      alpha1min=-20,
+                                      alpha1min=-30,
                                       alphan=0.5e2,
-                                      epsilon=1e-2) {
-  alpha1vec <- seq(alpha1min, min(-epsilon, x-epsilon), length.out=alphan)
+                                      epsilon=1e-3) {
+  alpha1vec <- head(seq(alpha1min, min(0, x), length.out=alphan)-epsilon, -1)
   
   ss <- sapply(alpha1vec, function(a1) {
-    alpha2vec <- seq(a1+epsilon, x, length.out=alphan)
+    alpha2vec <- seq(a1+epsilon, x-epsilon, length.out=alphan)
     
-    sum(head(serialfun_lognormal_internal(x, a1, alpha2vec, 
+    sum(serialfun_lognormal_internal(x, a1, alpha2vec, 
                                           logmean_gen=logmean_gen,
                                           logsd_gen=logsd_gen,
                                           logmean_inc=logmean_inc,
                                           logsd_inc=logsd_inc,
                                           rho=rho,
-                                          r=r), -1) * diff(alpha2vec))  
+                                          r=r) * diff(alpha2vec)[1])  
   })
   
-  sum(head(ss, -1) * diff(alpha1vec))
+  sum(ss * diff(alpha1vec)[1])
 }
 
 serialfun_lognormal <- function(x, 
@@ -75,10 +75,10 @@ serialfun_lognormal <- function(x,
                                 logsd_inc=0.5,
                                 rho=0.5,
                                 r=0.1,
-                                alpha1min=-40,
+                                alpha1min=-30,
                                 alphan=0.5e2,
                                 epsilon=1e-3,
-                                xmin=-20,
+                                xmin=-10,
                                 xmax=20,
                                 xby=0.2) {
   xvec <- seq(xmin, xmax, by=xby)
@@ -175,26 +175,27 @@ serialfun_he_numer <- function(x,
                                logmean_inc=1,
                                logsd_inc=0.5,
                                r=0.1,
-                               alpha1min=-20,
-                               alphan=1e2,
-                               epsilon=1e-2) {
[…]

### 2022-03-02 omicron-generation.tex
almost done

@@
 
-Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron varaints from Netherlands between 13 and 26 December 2021.
+Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron varaints from the Netherlands between 13 and 26 December 2021.
 The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with and without S gene target failures, but they neglected growth rate differences in their inference.
-We first discuss epidemiological context in Netherlands during the study period and provide corrected estimates for the incubation period.
+We first discuss epidemiological context in the Netherlands during the study period and provide corrected estimates for the incubation period.
 We combine incubation-period estimates with serial-interval data to infer generation-interval distributions of the Delta and Omicron variants.
@@
 
-We analyze time series of reported COVID-19 cases (\url{https://data.rivm.nl/covid-19/}) and proportions of SARS-CoV-2 variants detected (\url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}) from Netherlands between 29 November 2021 and 30 January 2022.
+We analyze time series of reported COVID-19 cases (\url{https://data.rivm.nl/covid-19/}) and proportions of SARS-CoV-2 variants detected (\url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}) from the Netherlands between 29 November 2021 and 30 January 2022.
 Both data are publicly available on the National Institute for Public Health and the Environment (RIVM) website.
@@
 \begin{equation}
-b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r\sigma) f_I(\sigma)  \dsigma},
+b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx},
 \label{eq:backward}
@@
 \begin{equation}
-f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(r\sigma)  b_I(\sigma)\dsigma},
+f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx)  b_I(x)\dx},
 \label{eq:forward}
@@
 \begin{align}
-\mathbb E[\tau_{g}] &= \int_0^\infty \int_0^\infty \sigma_g f_{G|I}(\sigma_g|\sigma_{i,1}) b_I(\sigma_{i,1}) \dd \sigma_g \sigma_{i, 1}\\
-&= \frac{\int_0^\infty \int_0^\infty \sigma_g \exp(-r\sigma_{i,1})  f_{G|I}(\sigma_g|\sigma_{i,1}) f_I(\sigma_{i,1}) \dd \sigma_g \sigma_{i, 1}}{\int_0^\infty  \exp(-r\sigma) f_I(\sigma) \dsigma}
+\mathbb E[\tau_{g}] &= \int_0^\infty \int_0^\infty x_g f_{G|I}(x_g|x_{i,1}) b_I(x_{i,1}) \dd x_g x_{i, 1}\\
+&= \frac{\int_0^\infty \int_0^\infty x_g \exp(-rx_{i,1})  f_{G|I}(x_g|x_{i,1}) f_I(x_{i,1}) \dd x_g x_{i, 1}}{\int_0^\infty  \exp(-r x) f_I(x) \dx}
 \end{align}
@@
 \begin{equation}
-\mathbb E[\hat{\tau}_{g}] = \int_0^\infty \int_0^\infty \sigma_g f_{G|I}(\sigma_g|\sigma_{i,1}) f_I(\sigma_{i,1}) \dd \sigma_g \sigma_{i, 1}.\\
+\mathbb E[\hat{\tau}_{g}] = \int_0^\infty \int_0^\infty x_g f_{G|I}(x_g|x_{i,1}) f_I(x_{i,1}) \dd x_g x_{i, 1}.\\
+\label{eq:meangen}
 \end{equation}
+Therefore, while we generally expect the mean forward generation interval to be shorter than the mean forward serial interval for a growing epidemic, correlation between symptom onset and transmissibility can also have opposite effects.
 
-\begin{figure}[!th]
+Here, we model forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions with lognormal distributions.
+Both distributions are parameterized by log means, $\mu_I$ and $\mu_G$, and log variances, $\sigma_I^2$ and $\sigma_G^2$.
+Then, given the correlation coefficient $\rho$, the forward generation-interval distribution conditional conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
+Assuming that the incidence of infection is changing exponentially at rate $r$, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be calculated by integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ \citep{park2021forward}:
+\begin{equation}
+f_S(\tau) = \frac{1}{\phi} \int_{-\infty}^0\int_{\alpha_1}^\tau \exp(r \alpha_1) f_{G|I}(\alpha_2 - \alpha_1|- \alpha_1) f_I(- \alpha_1) f_I(\tau - \alpha_2) \dd \alpha_2 \dd \alpha_1,
+\label{eq:forwardserial}
+\end{equation}
+where $\phi$ represents the normalization constant to satisfy $\int f_S(x) \dx = 1$;
+$-\alpha_1$ corresponds to the incubation period of the infector;
+$\alpha_2 - \alpha_1$ corresponds to the generation interval;
+and $\tau - \alpha_2$ corresponds to the incubation period of the infectee.
+
+For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
+We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation.
+Using this incubation-period distribution, we fit \fref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
+We then calculate the mean forward generation interval using \eref{meangen} and approximate the 95\% confidence interval using the Delta method.
+We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be practically unidentifiable.
+
+\subsection{Estimating instantaneous reproduction number}
+
+We use our estimates of the generation-interval distributions to infer instantaneous reproduction numbers $\mathcal R(t)$ of the Delta and Omicron variant, as well as the ratio between two reproduction numbers.
+Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
+\begin{equation}
+\mathcal R(t) = \frac{i(t)}{\int_0^\infty i(t-x) g(x) \dx},
+\end{equation}
+where $i(t)$ represents incidence of infection.
+Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for week 50---when epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
+Then, the instantaneous reproduction number is estimated using the renewal equation \citep{fraser2007estimating}:
+Incidence of infection is approximated by shifting the smoothed case trajectories by one week to account for reporting delays.
+Confidence intervals are calculated by sampling parameters of the smoothed case trajectories as well as the generation-interval distributions from multivariate normal distributions and repeating the analysis 1000 times.
+
+\section{Results}
+
+\fref{epidemic} summarizes epidemiological context in the Netherlands during the study period.
+The first Omicron cases was detected in the Netherlands from a sample collected on 19 November 2021 \citep{backer2021omicron}, during which the number of COVID-19 cases were decreasing (\fref{epidemic}A).
+As the Omicron variant continued to spread and increase in proportion (\fref{epidemic}B), the number of COVID-19 cases  increased (\fref{epidemic}A).
+Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to tease apart differences in their epidemiological dynamics (\fref{epidemic}C).
+The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r = -0.01/\mathrm{day}$ (95\% CI: $-0.04/\mathrm{day}$--$0.02/\mathrm{day}$) to $r = -0.09/\mathrm{day}$ (95\% CI: $-0.11/\mathrm{day}$--$-0.07/\mathrm{day}$) by the week of January 16, 2022, and increasing back up to $r = -0.04/\mathrm{day}$ (95\% CI: $-0.07/\mathrm{day}$--$-0.01/\mathrm{day}$).
+The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ (95\% CI: $0.15/\mathrm{day}$--$0.21/\mathrm{day}$) on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ (95\% CI: $0.00/\mathrm{day}$--$0.09/\mathrm{day}$).
+Finally, we find that the growth rate difference between the Delta and Omicron variants decreased over time---this finding echoes our previous work, which showed that growth rate differences change over the course of an epidemic even when the ratio of the reproduction numbers remain constant \citep{park2021roles}.
+
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{figure_epidemic.pdf}
 \caption{
-\textbf{Epidemic dynamics on the Delta and Omicron variants in Netherlands.}
-(A) Daily numbers of reported COVID-19 cases in Netherlands (points).
+\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands.}
+(A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
 The solid line represents the 7-day moving average.
 Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
-(B) Proportion of SARS-CoV-2 variants detected from Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
+(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
 (C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
@@
 Growth rates are estimated by taking the derivative of the generalized additive model estimates.
+\label{fig:epidemic}
 }
@@
 
-\pagebreak
+For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} previoulsy found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
+However, during this period, the number of COVID-19 cases caused by the Delta and Omicron variants showed opposite patterns of epidemic growth/decay (\fref{epidemic}).
+Since the number of COVID-19 cases caused by the Delta variant was decreasing, longer incubation periods would have been more likely to be observed.
+In contrast, the rapid growth in the number of COVID-19 cases caused by the Omicron variant would have caused shorter incubation periods to be more likely to be observed.
+When we account for these growth rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
+Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C).
+For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$.
+The mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.3--4.4 days) to 4.5 days (95\% CI: 3.8--5.6 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.
 
@@
 \textbf{Observed and corrected differences in incubation period distributions of Delta and Omicron variants.}
-(A) Maximum likelihood estimates of the observed (backward) incubation periods of the Delta (black) and Omicron (orange) variants by \cite{backer2021omicron}.
-(B) Corrected estimates of the forward incubation periods for different assumptions about the growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
+(A) Posterior median estimates of the observed (backward) incubation periods of the Delta (black) and Omicron (orange) variants by \cite{backer2021omicron}.
+(B) Forward incubation-period distributions assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta (black) and Omicron (orange) variants, respectively.
+(C) Corrected estimates of the forward incubation-period distributions for different assumptions about the growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
 Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
+\label{fig:incubation}
 }
@@
 
-\pagebreak
+Our corrected estimates of the forward incubation period distributions further allow us to infer the forward generation-interval distributions.
+For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021).
+\citep{backer2021omicron} previously reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.5 days) for this cohort, but the overall shapes of the distributions are similar (\fref{serial}A).
+However, when we account for growth rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
+Notably, both variants have similar mode, but the transmissibility of the omicron variant decays much faster (\fref{serial}B).
+
+While we estimate shorter mean forward generation intervals for the Omicron variant across all stratifications (\fref{serial}C), the differences in the mean generation intervals are unclear for between-household pairs.
+Consistent with the original study, which also repoted shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
+We also estimate a reduction in the mean forward generation intervals from week 50 to week 51.
 
@@
 \textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
-(A) Observed forward serial intervals from household transmission pairs in Netherlands for the Delta (black points) and Omicron (orange trianlges) variants \citep{backer2021omicron}.
-Serial intervals are calculated from infectors who developed symptoms between 13 and 19 December 2021.
-The black sold line and the orange dashed lines represent the maximum likelihood fits for forward serial-interval distributions of Delta ($r=-0.05/\textrm{day}$) and Omicron ($r=0.15/\textrm{day}$) variants, respectively.
+(A) Observed and fitted forward serial-interval distributions fpr within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
+Serial intervals are calculated for infectors who developed symptoms on week 50 (13 and 19 December, 2021).
+Points represent the observed data.
+Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant. 
 Bivariate lognormal distributions with correlation coefficient of 0.75 are used to model relationships between incubation and generation intervals.
-(B) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
-Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
-(C--D) Maximum likelihood estimaes for the bivariate relationship between incubation and generation intervals of the Omicron (C) and Delta (D) variants.
+(B) Estimated forward generation-interval distributions for within-household transmission pairs in the Netherlands.
+(C) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines) stratified by the week of infectors' symptom onset and the type of transmission pairs.
+\label{fig:serial}
 }
@@
 
-\pagebreak
+Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.83 (95\% CI: 1.65--2.05) to 1.17 (95\% CI: 1.00--1.36) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
+On the other hand, the reproduction number of the Delta variant decreased from 0.89 (95\% CI: 0.82--0.97) to 0.68 (95\% CI: 0.62--0.74) between December 5, 2021, and January 9, 2022, and increased back up to 0.82 (95\% CI: 0.72--0.93) by January 23, 2022 (\fref{reproduction}A).
+We estimate the reproduction number ratios stayed rouhgly constant at around 2.2 (95\% CI: 2.0--2.5) between December 12--26, 2021, and slowly decreased to 1.4 (95\% CI: 1.2--1.7).
+However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
+In this case, the reproduction ratio decreases from 2.5 (95\% CI: 2.2--2.8) to 1.5 (95\% CI: 1.2--1.8), corresponding to roughly 7--13\% bias.
 
@@
 \textbf{Estimated time-varying reproduction number advantages of the Omicron variant.}
-The effective reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (Figure 1C) and taking the maximum likelihood estimates of the generation-interval distributions based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
-Purple lines illustrate the ratio between the effective reproduction numbers of the Delta and Omicron variants.
+(A) Estimated instantaneous reproduction numbers and their ratios over time while accounting for differences in the generation-interval distributions.
+(B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
+(C--D) Phase diagrams of the reproduction ratio against instantaneous reproduction numbers of the Delta (C) and Omicron (D) variants.
+The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
+The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the generation-interval distributions for within-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
+Purple lines represent the ratio between the effective reproduction numbers of the Delta and Omicron variants.
 Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
+\label{fig:reproduction}
 }
@@
 
-\pagebreak
+Comparing phase diagrams of the reproduction ratio against instantaneous reproduction numbers of the Delta (\fref{reproduction}C) and Omicron (\fref{reproduction}D) variants helps us better understand the reduction in transmission advantage over time.
+As the reproduction number of the Delta variant decreases, the reproduction number ratio initially increases and stays constant. 
+Then, the reproduction number ratio and the reproduction number of the Delta variant both decreases at the same time, and the reproduction number of the Delta variant increases slightly near the end (\fref{reproduction}C).
+Similarly, the decrease in the reproduction number ratio coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant (\fref{reproduction}D);
+larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the growth rate difference (\fref{epidemic}D).
 
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_household.pdf}
-\caption{
-\textbf{Effects of network structures on realized generation intervals.}
-We simulate transmission from 2000 fully connected networks with varying sizes (4--20), reproduction number ratios (1.5--2.5), and over-dispersion parameters ($k=0.1$--$10$) and compare mean generation intervals.
-Egocentric simulations (solid lines) only consider onward transmission from a single, primary infector in each network.
-Local simulations (dashed lines) allow infection to continue to spread within a network until infections can no longer spread (due to herd immunity, stochastic fadeouts, and/or a lack of susceptible individuals).
-Lines represents mean across 2000 networks, and shaded regions represent corresponding 95\% confidence intervals.
-The Delta variant (black) is assumed to have effective reproduction number of 0.8.
-The Omicron variant (orange) is assumed to have effective reproduction number between 1.2 and 2.
-Dotted horizontal lines represent the maximum likelihood estimates of the generation-interval distributions based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant.
-}
-\end{figure}
+\section{Discussion}
+
+We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands.
+The original analysis detailing the data set previously reported shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
+Accounting for differences in epidemic growth rates, we find similar incubation-period distributions for both variants but a considerably shorter (0.5--0.9 days) mean generation interval for the Omicron variant.
+The mean forward generation interval of both variants further decreased between week 50 (13--19 December, 2021) and week 51 (20--26 December, 2021).
+Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.2-fold to 1.4-fold between early December and late January. 
+Neglecting differences in the generation-interval distributions can result in $\approx 10\%$ bias in estimates of the transmission advantage.
+
+The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides cruicial information for epidemic control.
 

### 2022-03-02 figure_compare.R

@@ -2,10 +2,14 @@ library(dplyr)
 library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(egg)
+library(gridExtra)
 source("serialfun.R")
 load("calculate_incubation_mle.rda")
 load("fit_lognormal_base.rda")
 load("fit_lognormal_r.rda")
+load("fit_lognormal_r_50_between.rda")
+load("fit_lognormal_r_51_between.rda")
+load("fit_lognormal_r_51.rda")
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
@@ -65,17 +69,45 @@ g1 <- ggplot(serialdata_50_nsgtf_within) +
   geom_line(data=serial_density_lognormal_nsgtf, aes(x, y, col="Delta", lty="Delta"), lwd=1) +
   geom_line(data=serial_density_lognormal_sgtf, aes(x, y, col="Omicron", lty="Omicron"), lwd=1) +
   scale_x_continuous("Forward serial intervals (days)", expand=c(0, 0), limits=c(-6, 16)) +
-  scale_y_continuous("Density", expand=c(0, 0), limits=c(0, 0.22)) +
+  scale_y_continuous("Density (1/day)", expand=c(0, 0), limits=c(0, 0.22)) +
   scale_color_manual("", values=c("black", "orange")) +
   scale_linetype_discrete("") +
   scale_shape_discrete("") +
   theme(
     panel.grid = element_blank(),
-    legend.position = c(0.8, 0.8),
+    legend.position = c(0.8, 0.85),
     legend.background = element_rect(fill=NA)
   )
 
-g2 <- ggplot(filter(fit_lognormal_r_50_nsgtf_within, param=="mean")) +
+genden_nsgtf <- tibble(
+  time=seq(0, 14, by=0.1),
+  density=dlnorm(time, filter(fit_lognormal_r_50_nsgtf_within, r==-0.05, param=="logmean")$est, filter(fit_lognormal_r_50_nsgtf_within, r==-0.05, param=="logsd")$est)
+)
+
+genden_sgtf <- tibble(
+  time=seq(0, 14, by=0.1),
+  density=dlnorm(time, filter(fit_lognormal_r_50_sgtf_within, round(r, 2)==0.15, param=="logmean")$est, filter(fit_lognormal_r_50_sgtf_within, round(r, 2)==0.15, param=="logsd")$est)
+)
+
[…]

### 2022-03-02 figure_incubation.R

@@ -6,6 +6,9 @@ source("sample_incubation.R")
 rvec_nsgtf <- seq(-0.1, 0, length.out=11)
 rvec_sgtf <- seq(0.1, 0.2, length.out=11)
 
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+
 corrected_mean_nsgtf <- lapply(1:length(rvec_nsgtf), function(x) {
   numer <- sapply(1:nsample, function(y) {
     integrate(function(z) {
@@ -54,14 +57,26 @@ corrected_mean_sgtf <- lapply(1:length(rvec_sgtf), function(x) {
 }) %>%
   bind_rows
 
-observed_nsgtf <- data.frame(
+observed_nsgtf <- tibble(
   time=seq(0, 14, by=0.1),
-  density=dweibull(seq(0, 14, by=0.1), shape=backward_shape_nsgtf, scale=backward_scale_nsgtf)
+  density=dweibull(time, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf)
 )
 
-observed_sgtf <- data.frame(
+observed_sgtf <- tibble(
   time=seq(0, 14, by=0.1),
-  density=dweibull(seq(0, 14, by=0.1), shape=backward_shape_sgtf, scale=backward_scale_sgtf)
+  density=dweibull(time, shape=backward_shape_sgtf, scale=backward_scale_sgtf)
+)
+
+corrected_nsgtf <- tibble(
+  time=seq(0, 28, by=0.1),
+  density0=dweibull(time, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * time),
+  density=density0/sum(density0*0.1)
+)
+
+corrected_sgtf <- tibble(
+  time=seq(0, 28, by=0.1),
+  density0=dweibull(time, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * time),
+  density=density0/sum(density0*0.1)
 )
 
 g1 <- ggplot(observed_nsgtf) +
@@ -76,7 +91,18 @@ g1 <- ggplot(observed_nsgtf) +
     legend.title = element_blank()
   )
 
-g2 <- ggplot(corrected_mean_nsgtf) +
+g2 <- ggplot(corrected_nsgtf) +
+  geom_line(aes(time, density, col="Delta"), lwd=1) +
+  geom_line(data=corrected_sgtf, aes(time, density, col="Omicron"), lwd=1) +
[…]

### 2022-03-02 figure_reproduction_advantage_between.R

@@ -8,7 +8,7 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(mgcv)
 library(mvtnorm)
 library(egg)
-load("fit_lognormal_base.rda")
+load("fit_lognormal_base_50_between.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 
@@ -85,14 +85,15 @@ gfit2_Xp <- (gfit2_p2 - gfit2_p1) / 0.01/7
 set.seed(101)
 gfit1_sim <- rmvnorm(nsample, mean = coef(gfit1), sigma = vcov(gfit1))
 gfit2_sim <- rmvnorm(nsample, mean = coef(gfit2), sigma = vcov(gfit2))
-lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_nsgtf_within)[1:2], sigma=vcov(fit_lognormal_base_50_nsgtf_within))
-lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_sgtf_within)[1:2], sigma=vcov(fit_lognormal_base_50_sgtf_within))
+lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_nsgtf_between)[1:2], sigma=vcov(fit_lognormal_base_50_nsgtf_between))
+lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_sgtf_between)[1:2], sigma=vcov(fit_lognormal_base_50_sgtf_between))
 
 gfit1_post <- apply(gfit1_sim, 1, function(x) {gfit1_Xp %*% x})
 gfit2_post <- apply(gfit2_sim, 1, function(x) {gfit2_Xp %*% x})
 advantage_post <- gfit2_post - gfit1_post
 
 advantagedata <- lapply(1:nsample, function(x) {
+  if (lognormal_sim2[x,2] < 0) return(NULL)
   i1 <- c(exp(gfit1_p1 %*% c(gfit1_sim[x,]))/14)
   i2 <- c(exp(gfit2_p1 %*% c(gfit2_sim[x,]))/14)
   
@@ -185,4 +186,4 @@ g4 <- ggplot(advantagesumm2) +
 
 gfinal <- ggarrange(g1, g2, g3, g4, nrow=2, labels=c("A", "B", "C", "D"), draw=FALSE)
 
[…]

### 2022-03-02 fit_lognormal_base_50_between.R

@@ -1,0 +1,38 @@
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+load("calculate_incubation_mle.rda")
+
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+rho <- 0.75
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_50_sgtf_between <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="between")
+
+serialdata_50_nsgtf_between <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="between")
+
+data_50_sgtf_between <- rep(serialdata_50_sgtf_between$serial, serialdata_50_sgtf_between$n)
+data_50_nsgtf_between <- rep(serialdata_50_nsgtf_between$serial, serialdata_50_nsgtf_between$n)
+
+fit_lognormal_base_50_sgtf_between <- fitfun_lognormal(data=data_50_sgtf_between, 
+                                                      logmean_gen=logmean_inc_sgtf, 
+                                                      logsd_gen=logsd_inc_sgtf, 
+                                                      logmean_inc=logmean_inc_sgtf, 
+                                                      logsd_inc=logsd_inc_sgtf, 
+                                                      rho=rho, 
+                                                      r=r_sgtf)
+
+fit_lognormal_base_50_nsgtf_between <- fitfun_lognormal(data=data_50_nsgtf_between, 
+                                                       logmean_gen=logmean_inc_nsgtf, 
+                                                       logsd_gen=logsd_inc_nsgtf, 
+                                                       logmean_inc=logmean_inc_nsgtf, 
+                                                       logsd_inc=logsd_inc_nsgtf, 
+                                                       rho=rho, 
+                                                       r=r_nsgtf)
+
+save("fit_lognormal_base_50_sgtf_between", "fit_lognormal_base_50_nsgtf_between", file="fit_lognormal_base_50_between.rda")

### 2022-03-02 figure_compare_he.R
adding rdaout folder for Jonathan

@@ -3,9 +3,9 @@ library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(egg)
 source("serialfun.R")
-load("calculate_incubation_mle.rda")
-load("fit_he_r.rda")
-load("fit_he_base.rda")
+load("rdaout/calculate_incubation_mle.rda")
+load("rdaout/fit_he_r.rda")
+load("rdaout/fit_he_base.rda")
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 

### 2022-03-02 figure_reproduction_advantage.R

@@ -8,7 +8,7 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(mgcv)
 library(mvtnorm)
 library(egg)
-load("fit_lognormal_base.rda")
+load("rdaout/fit_lognormal_base.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 

### 2022-03-02 figure_reproduction_advantage_between.R

@@ -8,7 +8,7 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(mgcv)
 library(mvtnorm)
 library(egg)
-load("fit_lognormal_base_50_between.rda")
+load("rdaout/fit_lognormal_base_50_between.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 

### 2022-03-03 omicron-generation.tex
done

@@
 	\textbf\newline{
-		Epidemic dynamics (and network structures) affect differences in incubation-period and generation-interval distributions of SARS-CoV-2 variants
+		Epidemic dynamics shape inferences of the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants
 	}
@@
 
-\section*{Summary}
+\section*{Abstract}
 
-\begin{itemize}
-  \item Neglecting growth rate differences can exaggerate or diminish the observed differences in epidemiological distributions (e.g., incubation period distribution and generation-interval distribution) of SARS-CoV-2 variants
-  \item Accounting for growth rate differences yield similar incubation period distribution estimates for Delta and Omicron variants
-  \item Despite similaries in the incubation period distributions, we generally estimate shorter mean generation intervals for the Omicron variant (2.7--3.9 days) than for the Delta variant (3.9--4.5 days).
-  \item Higher reproduction number of the Omicron variant can lead to a faster depletion of susceptible contacts, esepcially within a household, resulting in shorter realized generation intervals.
-Simulations suggest that the estimated differences in the mean generation interval are consistent with differences driven by susceptible depletion effects in small networks.
-\end{itemize}
+Estimating the differences in the incubation-period and serial-interval distributions of SARS-CoV-2 variants is critical to understanding changes in their transmission dynamics and therefore controlling their outbreaks.
+However, these comparisons are inherently difficult due to their dependence on epidemic growth rates---for example, when an epidemic is growing exponentially, we are more likely to observe individuals who have been infected recently, decreasing the mean observed incubation period and thereby increasing the mean observed serial interval.
+Here, we analyze serial-interval data describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021, during which the number of infections caused by the Delta variant decreased as the number of Omicron infections increased.
+Accounting for growth rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
+These differences may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
+Neglecting differences in the generation-interval distributions can further bias in estimates of transmission advantages of the Omicron variant.
 
@@
 The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides cruicial information for epidemic control.
+A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
+Here, we build on our previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
+Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
+These effects are also pertinent to all epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
+
+To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant.
+Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
+We tentatively hypothesize that these differences may be primarily driven by the network effect: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections \citep{park2020inferring,hart2022generation}.
+While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structres that involve repeated contacts between the same group of individuals (beacuse only the first infectious contact results in infection).
+The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
+Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shoter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.
+
+While our study indicate that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how much their infectiousness profiles differ intrinsically.
+In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their intrinsic infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
+Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accouting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation interval distribution, which may have a longer mean).
+
+Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
+In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction number ratios; however, the degree of ratio is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
+For example, \cite{pearson2021bounding} estimated a much higher reproduction ratios ($> 4$ fold) in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
+With our generation-interval estimates, we estimate that the reproduction number ratio of 2.8 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+We also show that both growth rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.
+
+We primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
+In doing so, we implicitly assume thta the delay between infection and reports is fixed.
+However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
+While this limitation does not affect our generation-interval estimates, our inferences of the transmission advantages of the Omicron variant should be interpreted with care.
+
+Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
+Our study sythesizes previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters;
+however, uncertainty remains in the intrinsic infectiousness profiles of SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.
+Future studies extending our framework and combining more detailed epidemiological data will be critical to narrowing down these uncertainties.
 
 \pagebreak
-
-\section*{Supplementary Materials}
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_compare_he.pdf}
-\caption{
-\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
-We model relationships between incubation and generation intervals using a shifted gamma distribution as suggested by \cite{he2020temporal} as a sensitivity analysis.
-We obtain similar mean generation-interval estimates, but the predited bivariate relationships (C--D) are considerably different.
-This model also gives overall worse fits to the observed serial-interval distribution.
-See Figure 3 in the main text for details.
-}
-\end{figure}
 

### 2022-03-03 omicron-generation.tex
spelling check

@@
 
-There is currently a lack of generation-interval estimates of SARS-CoV-2 varaints, in part, due to difficulties in observing actual infection events.
+There is currently a lack of generation-interval estimates of SARS-CoV-2 variants, in part, due to difficulties in observing actual infection events.
 Instead, many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or testing \citep{abbott2022test} of the infector and the infectee.
@@
 
-Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron varaints from the Netherlands between 13 and 26 December 2021.
+Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
 The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with and without S gene target failures, but they neglected growth rate differences in their inference.
@@
 To further account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously-inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parametrize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
+First, we approximate the previously-inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parameterize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
 We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified Gamma distributions and estimate the corresponding forward distribution using \eref{forward}.
@@
 
-For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} previoulsy found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
+For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} previously found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
 However, during this period, the number of COVID-19 cases caused by the Delta and Omicron variants showed opposite patterns of epidemic growth/decay (\fref{epidemic}).
@@
 
-While we estimate shorter mean forward generation intervals for the Omicron variant across all stratifications (\fref{serial}C), the differences in the mean generation intervals are unclear for between-household pairs.
-Consistent with the original study, which also repoted shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
+While we estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), the differences in the mean generation intervals are unclear for between-household pairs.
+Consistent with the original study, which also reported shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
 We also estimate a reduction in the mean forward generation intervals from week 50 to week 51.
@@
 \textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
-(A) Observed and fitted forward serial-interval distributions fpr within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
+(A) Observed and fitted forward serial-interval distributions for within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
 Serial intervals are calculated for infectors who developed symptoms on week 50 (13 and 19 December, 2021).
@@
 On the other hand, the reproduction number of the Delta variant decreased from 0.89 (95\% CI: 0.82--0.97) to 0.68 (95\% CI: 0.62--0.74) between December 5, 2021, and January 9, 2022, and increased back up to 0.82 (95\% CI: 0.72--0.93) by January 23, 2022 (\fref{reproduction}A).
-We estimate the reproduction number ratios stayed rouhgly constant at around 2.2 (95\% CI: 2.0--2.5) between December 12--26, 2021, and slowly decreased to 1.4 (95\% CI: 1.2--1.7).
+We estimate the reproduction number ratios stayed roughly constant at around 2.2 (95\% CI: 2.0--2.5) between December 12--26, 2021, and slowly decreased to 1.4 (95\% CI: 1.2--1.7).
 However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
@@
 
-The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides cruicial information for epidemic control.
+The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides crucial information for epidemic control.
 A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
@@
 We tentatively hypothesize that these differences may be primarily driven by the network effect: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections \citep{park2020inferring,hart2022generation}.
-While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structres that involve repeated contacts between the same group of individuals (beacuse only the first infectious contact results in infection).
+While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
 The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
-Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shoter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.
+Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shorter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.
 
@@
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their intrinsic infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accouting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation interval distribution, which may have a longer mean).
+Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation interval distribution, which may have a longer mean).
 
@@
 We primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
-In doing so, we implicitly assume thta the delay between infection and reports is fixed.
+In doing so, we implicitly assume that the delay between infection and reports is fixed.
 However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
@@
 Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
-Our study sythesizes previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters;
+Our study synthesizes previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters;
 however, uncertainty remains in the intrinsic infectiousness profiles of SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.

### 2022-03-07 omicron-generation.tex
add libraries, start editing

@@
 \documentclass[12pt]{article}
-\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
+% \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
+\usepackage[top=1in,footskip=1in]{geometry}
 
@@
 These two quantities are linked by the generation-interval distributions of the resident and invading variants.
-For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical  $\Rinv = \Rres$.
+For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical  $\Rinv = \Rres > 1$.
 
-There is currently a lack of generation-interval estimates of SARS-CoV-2 variants, in part, due to difficulties in observing actual infection events.
-Instead, many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or testing \citep{abbott2022test} of the infector and the infectee.
-However, transmission-interval distributions are subject to dynamical biases and may not be an accurate proxy for the generation-interval distribution.
-For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
-Therefore, the observed differences (or the lack thereof) in transmission-interval distributions of the invading and resident variants can exaggerate or undermine true differences in their generation-interval distributions, depending on their growth rate differences.
+Estimating generation-interval is hard, largely due to difficulties in observing actual infection events.
+Many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
+However, transmission-interval distributions are subject to dynamical effects which can bias estimation.
+For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
+Therefore, observed differences in transmission-interval distributions between variants can give a biased picture of true differences in their generation-interval distributions, especially if their growth rates differ.
 
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with and without S gene target failures, but they neglected growth rate differences in their inference.
-We first discuss epidemiological context in the Netherlands during the study period and provide corrected estimates for the incubation period.
-We combine incubation-period estimates with serial-interval data to infer generation-interval distributions of the Delta and Omicron variants.
+The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly omicron during the study period) and without (mostly delta), but did not consider growth rate differences in their inference.
+Here, we discuss the epidemiological context in the Netherlands during the study period and take dynamics into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
 

### 2022-03-07 figure_compare.R

@@ -1,3 +1,4 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))

### 2022-03-07 figure_reproduction_advantage_between.R

@@ -1,3 +1,4 @@
+library(bbmle)
 library(vroom)
 library(readxl)
 library(tidyr)

### 2022-03-10 omicron-generation.tex
Minor changes and simplifications

@@
 
-To estimate the differences in growth rates of the Delta and Omicron variants, we first calculate the number of COVID-19 cases caused by each variant by multiplying the weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly, instead of daily, time series to remove changes in the numbers of cases caused testing patterns.
-We then fit a generalized linear model to obtain smooth trajectories for case time series \citep{wood2001mgcv}.
-More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood.
+To estimate the differences in growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly cidence numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
+We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
+%% More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood.
+%% JD: clarified previous s. Don't think we need this, with the cite.
 Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.
 
-To obtain confidence intervals on the estimated time-varying growth rates, we generate 1000 parameter sets using estimated spline coefficients and the corresponding variance-covariance matrices as means and variance-covariance matrices of multivariate normal distributions.
-We calculate time-varying growth rates from each parameter set and take 95\% equi-tailed quantiles to generate 95\% confidence limits.
+To obtain confidence intervals on the estimated time-varying growth rates, we generate 1000 parameter sets by resampling spline coefficients from a multivariate normal distribution using the estimated variance-covariance matrices.
+We calculate time-varying growth rates from each parameter set and use equi-tailed quantiles to generate 95\% confidence limits.
 
@@
 
-\cite{backer2021omicron} estimated the incubation periods from 513 individuals (258 Omicron and 255 Delta cases), who showed symptom onsets between 1 December 2021 and 2 January 2022.
-They used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring probability distributions of time of infection for each individual from their known exposure dates (and then taking the difference between symptom onset time and potential infection times).
-This method further assumes a uniform distribution over exposure intervals.
+\cite{backer2021omicron} estimated the incubation periods from 513 individuals (258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022.
+They used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates, using a uniform distribution, and compares this to a known symptom-onset time.
+\jd{Simplified. OK?}
 
-In practice, incubation periods (and any other epidemiological delays) can be measured in two ways.
+In practice, incubation periods (and other epidemiological delays) can be measured in two ways.
 The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
-We expect this forward incubation-period distribution $f_I(\tau)$ to remain constant over the course of an epidemic and therefore provide accurate descriptions of incubation periods.
-On the other hand, the backward incubation periods, measured from a cohort of individuals, who developed symptoms at the same time, are sensitive to epidemic growth.
+We expect this forward incubation-period distribution $f_I(\tau)$ to remain constant over the course of an epidemic and provide reliable estimates of the distribution across individuals. 
+Backward incubation periods are measured from a cohort of individuals who developed symptoms at the same time. 
+We expect the backward incubation-period distribution to be sensitive to epidemic dynamics.
+The difference arises because forward incubation periods look forward from the reference point towards symptom development, which is an individual-level process, while backward incubation periods look backwards towards an infection event, which requires an interaction with an infectious individual.
+
 In particular, when incidence of infection is growing exponentially, we are more likely to observe shorter backward incubation periods because there will be relatively more individuals who were infected recently.
@@
 where $r$ represents the epidemic growth rate.
-The method of \cite{backer2020incubation} effectively estimates the backward incubation-period distribution as it does not take growth rates into account.
+The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution without taking growth rates into account.
 
-To account for growth rate differences between the Delta and Omicron variants, we take the backward incubation-period distributions, estimated by \cite{backer2021omicron}, and calculate the corresponding forward incubation-period distributions by inverting \eref{backward}:
+We adjust for observed growth rate of each variant by taking the backward incubation-period distributions estimated by \cite{backer2021omicron} and calculating the corresponding forward incubation-period distributions by inverting \eref{backward}:
 \begin{equation}
@@
 where $b_I(\tau)$ is assumed to follow a Weibull distribution.
+\jd{Can this be explained better? Is this what backer assumed? We don't seem to say so.}
 To further account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously-inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parameterize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
+First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parameterize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
+\jd{Why do we match a mean to a median?, sounds crazy.}
 We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified Gamma distributions and estimate the corresponding forward distribution using \eref{forward}.
@@
 \end{equation}
-where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), $\tau_{g|i,1}$ represents the forward generation interval, and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
-When incidence of infection is growing exponentially, infectors are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean forward generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g}]$).
+where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), $\tau_{g|i,1}$ \jd{why does notation here not match equation?} represents the forward generation interval, and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
+When incidence of infection is growing exponentially, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean forward generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g}]$).
 
@@
 Then, we can see that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} depends on the exponential growth rate:
+
 \begin{align}

### 2022-03-10 omicron-generation.tex
good night (Te)

@@
 \end{align}
-On the other hand, the mean forward generation interval of our interest does not depend on the exponential growth rate:
+
+\jd{What's going on here?
+I'm having trouble wading through the equations, but I guess the idea is that $\tau_g$ is the mean generation interval for a symptom-based cohort, and that what we want is the mean for an infection-based cohort.
+Can we say that more clearly?
+Can we also work on accompanying notation?
+In particular, I don't like giving the weird one the more-normal name.}
+On the other hand, the mean forward generation interval of our interest \jd{“of our interest” requires some explanation, and maybe some thought} does not depend on the exponential growth rate:
 \begin{equation}
@@
 Therefore, while we generally expect the mean forward generation interval to be shorter than the mean forward serial interval for a growing epidemic, correlation between symptom onset and transmissibility can also have opposite effects.
+\jd{I'm not quite following here. Can we disentangle the two effects a bit better?}
 
+\jd{Didn't we just finish modeling incubation periods with both Weibull and lognormal? Is there a simpler way?}
 Here, we model forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions with lognormal distributions.

### 2022-03-10 omicron-generation.tex
small edit before bed

@@
 These two quantities are linked by the generation-interval distributions of the resident and invading variants.
-For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical  $\Rinv = \Rres > 1$.
+For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical ($\Rinv = \Rres > 1$).
 
@@
 They used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates, using a uniform distribution, and compares this to a known symptom-onset time.
-\jd{Simplified. OK?}
 
@@
 \begin{equation}
-f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx)  b_I(x)\dx},
+f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx) b_I(x)\dx}.
 \label{eq:forward}
 \end{equation}
-where $b_I(\tau)$ is assumed to follow a Weibull distribution.
-\jd{Can this be explained better? Is this what backer assumed? We don't seem to say so.}
-To further account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a Gamma distribution---we parameterize the Gamma distribution such that (i) its mean matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
-\jd{Why do we match a mean to a median?, sounds crazy.}
-We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified Gamma distributions and estimate the corresponding forward distribution using \eref{forward}.
+We model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
+To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
+First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
+\jd{Why do we match a mean to a median?, sounds crazy.} \swp{I changed to lognormal and now match median. I was just doing something quick and dirty when I did this...}
+We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.
 We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.

### 2022-03-10 sample_incubation.R

@@ -2,17 +2,17 @@ nsample <- 1000
 
 set.seed(101)
 backward_scale_nsgtf <- 4.93
-# diff(pgamma(c(4.51, 5.37), shape=500, rate=500/backward_scale_nsgtf))
-backward_scale_nsgtf_sample <- rgamma(nsample, shape=500, rate=500/backward_scale_nsgtf)
+# diff(plnorm(c(4.51, 5.37), meanlog=log(backward_scale_nsgtf), 0.0445))
+backward_scale_nsgtf_sample <- rlnorm(nsample, meanlog=log(backward_scale_nsgtf), 0.0445)
 
 backward_shape_nsgtf <- 1.83
-# diff(pgamma(c(1.59, 2.08), shape=210, rate=210/backward_shape_nsgtf))
-backward_shape_nsgtf_sample <- rgamma(nsample, shape=210, rate=210/backward_shape_nsgtf)
+# diff(plnorm(c(1.59, 2.08), meanlog=log(backward_shape_nsgtf), 0.068))
+backward_shape_nsgtf_sample <- rlnorm(nsample, meanlog=log(backward_shape_nsgtf), 0.068)
 
 backward_scale_sgtf <- 3.60
-# diff(pgamma(c(3.23, 3.98), shape=350, rate=350/backward_scale_sgtf))
-backward_scale_sgtf_sample <- rgamma(nsample, shape=350, rate=350/backward_scale_sgtf)
+# diff(plnorm(c(3.23, 3.98), meanlog=log(backward_scale_sgtf), 0.053))
+backward_scale_sgtf_sample <- rlnorm(nsample, meanlog=log(backward_scale_sgtf), 0.053)
 
 backward_shape_sgtf <- 1.50
-# diff(pgamma(c(1.32, 1.70), shape=240, rate=240/backward_shape_sgtf))
-backward_shape_sgtf_sample <- rgamma(nsample, shape=240, rate=240/backward_shape_sgtf)
+# diff(plnorm(c(1.32, 1.70), meanlog=log(backward_shape_sgtf), 0.0645))
+backward_shape_sgtf_sample <- rlnorm(nsample, meanlog=log(backward_shape_sgtf), 0.0645)

### 2022-03-10 omicron-generation.tex
more edits

@@
 To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median estimates and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
-\jd{Why do we match a mean to a median?, sounds crazy.} \swp{I changed to lognormal and now match median. I was just doing something quick and dirty when I did this...}
+First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
 We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.
@@
 \begin{equation}
-\tau_s =  - \tau_{i, 1} + \tau_{g} + \tau_{i, 2},
+\tau_s =  - \tau_{i, 1} + \tau_{g, \mathrm{symp}} + \tau_{i, 2},
 \label{eq:serial}
 \end{equation}
-where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), $\tau_{g|i,1}$ \jd{why does notation here not match equation?} represents the forward generation interval, and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
-When incidence of infection is growing exponentially, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean forward generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g}]$).
+where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
+Here, $\tau_{g, \mathrm{symp}}$ represents the generation interval between the infector and the infectee; we use the subscript $\mathrm{symp}$ to indicate that these generation intervals are measured from infectors who developed symptoms at the same time.
 
-We further note that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} is also likely shorter than the mean forward generation interval $\mathbb E[\hat{\tau}_{g}]$ that we wish to estimate because infectors with shorter incubation periods (due to exponential growth) also transmit faster.
-To illustrate this point, we write $f_{G|I}(\tau_g|\tau_{i,1})$ to represent forward generation-interval distribution conditional on the incubation period.
-Then, we can see that the mean forward generation interval $\mathbb E[\tau_{g}]$ defined in \eref{serial} depends on the exponential growth rate:
-
+The symptom-based generation-interval distribution describing \eref{serial} gives a biased picture because infectors who developed symptoms at the same time will have shorter incubation periods (when epidemic is growing) and therefore transmit earlier.
+This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
 \begin{align}
-\mathbb E[\tau_{g}] &= \int_0^\infty \int_0^\infty x_g f_{G|I}(x_g|x_{i,1}) b_I(x_{i,1}) \dd x_g x_{i, 1}\\
-&= \frac{\int_0^\infty \int_0^\infty x_g \exp(-rx_{i,1})  f_{G|I}(x_g|x_{i,1}) f_I(x_{i,1}) \dd x_g x_{i, 1}}{\int_0^\infty  \exp(-r x) f_I(x) \dx}
+f_{G, \mathrm{symptom}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) b_I(x) \dx,
+\end{align}
+where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$.
+Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time):
+\begin{align}
+f_{G, \mathrm{inf}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) f_I(x) \dx.
 \end{align}
 
-\jd{What's going on here?
-I'm having trouble wading through the equations, but I guess the idea is that $\tau_g$ is the mean generation interval for a symptom-based cohort, and that what we want is the mean for an infection-based cohort.
-Can we say that more clearly?
-Can we also work on accompanying notation?
-In particular, I don't like giving the weird one the more-normal name.}
-On the other hand, the mean forward generation interval of our interest \jd{“of our interest” requires some explanation, and maybe some thought} does not depend on the exponential growth rate:
-\begin{equation}
-\mathbb E[\hat{\tau}_{g}] = \int_0^\infty \int_0^\infty x_g f_{G|I}(x_g|x_{i,1}) f_I(x_{i,1}) \dd x_g x_{i, 1}.\\
-\label{eq:meangen}
-\end{equation}
-Therefore, while we generally expect the mean forward generation interval to be shorter than the mean forward serial interval for a growing epidemic, correlation between symptom onset and transmissibility can also have opposite effects.
-\jd{I'm not quite following here. Can we disentangle the two effects a bit better?}
+When epidemic is growing is exponentially, we have two opposing effects affecting the relationship between the mean serial and generation interval.
+First, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
+Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}} > \mathbb E[\tau_{g, \mathrm{symp}}]$.
+Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}$ \citep{park2021forward}.
+Hereafter, we use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time) for simplicity and drop the subscript $\mathrm{inf}$.
 
-\jd{Didn't we just finish modeling incubation periods with both Weibull and lognormal? Is there a simpler way?}
 Here, we model forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions with lognormal distributions.

### 2022-03-10 omicron-generation.tex
done?

@@
 
-To estimate the differences in growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly cidence numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
+To estimate the differences in growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
 We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
@@
 For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$.
-The mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.3--4.4 days) to 4.5 days (95\% CI: 3.8--5.6 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.
+The mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.
 
@@
 
+In both cases, the decrease in the reproduction number ratio coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant;
+a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the growth rate difference (\fref{epidemic}D).
+
 \begin{figure}[!th]
@@
 (B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
-(C--D) Phase diagrams of the reproduction ratio against instantaneous reproduction numbers of the Delta (C) and Omicron (D) variants.
 The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
@@
 \end{figure}
-
-Comparing phase diagrams of the reproduction ratio against instantaneous reproduction numbers of the Delta (\fref{reproduction}C) and Omicron (\fref{reproduction}D) variants helps us better understand the reduction in transmission advantage over time.
-As the reproduction number of the Delta variant decreases, the reproduction number ratio initially increases and stays constant. 
-Then, the reproduction number ratio and the reproduction number of the Delta variant both decreases at the same time, and the reproduction number of the Delta variant increases slightly near the end (\fref{reproduction}C).
-Similarly, the decrease in the reproduction number ratio coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant (\fref{reproduction}D);
-larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the growth rate difference (\fref{epidemic}D).
 

### 2022-03-10 figure_reproduction_advantage.R

@@ -177,13 +177,25 @@ g3 <- ggplot(advantagesumm2) +
   )
 
 g4 <- ggplot(advantagesumm2) +
-  geom_path(aes(R_omicron, R_advantage), lwd=1, arrow = arrow(length=unit(0.30,"cm")), col="Orange") +
+  geom_path(aes(R_omicron, R_advantage), lwd=1, arrow = arrow(length=unit(0.30,"cm")), col="black") +
   scale_x_log10("Omicron reproduction number") +
   scale_y_log10("Reproduction number ratio") +
   theme(
     panel.grid = element_blank()
   )
 
-gfinal <- ggarrange(g1, g2, g3, g4, nrow=2, labels=c("A", "B", "C", "D"), draw=FALSE)
+g5 <- ggplot(advantagesumm2) +
+  geom_path(aes(R_delta, R_omicron), lwd=1, arrow = arrow(length=unit(0.30,"cm")), col="black") +
+  scale_x_log10("Delta reproduction number") +
+  scale_y_log10("Omicron reproduction number") +
+  theme(
+    panel.grid = element_blank()
+  )
 
-ggsave("figure_reproduction_advantage.pdf", gfinal, width=8, height=6)
+gfinal <- ggarrange(g1, g2, nrow=1, labels=c("A", "B"), draw=FALSE)
+
+ggsave("figure_reproduction_advantage.pdf", gfinal, width=8, height=3)
+
+gsupp <- ggarrange(g5, g3, g4, nrow=1, labels=c("A", "B", "C"), draw=FALSE)
+
+ggsave("figure_reproduction_advantage_phase.pdf", gsupp, width=9, height=3)

### 2022-03-10 omicron-generation.tex
done

@@
 For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021).
-\citep{backer2021omicron} previously reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.5 days) for this cohort, but the overall shapes of the distributions are similar (\fref{serial}A).
+\citep{backer2021omicron} previously reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort, but the overall shapes of the distributions are similar (\fref{serial}A).
 However, when we account for growth rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
 Notably, both variants have similar mode, but the transmissibility of the omicron variant decays much faster (\fref{serial}B).
+We note that the mean forward generation interval of the Omicron (and Delta) variant is shorter (longer) than its mean serial interval due to the effect of a growing (declining) epidemic.
+Instead, if we neglect growth rate differences and assume $r= 0/\mathrm{day}$ for both variants, the estimated shape of the generation-interval distributions become more similar and the differences in their means become smaller (Supplementary Figure S1): 3.5 days (95\% CI: 3.1--3.9 days) for the Omicron variant and 4.1 days (95\% CI: 3.9--4.3 days).
 
-While we estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), the differences in the mean generation intervals are unclear for between-household pairs.
+We estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), but the differences in the mean generation intervals are unclear for between-household pairs.
 Consistent with the original study, which also reported shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
@@
 \caption{
-\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
+\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants.}
 (A) Observed and fitted forward serial-interval distributions for within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
@@
 
+\section*{Supplementary Materials}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{figure_compare_r_0.pdf}
+\caption{
+\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants while neglecting growth rate differences.}
+See Figure 3 in the main text for figure caption.
+}
+\end{figure}
+
+\pagebreak
+
 \bibliography{omicron.bib}

### 2022-03-10 figure_compare_r_0.R
done?

@@ -1,0 +1,126 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(gridExtra)
+source("serialfun.R")
+load("rdaout/fit_lognormal_base_r_0.rda")
+
+mean_nsgtf <- exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[1]]+coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]^2/2)
+
+tmp_nsgtf <- c(
+  exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]^2/2),
+  exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]^2/2) * coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]
+)
+
+sd_mean_nsgtf <- sqrt(c(tmp_nsgtf %*% vcov(fit_lognormal_base_r_0_50_nsgtf_within) %*% tmp_nsgtf))
+
+mean_sgtf <- exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[1]]+coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]^2/2)
+
+tmp_sgtf <- c(
+  exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]^2/2),
+  exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]^2/2) * coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]
+)
+
+sd_mean_sgtf <- sqrt(c(tmp_sgtf %*% vcov(fit_lognormal_base_r_0_50_sgtf_within) %*% tmp_sgtf))
+
+c(mean_nsgtf - 1.96 * sd_mean_nsgtf, mean_nsgtf, mean_nsgtf + 1.96 * sd_mean_nsgtf)
[…]

### 2022-03-10 fit_lognormal_base_r_0.R

@@ -1,0 +1,74 @@
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+source("sample_incubation.R")
+
+r_nsgtf <- 0
+r_sgtf <- 0
+rho <- 0.75
+
+moment_0_nsgtf <- integrate(function(z) {
+  dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
+}, 0, Inf)[[1]]
+
+moment_1_nsgtf <- integrate(function(z) {
+  z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
+}, -100, 100)[[1]]
+
+moment_2_nsgtf <- integrate(function(z) {
+  z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
+}, -100, 100)[[1]]
+
+moment_0_sgtf <- integrate(function(z) {
+  dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * z)
+}, 0, Inf)[[1]]
+
+moment_1_sgtf <- integrate(function(z) {
+  z * dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * z)
+}, -100, 100)[[1]]
+
+moment_2_sgtf <- integrate(function(z) {
+  z^2 * dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * z)
+}, -100, 100)[[1]]
+
+mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
+var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
+
+logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
+logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
+
+mean_inc_sgtf <- moment_1_sgtf/moment_0_sgtf
+var_inc_sgtf <- moment_2_sgtf/moment_0_sgtf - mean_inc_sgtf^2
+
[…]

### 2022-03-10 omicron-generation.tex
Done with Methods for now.

@@
 \newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
-\newcommand{\Ro}{\Rx{0}}
-\newcommand{\Rc}{\Rx{\mathrm{c}}}
-\newcommand{\Rs}{\Rx{\mathrm{s}}}
+\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
 \newcommand{\Rres}{\Rx{\mathrm{res}}}
 \newcommand{\Rinv}{\Rx{\mathrm{inv}}}
-\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
 \newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
-\newcommand{\Rintrinsic}{\ensuremath{{\mathcal R}_{\textrm{\tiny intrinsic}}}\xspace}
+\newcommand{\Rt}{\ensuremath{{\mathcal R}(t)}\xspace}
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
@@
 Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}$ \citep{park2021forward}.
-Hereafter, we use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time) for simplicity and drop the subscript $\mathrm{inf}$.
+For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time), and drop the subscript $\mathrm{inf}$.
 
-Here, we model forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions with lognormal distributions.
-Both distributions are parameterized by log means, $\mu_I$ and $\mu_G$, and log variances, $\sigma_I^2$ and $\sigma_G^2$.
-Then, given the correlation coefficient $\rho$, the forward generation-interval distribution conditional conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
+We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
+The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient $\rho$.
+Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
+\jd{There is some scale mixing going on here and I can't figure it all out.
+Did you mean to write ``log mean'' just above?
+That might fix most of it, but then you would need to also say that $\rho$ is the log-scale correlation, or something like that.}
 Assuming that the incidence of infection is changing exponentially at rate $r$, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be calculated by integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ \citep{park2021forward}:
@@
 \end{equation}
-where $\phi$ represents the normalization constant to satisfy $\int f_S(x) \dx = 1$;
+where $\phi$ is a normalization constant chosen so that $\int f_S(x) \dx = 1$;
 $-\alpha_1$ corresponds to the incubation period of the infector;
@@
 We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation.
-Using this incubation-period distribution, we fit \fref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
+Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
 We then calculate the mean forward generation interval using \eref{meangen} and approximate the 95\% confidence interval using the Delta method.
-We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be practically unidentifiable.
+\jd{Fixed one eref, but failed to fix the other.}
+We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.
+\jd{Consider more discussion and/or sensitivity explorations.}
 
@@
 
-We use our estimates of the generation-interval distributions to infer instantaneous reproduction numbers $\mathcal R(t)$ of the Delta and Omicron variant, as well as the ratio between two reproduction numbers.
+We use our estimates of the generation-interval distributions to infer instantaneous reproduction numbers \Rt\ of the Delta and Omicron variant, as well as the ratio between two reproduction numbers.
 Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
+\jd{We should think about this, too -- is it really just the forward if it's \Rt?}
 \begin{equation}
-\mathcal R(t) = \frac{i(t)}{\int_0^\infty i(t-x) g(x) \dx},
+\Rt = \frac{i(t)}{\int_0^\infty i(t-x) g(x) \dx},
 \end{equation}
@@
 Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for week 50---when epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
-Then, the instantaneous reproduction number is estimated using the renewal equation \citep{fraser2007estimating}:
+%% Then, the instantaneous reproduction number is estimated using the renewal equation \citep{fraser2007estimating}: This looks like an orphan to me  \jd
+\jd{Another point that may need a bit more discussion:}
 Incidence of infection is approximated by shifting the smoothed case trajectories by one week to account for reporting delays.

### 2022-03-11 omicron-generation.tex
Hand back to Park.

@@
 \newcommand{\gdist}{g} %% generation-interval distribution
-\newcommand{\idist}{\ell} %% incubation period distribution
+\newcommand{\idist}{\ell} %% incubation-period distribution
 
@@
 Here, we analyze serial-interval data describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021, during which the number of infections caused by the Delta variant decreased as the number of Omicron infections increased.
-Accounting for growth rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
+Accounting for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
 These differences may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
@@
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly omicron during the study period) and without (mostly delta), but did not consider growth rate differences in their inference.
+The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly omicron during the study period) and without (mostly delta), but did not consider growth-rate differences in their inference.
 Here, we discuss the epidemiological context in the Netherlands during the study period and take dynamics into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
@@
 
-\fref{epidemic} summarizes epidemiological context in the Netherlands during the study period.
-The first Omicron cases was detected in the Netherlands from a sample collected on 19 November 2021 \citep{backer2021omicron}, during which the number of COVID-19 cases were decreasing (\fref{epidemic}A).
-As the Omicron variant continued to spread and increase in proportion (\fref{epidemic}B), the number of COVID-19 cases  increased (\fref{epidemic}A).
-Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to tease apart differences in their epidemiological dynamics (\fref{epidemic}C).
+\fref{epidemic} summarizes the epidemiological context in the Netherlands during the study period.
+The first known Omicron case in the Netherlands was sampled on 19 November 2021 \citep{backer2021omicron}, during a period when COVID-19 incidence was decreasing (\fref{epidemic}A).
+As the Omicron variant continued to spread and increase in proportion (\fref{epidemic}B), the number of COVID-19 cases started to increase (\fref{epidemic}A).
+Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
+\jd{Can we simplify following s.~by eliminating CIs? Maybe add $\approx$ to indicate that these are just estimates.}
 The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r = -0.01/\mathrm{day}$ (95\% CI: $-0.04/\mathrm{day}$--$0.02/\mathrm{day}$) to $r = -0.09/\mathrm{day}$ (95\% CI: $-0.11/\mathrm{day}$--$-0.07/\mathrm{day}$) by the week of January 16, 2022, and increasing back up to $r = -0.04/\mathrm{day}$ (95\% CI: $-0.07/\mathrm{day}$--$-0.01/\mathrm{day}$).
 The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ (95\% CI: $0.15/\mathrm{day}$--$0.21/\mathrm{day}$) on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ (95\% CI: $0.00/\mathrm{day}$--$0.09/\mathrm{day}$).
-Finally, we find that the growth rate difference between the Delta and Omicron variants decreased over time---this finding echoes our previous work, which showed that growth rate differences change over the course of an epidemic even when the ratio of the reproduction numbers remain constant \citep{park2021roles}.
+Finally, we find that the growth-rate difference between the Delta and Omicron variants decreased over time---
+\jd{The following comes across too strong; the roles work doesn't explain all of the decrease.
+Something like: this effect is partly due to changes in the overall growth rate and partly due to changes in differential immunity… I'm kind of thinking simplify or omit and come back in Discussion, now.}
+this finding echoes our previous work, which showed that growth-rate differences change over the course of an epidemic even when the ratio of the reproduction numbers remain constant \citep{park2021roles}.
 
@@
 Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
-(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth rate differences (purple).
+(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth-rate differences (purple).
 Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
@@
 
-For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} previously found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
-However, during this period, the number of COVID-19 cases caused by the Delta and Omicron variants showed opposite patterns of epidemic growth/decay (\fref{epidemic}).
-Since the number of COVID-19 cases caused by the Delta variant was decreasing, longer incubation periods would have been more likely to be observed.
-In contrast, the rapid growth in the number of COVID-19 cases caused by the Omicron variant would have caused shorter incubation periods to be more likely to be observed.
-When we account for these growth rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
+For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
+However, these measurements were done during a period when the incidence of Omicron was increasing while the increasing of Delta was decreasing (\fref{epidemic}).
+Thus, dynamical bias would be expected to lead to shorter observed (backward) incubation periods in Omicron, and longer observed incubation periods in Delta. 
+When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
 Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C).
-For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$.
-The mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.
+\jd{Do we need the ff.~s.? What does it add to the Fig?}
+For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$, while
+the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.
 
@@
 \caption{
-\textbf{Observed and corrected differences in incubation period distributions of Delta and Omicron variants.}
+\textbf{Observed and corrected differences in incubation-period distributions of Delta and Omicron variants.}
 (A) Posterior median estimates of the observed (backward) incubation periods of the Delta (black) and Omicron (orange) variants by \cite{backer2021omicron}.
@@
 
-Our corrected estimates of the forward incubation period distributions further allow us to infer the forward generation-interval distributions.
+Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
 For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021).
-\citep{backer2021omicron} previously reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort, but the overall shapes of the distributions are similar (\fref{serial}A).
-However, when we account for growth rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
-Notably, both variants have similar mode, but the transmissibility of the omicron variant decays much faster (\fref{serial}B).
-We note that the mean forward generation interval of the Omicron (and Delta) variant is shorter (longer) than its mean serial interval due to the effect of a growing (declining) epidemic.
-Instead, if we neglect growth rate differences and assume $r= 0/\mathrm{day}$ for both variants, the estimated shape of the generation-interval distributions become more similar and the differences in their means become smaller (Supplementary Figure S1): 3.5 days (95\% CI: 3.1--3.9 days) for the Omicron variant and 4.1 days (95\% CI: 3.9--4.3 days).
+\citep{backer2021omicron} reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort,
+\jd{I'm not clear what the last part of this s.~means.} but the overall shapes of the distributions are similar (\fref{serial}A).
+However, when we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
+Notably, both variants have similar mode, but the transmissibility of the omicron variant is estimated to decay much faster (\fref{serial}B).
+Note that the mean forward generation interval of the Omicron (resp., Delta) variant is shorter (longer) than its mean serial interval due to the effect of a growing (declining) epidemic.
+\jd{What is the relevace of the ff.~s.? Do we want it?}
+Instead, if we neglect growth-rate differences and assume $r= 0/\mathrm{day}$ for both variants, the estimated shape of the generation-interval distributions become more similar and the differences in their means become smaller (Supplementary Figure S1): 3.5 days (95\% CI: 3.1--3.9 days) for the Omicron variant and 4.1 days (95\% CI: 3.9--4.3 days).
 
+\jd{What is up with stratification here? I'm starting to get lost. Is panel C showing CIs again? If so can you say so again, and also use the same style as the previous figure?}
 We estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), but the differences in the mean generation intervals are unclear for between-household pairs.
@@
 In both cases, the decrease in the reproduction number ratio coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant;
-a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the growth rate difference (\fref{epidemic}D).
+a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the growth-rate difference (\fref{epidemic}D).
 
@@
 The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides crucial information for epidemic control.
-A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
+A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation-period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
 Here, we build on our previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
@@
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their intrinsic infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation interval distribution, which may have a longer mean).
+Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation-interval distribution, which may have a longer mean).
 
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction ratios ($> 4$ fold) in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate that the reproduction number ratio of 2.8 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
-We also show that both growth rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.
+With our generation-interval estimates, we estimate that the reproduction number ratio of 2.8 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+We also show that both growth-rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.
 
@@
 \caption{
-\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants while neglecting growth rate differences.}
+\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants while neglecting growth-rate differences.}
 See Figure 3 in the main text for figure caption.

### 2022-03-11 omicron-generation.tex
think done

@@
 f_{G, \mathrm{inf}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) f_I(x) \dx.
+\label{eq:meangen}
 \end{align}
@@
 We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
-The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient $\rho$.
-Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
-\jd{There is some scale mixing going on here and I can't figure it all out.
-Did you mean to write ``log mean'' just above?
-That might fix most of it, but then you would need to also say that $\rho$ is the log-scale correlation, or something like that.}
+The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient on a log scale $\rho$.
+Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
 Assuming that the incidence of infection is changing exponentially at rate $r$, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be calculated by integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ \citep{park2021forward}:
@@
 We then calculate the mean forward generation interval using \eref{meangen} and approximate the 95\% confidence interval using the Delta method.
-\jd{Fixed one eref, but failed to fix the other.}
 We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.
-\jd{Consider more discussion and/or sensitivity explorations.}
+In Supplementary Material, we explore how assumptions about $\rho$ affect inferences of the generation-interval distribution.
 
@@
 Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
-\jd{We should think about this, too -- is it really just the forward if it's \Rt?}
+% \jd{We should think about this, too -- is it really just the forward if it's \Rt?}
+% \swp{Not sure what you mean... during exponential growth, forward = intrinsic...}
 \begin{equation}
@@
 Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for week 50---when epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
-%% Then, the instantaneous reproduction number is estimated using the renewal equation \citep{fraser2007estimating}: This looks like an orphan to me  \jd
-\jd{Another point that may need a bit more discussion:}
 Incidence of infection is approximated by shifting the smoothed case trajectories by one week to account for reporting delays.
+This method of approximating incidence of infection assumes a fixed delay between infection and case reporting;
+in practice, deconvolution is required to accurately estimate the incidence of infection \citep{goldstein2009reconstructing}.
+Case reports are also sensitive to changes in testing behavior, and therefore our estimates of $\Rt$ must be interpreted with care.
 Confidence intervals are calculated by sampling parameters of the smoothed case trajectories as well as the generation-interval distributions from multivariate normal distributions and repeating the analysis 1000 times.
@@
 Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
-\jd{Can we simplify following s.~by eliminating CIs? Maybe add $\approx$ to indicate that these are just estimates.}
-The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r = -0.01/\mathrm{day}$ (95\% CI: $-0.04/\mathrm{day}$--$0.02/\mathrm{day}$) to $r = -0.09/\mathrm{day}$ (95\% CI: $-0.11/\mathrm{day}$--$-0.07/\mathrm{day}$) by the week of January 16, 2022, and increasing back up to $r = -0.04/\mathrm{day}$ (95\% CI: $-0.07/\mathrm{day}$--$-0.01/\mathrm{day}$).
-The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ (95\% CI: $0.15/\mathrm{day}$--$0.21/\mathrm{day}$) on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ (95\% CI: $0.00/\mathrm{day}$--$0.09/\mathrm{day}$).
-Finally, we find that the growth-rate difference between the Delta and Omicron variants decreased over time---
-\jd{The following comes across too strong; the roles work doesn't explain all of the decrease.
-Something like: this effect is partly due to changes in the overall growth rate and partly due to changes in differential immunity… I'm kind of thinking simplify or omit and come back in Discussion, now.}
-this finding echoes our previous work, which showed that growth-rate differences change over the course of an epidemic even when the ratio of the reproduction numbers remain constant \citep{park2021roles}.
+The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$.
+The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$.
+Finally, we find that the growth-rate difference between the Delta and Omicron variants decreased over time.
 
@@
 Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C).
-\jd{Do we need the ff.~s.? What does it add to the Fig?}
+% \jd{Do we need the ff.~s.? What does it add to the Fig?}
+% \swp{I think people like to see actual values...}
 For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$, while
@@
 Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
-For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021).
-\citep{backer2021omicron} reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort,
-\jd{I'm not clear what the last part of this s.~means.} but the overall shapes of the distributions are similar (\fref{serial}A).
-However, when we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
+For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021)---\citep{backer2021omicron} reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort (\fref{serial}A).
+When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
 Notably, both variants have similar mode, but the transmissibility of the omicron variant is estimated to decay much faster (\fref{serial}B).
-Note that the mean forward generation interval of the Omicron (resp., Delta) variant is shorter (longer) than its mean serial interval due to the effect of a growing (declining) epidemic.
-\jd{What is the relevace of the ff.~s.? Do we want it?}
-Instead, if we neglect growth-rate differences and assume $r= 0/\mathrm{day}$ for both variants, the estimated shape of the generation-interval distributions become more similar and the differences in their means become smaller (Supplementary Figure S1): 3.5 days (95\% CI: 3.1--3.9 days) for the Omicron variant and 4.1 days (95\% CI: 3.9--4.3 days).
+Note that the mean forward generation interval of the Omicron (respectively, Delta) variant is shorter (longer) than its mean serial interval due to the effect of a growing (declining) epidemic.
+The estimated differences of the mean generation interval increase if we assume that generation intervals and incubation periods are less correlated (and therefore, lower values of $\rho$; Supplementary Figure S1).
+On the other hand, neglecting growth-rate differences (i.e., assuming $r= 0/\mathrm{day}$ for both variants) leads to the underestimation of these differences (Supplementary Figure S2): 3.5 days (95\% CI: 3.1--3.9 days) for the Omicron variant and 4.1 days (95\% CI: 3.9--4.3 days).
 
-\jd{What is up with stratification here? I'm starting to get lost. Is panel C showing CIs again? If so can you say so again, and also use the same style as the previous figure?}
+We then estimate the mean forward generation intervals for all other cohorts, stratified by the week of symptom onset (weeks 50 and 51) and the type of transmission (within and between household(s)).
+As before (see \fref{incubation}C), we repeat the analysis for different values of growth rates for the Delta ($r=-0.1/\mathrm{days}$ to $0/\mathrm{days}$) and Omicron ($r=0.1/\mathrm{days}$ to $0.2/\mathrm{days}$) variants and plot the maximum likelihood estimates and the associated CIs.
 We estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), but the differences in the mean generation intervals are unclear for between-household pairs.
@@
 \begin{figure}[!th]
+\includegraphics[width=\textwidth]{figure_compare_rho.pdf}
+\caption{
+\textbf{Sensitivity of the estimates of the mean generation interval to the assumed values of the correlation coefficient of the lognormal distribution.}
+Lines represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
+For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during week 50.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!th]
 \includegraphics[width=\textwidth]{figure_compare_r_0.pdf}

### 2022-03-11 figure_incubation.R

@@ -104,11 +104,11 @@ g2 <- ggplot(corrected_nsgtf) +
 
 g3 <- ggplot(corrected_mean_nsgtf) +
   geom_line(aes(r, median), lwd=1) +
-  geom_line(aes(r, lwr))  +
-  geom_line(aes(r, upr)) +
+  geom_line(aes(r, lwr), lwd=1)  +
+  geom_line(aes(r, upr), lwd=1) +
   geom_line(data=corrected_mean_sgtf, aes(r-0.2, median), lty=2, col="orange", lwd=1)  +
-  geom_line(data=corrected_mean_sgtf, aes(r-0.2, lwr), lty=2, col="orange")  +
-  geom_line(data=corrected_mean_sgtf, aes(r-0.2, upr), lty=2, col="orange") +
+  geom_line(data=corrected_mean_sgtf, aes(r-0.2, lwr), lty=2, col="orange", lwd=1)  +
+  geom_line(data=corrected_mean_sgtf, aes(r-0.2, upr), lty=2, col="orange", lwd=1) +
   scale_x_continuous("Delta growth rate (1/day)",
                      sec.axis = sec_axis(trans=~.+0.2, "Omicron growth rate (1/day)")) +
   scale_y_continuous("Mean forward incubation period (days)") +

### 2022-03-11 fit_lognormal_base_rho.R

@@ -1,0 +1,85 @@
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+load("rdaout/calculate_incubation_mle.rda")
+load("rdaout/fit_lognormal_base.rda")
+
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+rhovec <- seq(0.5, 0.9, length.out=11)
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_50_sgtf_within <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="within")
+
+serialdata_50_nsgtf_within <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="within")
+
+data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
+data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+
+fit_lognormal_base_rho_50_nsgtf_within <- lapply(rhovec, function(rho) {
+  ff <- fitfun_lognormal(data=data_50_nsgtf_within, 
+                         logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
+                         logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
+                         logmean_inc=logmean_inc_nsgtf, 
+                         logsd_inc=logsd_inc_nsgtf, 
+                         rho=rho, 
+                         r=r_nsgtf)
+  
+  mean <- exp(coef(ff)[[1]]+coef(ff)[[2]]^2/2)
+  
+  cc <- confint(ff, method="quad")
+  
+  tmp <- c(
+    exp(coef(ff)[[1]]) * exp(coef(ff)[[2]]^2/2),
+    exp(coef(ff)[[1]]) * exp(coef(ff)[[2]]^2/2) * coef(ff)[[2]]
+  )
+  
+  sd_mean <- sqrt(c(tmp %*% vcov(ff) %*% tmp))
+  
+  data.frame(
+    rho=rho,
+    param=c("logmean", "logsd", "mean"),
+    est=c(coef(ff)[[1]], coef(ff)[[2]], mean),
+    lwr=c(cc[1,1], cc[2,1], mean - 1.96 * sd_mean),
[…]

### 2022-03-15 omicron-generation.tex
tweaking abstract

@@
 Estimating the differences in the incubation-period and serial-interval distributions of SARS-CoV-2 variants is critical to understanding changes in their transmission dynamics and therefore controlling their outbreaks.
-However, these comparisons are inherently difficult due to their dependence on epidemic growth rates---for example, when an epidemic is growing exponentially, we are more likely to observe individuals who have been infected recently, decreasing the mean observed incubation period and thereby increasing the mean observed serial interval.
-Here, we analyze serial-interval data describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021, during which the number of infections caused by the Delta variant decreased as the number of Omicron infections increased.
-Accounting for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
-These differences may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
+However, these comparisons are inherently difficult due to their dependence on epidemic growth rates---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have shorter incubation periods.
+Here, we re-analyze incubation-period and serial-interval data collected by Backer \textit{et al.}, describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
+Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
+When we account for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
+The differences in generation intervals may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
 Neglecting differences in the generation-interval distributions can further bias in estimates of transmission advantages of the Omicron variant.

### 2022-03-15 omicron-generation.tex
some more clarification

@@
 The data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates or symptom onset dates.
-The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure, week of infectors' symptom onset date (week 50 and 51), and the type of transmission (within or between households).
+The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (weeks 50 and 51), and the type of transmission (within or between households).
+In the main text, we combine data from weeks 50 and 51 and instead present a stratified analysis in Supplementary Material.
 For simplicity, we refer to transmission pairs with and without S gene target failure as Omicron and Delta transmission pairs, respectively.
+Incubation period data are not publicly available with the original article; instead, we rely on previously-estimated incubation-period distributions by \cite{backer2021omicron} to derive growth-rate-adjusted .
 See original article for details of data collection.
@@
 
-In practice, incubation periods (and other epidemiological delays) can be measured in two ways.
+In practice, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward.
 The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
@@
 
-We adjust for observed growth rate of each variant by taking the backward incubation-period distributions estimated by \cite{backer2021omicron} and calculating the corresponding forward incubation-period distributions by inverting \eref{backward}:
+For a given growth rate $r$, the corresponding forward incubation-period distributions can be calculated by inverting \eref{backward}:
 \begin{equation}
@@
 \end{equation}
-We model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
+Since incubation-period data are not provided, we are not able to fit \eref{forward} directly; 
+instead we take the backward incubation-period distributions $b_I(x)$ estimated by \cite{backer2021omicron}, which was originally assumed to follow a Weibull distribution, and apply \eref{forward}.
+In particular, \cite{backer2021omicron} estimated the scale and shape parameters of the Weibull distribution to be 4.93 (95\% CI: 4.51--5.37) and 1.83 (95\% CI: 1.59--2.08), respectively, for the non-SGTF cases, and 3.60 (95\% CI: 3.23--3.98) and 1.50 ((95\% CI: 1.32--1.70), respectively, for SGTF cases.
+
+We also model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
 To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified Gamma distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
+First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified lognormal distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
 We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.

### 2022-03-15 omicron-generation.tex
micro edit
update output

@@
 In the main text, we combine data from weeks 50 and 51 and instead present a stratified analysis in Supplementary Material.
-For simplicity, we refer to transmission pairs with and without S gene target failure as Omicron and Delta transmission pairs, respectively.
-Incubation period data are not publicly available with the original article; instead, we rely on previously-estimated incubation-period distributions by \cite{backer2021omicron} to derive growth-rate-adjusted .
+For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
+Incubation period data are not publicly available with the original article; instead, we rely on previous estimates by \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
 See original article for details of data collection.

### 2022-03-16 omicron-generation.tex
new files

@@
 In particular, when incidence of infection is growing exponentially, we are more likely to observe shorter backward incubation periods because there will be relatively more individuals who were infected recently.
-Specifically, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
+Assuming that incidence is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
 \begin{equation}
-b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx},
+b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx}.
 \label{eq:backward}
 \end{equation}
-where $r$ represents the epidemic growth rate.
 The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution without taking growth rates into account.
@@
 We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.
+We repeat the analysis across plausible ranges of $r$ for the Delta and Omicron variants separately (discussed later).
 
@@
 
-Dynamical biases in the serial-interval distributions are more complex because the serial interval depends on the incubation periods of the infector and the infectee as well as the generation interval between them.
+\begin{figure}[!tp]
+\includegraphics[width=\textwidth]{figure_diagram.pdf}
+\caption{
+\textbf{Schematic diagrams of serial and generation intervals from symptom- and infection-based infector cohorts.}
+(A) Serial intervals are typically measured from the cohort of infectors who develop symptoms at the same time. 
+In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
+(B) Generation intervals are better measured from the cohort of infectors who are infected at the same time because their incubation periods will be unbiased.
+\label{fig:diagram}
+}
+\end{figure}
+
+Dynamical biases in the serial-interval distributions are more complex because the serial interval depends on the incubation periods of the infector and the infectee as well as the generation interval between them (\fref{diagram}).
 For example, \cite{backer2021omicron} measured the forward serial-interval distributions from cohorts of infectors who developed symptoms during the same week.
@@
 
-The symptom-based generation-interval distribution describing \eref{serial} gives a biased picture because infectors who developed symptoms at the same time will have shorter incubation periods (when epidemic is growing) and therefore transmit earlier.
+The symptom-based generation-interval distribution describing \eref{serial} gives a biased picture because infectors who developed symptoms at the same time will have shorter incubation periods (when epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
 This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
@@
 where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$.
-Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time):
+Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time, \fref{diagram}B):
 \begin{align}
@@
 First, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
-Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}} > \mathbb E[\tau_{g, \mathrm{symp}}]$.
-Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}$ \citep{park2021forward}.
-For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time), and drop the subscript $\mathrm{inf}$.
+Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}}] > \mathbb E[\tau_{g, \mathrm{symp}}]$.
+Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
+For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time, \fref{diagram}B), and drop the subscript $\mathrm{inf}$.
 
@@
 and $\tau - \alpha_2$ corresponds to the incubation period of the infectee.
+As discussed earlier in Section 2.3, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
+As we show later in the results section, the exponential growth rate changes over the study period, including weeks 50 and 51;
+for illustrative purpose, we choose values of $r$ that represent the dynamics of Delta and Omicron infections during this period and repeat the analysis across plausible ranges of $r$ (discussed later in detail).
 
@@
 Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
-The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$.
-The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$.
+The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022.
+The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
 Finally, we find that the growth-rate difference between the Delta and Omicron variants decreased over time.
+Hereafter, we choose $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as our nominal growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
+We then evaluate the growth-rate effects across $r=-0.1/\mathrm{day}$--$0.0/\mathrm{day}$ for the Delta variant and $r=0.1/\mathrm{day}$--$0.2/\mathrm{day}$ for the Omicron variant as a sensitivity analysis.
 
@@
 For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021)---\citep{backer2021omicron} reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort (\fref{serial}A).
-When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days).
+When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days) for the Delta variant.
 Notably, both variants have similar mode, but the transmissibility of the omicron variant is estimated to decay much faster (\fref{serial}B).
@@
 \textbf{Estimated forward generation-interval distributions of Delta and Omicron variants while neglecting growth-rate differences.}
-See Figure 3 in the main text for figure caption.
+See Figure 4 in the main text for figure caption.
 }

### 2022-03-16 figure_epidemic.R

@@ -8,6 +8,7 @@ library(mgcv)
 library(mvtnorm)
 library(egg)
 
+datevec <- seq(as.Date("2021-11-28"), as.Date("2022-01-30"), by=1)
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 
 nsample <- 1000
@@ -101,10 +102,10 @@ g2 <- ggplot(variant3) +
 gfit1 <- gam(log(delta_cases)~s(time, bs="cs"), data=variant4, method="REML")
 gfit2 <- gam(log(omicron_cases)~s(time, bs="cs"), data=variant4, method="REML")
 
-gfit1_p1 <- predict(gfit1, newdata = data.frame(time=seq(1, 10, by=0.1)), type = "lpmatrix")
-gfit1_p2 <- predict(gfit1, newdata = data.frame(time=seq(1, 10, by=0.1)+0.01), type = "lpmatrix")
-gfit2_p1 <- predict(gfit2, newdata = data.frame(time=seq(1, 10, by=0.1)), type = "lpmatrix")
-gfit2_p2 <- predict(gfit2, newdata = data.frame(time=seq(1, 10, by=0.1)+0.01), type = "lpmatrix")
+gfit1_p1 <- predict(gfit1, newdata = data.frame(time=seq(1, 10, by=1/7)), type = "lpmatrix")
+gfit1_p2 <- predict(gfit1, newdata = data.frame(time=seq(1, 10, by=1/7)+0.01), type = "lpmatrix")
+gfit2_p1 <- predict(gfit2, newdata = data.frame(time=seq(1, 10, by=1/7)), type = "lpmatrix")
+gfit2_p2 <- predict(gfit2, newdata = data.frame(time=seq(1, 10, by=1/7)+0.01), type = "lpmatrix")
 
 gfit1_Xp <- (gfit1_p2 - gfit1_p1) / 0.01/7
 gfit2_Xp <- (gfit2_p2 - gfit2_p1) / 0.01/7
@@ -118,23 +119,27 @@ gfit2_post <- apply(gfit2_sim, 1, function(x) {gfit2_Xp %*% x})
 advantage_post <- gfit2_post - gfit1_post
 
 delta_growth <- data.frame(
-  time=seq(1, 10, by=0.1),
+  time=seq(1, 10, by=1/7),
   median=apply(gfit1_post, 1, median),
   lwr=apply(gfit1_post, 1, quantile, 0.025),
   upr=apply(gfit1_post, 1, quantile, 0.975),
[…]

### 2022-03-16 figure_incubation.R

@@ -80,11 +80,12 @@ corrected_sgtf <- tibble(
 )
 
 g1 <- ggplot(observed_nsgtf) +
-  geom_line(aes(time, density, col="Delta"), lwd=1) +
-  geom_line(data=observed_sgtf, aes(time, density, col="Omicron"), lwd=1) +
+  geom_line(aes(time, density, col="Delta", lty="Delta"), lwd=1) +
+  geom_line(data=observed_sgtf, aes(time, density, col="Omicron", lty="Omicron"), lwd=1) +
   scale_x_continuous("Observed incubation period (days)", expand=c(0, 0)) +
   scale_y_continuous("Density (1/day)", expand=c(0, 0), limits=c(0, 0.23)) +
-  scale_color_manual(values=c("black", "orange")) +
+  scale_color_manual("", values=c("black", "orange")) +
+  scale_linetype_discrete("") +
   theme(
     panel.grid = element_blank(),
     legend.position = c(0.8, 0.8),
@@ -93,7 +94,7 @@ g1 <- ggplot(observed_nsgtf) +
 
 g2 <- ggplot(corrected_nsgtf) +
   geom_line(aes(time, density, col="Delta"), lwd=1) +
-  geom_line(data=corrected_sgtf, aes(time, density, col="Omicron"), lwd=1) +
+  geom_line(data=corrected_sgtf, aes(time, density, col="Omicron"), lwd=1, lty=2) +
   scale_x_continuous("Forward incubation period (days)", expand=c(0, 0), limits=c(0, 14)) +
   scale_y_continuous("Density (1/day)", expand=c(0, 0), limits=c(0, 0.23)) +
   scale_color_manual(values=c("black", "orange")) +
@@ -106,20 +107,22 @@ g3 <- ggplot(corrected_mean_nsgtf) +
   geom_line(aes(r, median), lwd=1) +
   geom_line(aes(r, lwr), lwd=1)  +
   geom_line(aes(r, upr), lwd=1) +
-  geom_line(data=corrected_mean_sgtf, aes(r-0.2, median), lty=2, col="orange", lwd=1)  +
-  geom_line(data=corrected_mean_sgtf, aes(r-0.2, lwr), lty=2, col="orange", lwd=1)  +
-  geom_line(data=corrected_mean_sgtf, aes(r-0.2, upr), lty=2, col="orange", lwd=1) +
-  scale_x_continuous("Delta growth rate (1/day)",
[…]

### 2022-03-16 fit_lognormal_base_comb_between.R

@@ -1,0 +1,47 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+load("rdaout/calculate_incubation_mle.rda")
+
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+rho <- 0.75
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_comb_sgtf_between <- serialdata %>%
+  filter(strain=="SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_between <- serialdata %>%
+  filter(strain=="non-SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+data_comb_sgtf_between <- rep(serialdata_comb_sgtf_between$serial, serialdata_comb_sgtf_between$n)
+data_comb_nsgtf_between <- rep(serialdata_comb_nsgtf_between$serial, serialdata_comb_nsgtf_between$n)
+
+fit_lognormal_base_comb_sgtf_between <- fitfun_lognormal(data=data_comb_sgtf_between, 
+                                                      logmean_gen=logmean_inc_sgtf, 
+                                                      logsd_gen=logsd_inc_sgtf, 
+                                                      logmean_inc=logmean_inc_sgtf, 
+                                                      logsd_inc=logsd_inc_sgtf, 
+                                                      rho=rho, 
+                                                      r=r_sgtf)
+
+fit_lognormal_base_comb_nsgtf_between <- fitfun_lognormal(data=data_comb_nsgtf_between, 
+                                                       logmean_gen=logmean_inc_nsgtf, 
+                                                       logsd_gen=logsd_inc_nsgtf, 
+                                                       logmean_inc=logmean_inc_nsgtf, 
+                                                       logsd_inc=logsd_inc_nsgtf, 
+                                                       rho=rho, 
+                                                       r=r_nsgtf)
+
+save("fit_lognormal_base_comb_sgtf_between", "fit_lognormal_base_comb_nsgtf_between", file="fit_lognormal_base_comb_between.rda")

### 2022-03-16 fit_lognormal_base_comb_within.R

@@ -1,0 +1,47 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+load("rdaout/calculate_incubation_mle.rda")
+
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+rho <- 0.75
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
+data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
+
+fit_lognormal_base_comb_sgtf_within <- fitfun_lognormal(data=data_comb_sgtf_within, 
+                                                      logmean_gen=logmean_inc_sgtf, 
+                                                      logsd_gen=logsd_inc_sgtf, 
+                                                      logmean_inc=logmean_inc_sgtf, 
+                                                      logsd_inc=logsd_inc_sgtf, 
+                                                      rho=rho, 
+                                                      r=r_sgtf)
+
+fit_lognormal_base_comb_nsgtf_within <- fitfun_lognormal(data=data_comb_nsgtf_within, 
+                                                       logmean_gen=logmean_inc_nsgtf, 
+                                                       logsd_gen=logsd_inc_nsgtf, 
+                                                       logmean_inc=logmean_inc_nsgtf, 
+                                                       logsd_inc=logsd_inc_nsgtf, 
+                                                       rho=rho, 
+                                                       r=r_nsgtf)
+
+save("fit_lognormal_base_comb_sgtf_within", "fit_lognormal_base_comb_nsgtf_within", file="fit_lognormal_base_comb_within.rda")

### 2022-03-16 fit_lognormal_comb_between_r.R

@@ -1,0 +1,133 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+source("sample_incubation.R")
+load("rdaout/fit_lognormal_base_comb_between.rda")
+
+r_nsgtf <- seq(-0.1, 0, length.out=11)
+r_sgtf <- seq(0.1, 0.2, length.out=11)
+
+rho <- 0.75
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_comb_sgtf_between <- serialdata %>%
+  filter(strain=="SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_between <- serialdata %>%
+  filter(strain=="non-SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+data_comb_sgtf_between <- rep(serialdata_comb_sgtf_between$serial, serialdata_comb_sgtf_between$n)
+data_comb_nsgtf_between <- rep(serialdata_comb_nsgtf_between$serial, serialdata_comb_nsgtf_between$n)
+
+fit_lognormal_r_comb_nsgtf_between <- lapply(r_nsgtf, function(r) {
+  print(r)
+  moment_0_nsgtf <- integrate(function(z) {
+    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_1_nsgtf <- integrate(function(z) {
+    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_2_nsgtf <- integrate(function(z) {
+    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
+  var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
+  
[…]

### 2022-03-16 fit_lognormal_comb_within_r.R

@@ -1,0 +1,133 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+source("sample_incubation.R")
+load("rdaout/fit_lognormal_base_comb_within.rda")
+
+r_nsgtf <- seq(-0.1, 0, length.out=11)
+r_sgtf <- seq(0.1, 0.2, length.out=11)
+
+rho <- 0.75
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
+data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
+
+fit_lognormal_r_comb_nsgtf_within <- lapply(r_nsgtf, function(r) {
+  print(r)
+  moment_0_nsgtf <- integrate(function(z) {
+    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_1_nsgtf <- integrate(function(z) {
+    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_2_nsgtf <- integrate(function(z) {
+    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
+  var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
+  
[…]

### 2022-03-16 figure_diagram.R
new file

@@ -1,0 +1,144 @@
+library(ggplot2); theme_set(theme_bw(base_family="Times"))
+library(egg)
+
+size <- 4
+stroke <- 1
+epsilon <- 0.08
+
+g1 <- ggplot() +
+  annotate("text", x=5, y=0.55, label="Same symptom onset time", family="Times") +
+  geom_rect(aes(xmin=4.3, xmax=5.7, ymin=-3.7, ymax=0.4), col="black", lty=3, fill=NA, lwd=stroke) +
+  geom_segment(aes(x=-3, xend=10, y=0, yend=0), col="black", lwd=stroke) +
+  geom_segment(aes(x=-3, xend=10, y=-1.5, yend=-1.5), col="black", lwd=stroke) +
+  geom_segment(aes(x=-3, xend=10, y=-3, yend=-3), col="black", lwd=stroke) +
+  geom_segment(aes(x=-3, xend=10, y=-4.5, yend=-4.5), col="black", lwd=stroke) +
+  geom_segment(aes(x=3, xend=3, y=0, yend=-1.5), col="black", lty=2) +
+  geom_segment(aes(x=2.5, xend=2.5, y=-3, yend=-4.5), col="black", lty=2) +
+  geom_rect(aes(xmin=-4, xmax=-2, ymin=-0.2, ymax=0.2), fill="gray90", col="black", size=stroke) +
+  annotate("text", x=-3, y=0, label="Infector A", family="Times") +
+  geom_rect(aes(xmin=-4, xmax=-2, ymin=-1.7, ymax=-1.3), fill="gray90", col="black", size=stroke) +
+  annotate("text", x=-3, y=-1.5, label="Infectee B", family="Times") +
+  geom_rect(aes(xmin=-4, xmax=-2, ymin=-0.2-3, ymax=0.2-3), fill="gray90", col="black", size=stroke) +
+  annotate("text", x=-3, y=0-3, label="Infector C", family="Times") +
+  geom_rect(aes(xmin=-4, xmax=-2, ymin=-1.7-3, ymax=-1.3-3), fill="gray90", col="black", size=stroke) +
+  annotate("text", x=-3, y=-1.5-3, label="Infectee D", family="Times") +
+  annotate("text", x=0, y=0.25, label="Infection", family="Times") +
+  annotate("text", x=3, y=0.25, label="Transmission, A to B", family="Times") +
+  annotate("text", x=5, y=0.25, label="Symptom", family="Times") +
[…]

### 2022-03-17 omicron-generation.tex
update

@@
 When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
-Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C).
+Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C--D).
 % \jd{Do we need the ff.~s.? What does it add to the Fig?}
 % \swp{I think people like to see actual values...}
-For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$, while
-the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$.
+For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$ (\fref{incubation}C), while
+the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$ (\fref{incubation}D).
 
@@
 (B) Forward incubation-period distributions assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta (black) and Omicron (orange) variants, respectively.
-(C) Corrected estimates of the forward incubation-period distributions for different assumptions about the growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
-Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
+(C--D) Corrected estimates of the mean forward incubation-period for different assumptions about the growth rates of the Delta (C) and Omicron variants (D).
+Lines represent median estimates.
+Shaded regions represent the corresponding 95\% confidence intervals.
 \label{fig:incubation}

### 2022-03-17 figure_compare_main.R

@@ -1,0 +1,208 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(gridExtra)
+source("serialfun.R")
+load("rdaout/calculate_incubation_mle.rda")
+load("rdaout/fit_lognormal_base_comb_within.rda")
+load("rdaout/fit_lognormal_base_comb_between.rda")
+load("rdaout/fit_lognormal_comb_within_r.rda")
+load("rdaout/fit_lognormal_comb_between_r.rda")
+# load("rdaout/fit_lognormal_base.rda")
+# load("rdaout/fit_lognormal_r.rda")
+# load("rdaout/fit_lognormal_r_50_between.rda")
+# load("rdaout/fit_lognormal_r_51_between.rda")
+# load("rdaout/fit_lognormal_r_51.rda")
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+xvec <- seq(-6, 16, by=0.2)
+
+ivec <- seq(0, 16, by=0.2)
+gvec <- seq(0, 16, by=0.2)
+
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_sgtf_between <- serialdata %>%
+  filter(strain=="SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_between <- serialdata %>%
+  filter(strain=="non-SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serial_density_lognormal_nsgtf_within <- data.frame(
+  x=xvec,
+  y=do.call(serialfun_lognormal, c(list(x=xvec), as.list(coef(fit_lognormal_base_comb_nsgtf_within)))),
+  type="Delta"
+)
+
+serial_density_lognormal_sgtf_within <- data.frame(
+  x=xvec,
[…]

### 2022-03-17 figure_household.R

@@ -1,37 +1,0 @@
-library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
-load("rdaout/fit_lognormal_base.rda")
-load("rdaout/simulate_generation_household.rda")
-
-logmean_nsgtf <- coef(fit_lognormal_base_50_nsgtf_within)[1]
-logsd_nsgtf <- coef(fit_lognormal_base_50_nsgtf_within)[2]
-mean_nsgtf <- exp(logmean_nsgtf + logsd_nsgtf^2/2)
-
-logmean_sgtf <- coef(fit_lognormal_base_50_sgtf_within)[1]
-logsd_sgtf <- coef(fit_lognormal_base_50_sgtf_within)[2]
-mean_sgtf <- exp(logmean_sgtf + logsd_sgtf^2/2)
-
-simulate_generation_household2 <- simulate_generation_household %>%
-  mutate(
-    k=paste0("k=", k),
-    ratio=paste0("Reproduction number ratio: ", ratio),
-    type=factor(type, labels=c("Egocentric", "Local"))
-  )
-
-g1 <- ggplot(simulate_generation_household2) +
-  geom_hline(yintercept=mean_nsgtf, lty = "dotted") +
-  geom_hline(yintercept=mean_sgtf, lty = "dotted", col="orange") +
-  geom_ribbon(aes(household, ymin=lwr, ymax=upr, fill=strain, col=strain, lty=type), alpha=0.1) +
-  geom_line(aes(household, mean, col=strain, lty=type)) +
-  scale_x_continuous("Network sizes", breaks=1:5*4) +
-  scale_y_continuous("Mean generation interval (days)") +
-  scale_color_manual("Variant", values=c("black", "orange")) +
-  scale_fill_manual("Variant", values=c("black", "orange")) +
-  scale_linetype_discrete("Simulation") +
-  facet_grid(ratio~k) +
-  theme(
-    panel.grid = element_blank(),
-    strip.background = element_blank()
-  )
-
-ggsave("figure_household.pdf", g1, width=8, height=6)

### 2022-03-17 figure_incubation.R

@@ -104,9 +104,8 @@ g2 <- ggplot(corrected_nsgtf) +
   )
 
 g3 <- ggplot(corrected_mean_nsgtf) +
+  geom_ribbon(aes(r, ymin=lwr, ymax=upr), alpha=0.2) + 
   geom_line(aes(r, median), lwd=1) +
-  geom_line(aes(r, lwr), lwd=1)  +
-  geom_line(aes(r, upr), lwd=1) +
   scale_x_continuous("Delta growth rate (1/day)") +
   scale_y_continuous("Mean forward incubation period (days)", limits=c(3.3, 5.6)) +
   theme(
@@ -114,9 +113,8 @@ g3 <- ggplot(corrected_mean_nsgtf) +
   )
 
 g4 <- ggplot(corrected_mean_sgtf) +
+  geom_ribbon(aes(r, ymin=lwr, ymax=upr), alpha=0.2, fill="orange") + 
   geom_line(aes(r, median), lty=2, col="orange", lwd=1) +
-  geom_line(aes(r, lwr), lty=2, col="orange", lwd=1)  +
-  geom_line(aes(r, upr), lty=2, col="orange", lwd=1) +
   scale_x_continuous("Omicron growth rate (1/day)") +
   scale_y_continuous("Mean forward incubation period (days)", limits=c(3.3, 5.6)) +
   theme(

### 2022-03-18 omicron-generation.tex
lots of changes

@@
 Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
-For illustrative purposes, we first focus on within-household serial intervals from infectors who developed symptoms during week 50 (13-–19 December, 2021)---\citep{backer2021omicron} reported shorter mean serial interval of the Omicron variant (3.5 days) than that of the Delta variant (4.1 days) for this cohort (\fref{serial}A).
-When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a larger difference (\fref{serial}B): 3.3 days (95\% CI: 2.8--3.7 days) for the Omicron variant and 4.2 days (95\% CI: 4.0--4.4 days) for the Delta variant.
-Notably, both variants have similar mode, but the transmissibility of the omicron variant is estimated to decay much faster (\fref{serial}B).
-Note that the mean forward generation interval of the Omicron (respectively, Delta) variant is shorter (longer) than its mean serial interval due to the effect of a growing (declining) epidemic.
-The estimated differences of the mean generation interval increase if we assume that generation intervals and incubation periods are less correlated (and therefore, lower values of $\rho$; Supplementary Figure S1).
-On the other hand, neglecting growth-rate differences (i.e., assuming $r= 0/\mathrm{day}$ for both variants) leads to the underestimation of these differences (Supplementary Figure S2): 3.5 days (95\% CI: 3.1--3.9 days) for the Omicron variant and 4.1 days (95\% CI: 3.9--4.3 days).
+For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13-–19 December, 2021).
+For within-household transmission pairs (\fref{serial}A), the Omicron variant has shorter mean serial interval (3.1 days; 95\% CI: 2.9--3.3 days) than that of the Delta variant (3.7 days; 95\% CI: 3.5--3.8 days).
+When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a slightly larger difference (\fref{serial}B):
+3.0 days (95\% CI: 2.7--3.2 days) for the Omicron variant and 3.8 days (95\% CI: 3.7--4.0 days) for the Delta variant.
+Across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants, we estimate robust differences in their mean generation intervals (\fref{serial}C--D).
+Assuming lower values of the correlation between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
 
-We then estimate the mean forward generation intervals for all other cohorts, stratified by the week of symptom onset (weeks 50 and 51) and the type of transmission (within and between household(s)).
-As before (see \fref{incubation}C), we repeat the analysis for different values of growth rates for the Delta ($r=-0.1/\mathrm{days}$ to $0/\mathrm{days}$) and Omicron ($r=0.1/\mathrm{days}$ to $0.2/\mathrm{days}$) variants and plot the maximum likelihood estimates and the associated CIs.
-We estimate shorter mean forward generation intervals for the Omicron variant across all stratification (\fref{serial}C), but the differences in the mean generation intervals are unclear for between-household pairs.
-Consistent with the original study, which also reported shorter mean forward serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean forward generation intervals for between-household pairs.
-We also estimate a reduction in the mean forward generation intervals from week 50 to week 51.
+Similar pictures arise for between-household transmission pairs, but the differencse in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
+Consistent with the original study, which also reported shorter mean serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean generation intervals for between-household Delta pairs.
+While the difference in mean generation intervals is larger, there is greater uncertainty in their mean estimates (\fref{serial}F): 2.9 days (95\% CI: 2.5--3.3 days) for the Omicron variant and 3.5 days (95\% CI: 3.2--3.8 dyas) for the Delta variant.
+Once again, these patterns are robust across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants (\fref{serial}G--H).
+
+In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (weeks 50 and 51).
+While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50.
+We also estimate a reduction in the mean forward generation intervals from week 50 to week 51 (especially for the Delta variant).
 
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_compare.pdf}
+\includegraphics[width=\textwidth]{figure_compare_main.pdf}
 \caption{
 \textbf{Estimated forward generation-interval distributions of Delta and Omicron variants.}
-(A) Observed and fitted forward serial-interval distributions for within-household transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
-Serial intervals are calculated for infectors who developed symptoms on week 50 (13 and 19 December, 2021).
+(A, E) Observed and fitted forward serial-interval distributions for within-household (A) and between-household (E) transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
+Serial intervals are calculated for infectors who developed symptoms on weeks 50 and 51 (13--26 December, 2021).
 Points represent the observed data.
 Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant.
-Bivariate lognormal distributions with correlation coefficient of 0.75 are used to model relationships between incubation and generation intervals.
-(B) Estimated forward generation-interval distributions for within-household transmission pairs in the Netherlands.
-(C) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines) stratified by the week of infectors' symptom onset and the type of transmission pairs.
+(B, F) Estimated forward generation-interval distributions for within-household (B) and between-household (F) transmission pairs in the Netherlands.
+(C, D, G, H) Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta (C, G) and Omicron variants (G, H) for within-household (C, D) and between-household (G, H) transmission pairs.
+Lines represent median estimates.
+Shaded regions represent the corresponding 95\% confidence intervals.
 \label{fig:serial}
@@
 
-Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.83 (95\% CI: 1.65--2.05) to 1.17 (95\% CI: 1.00--1.36) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
-On the other hand, the reproduction number of the Delta variant decreased from 0.89 (95\% CI: 0.82--0.97) to 0.68 (95\% CI: 0.62--0.74) between December 5, 2021, and January 9, 2022, and increased back up to 0.82 (95\% CI: 0.72--0.93) by January 23, 2022 (\fref{reproduction}A).
-We estimate the reproduction number ratios stayed roughly constant at around 2.2 (95\% CI: 2.0--2.5) between December 12--26, 2021, and slowly decreased to 1.4 (95\% CI: 1.2--1.7).
+Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.73 (95\% CI: 1.59--1.89) to 1.14 (95\% CI: 1.00--1.32) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
+On the other hand, the reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
+We estimate the reproduction number ratios stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
 However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
-In this case, the reproduction ratio decreases from 2.5 (95\% CI: 2.2--2.8) to 1.5 (95\% CI: 1.2--1.8), corresponding to roughly 7--13\% bias.
+In this case, the reproduction ratio decreases from 2.38 (95\% CI: 2.13--2.67) to 1.43 (95\% CI: 1.17--1.75), corresponding to roughly 4--13\% bias.
+Using between-household generation intervals also gives similar conclusions about changes and biases in the reproduction number estimates (Supplementary Figure S3).
 
@@
 The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
-The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the generation-interval distributions for within-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
+The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the forward generation-interval distributions for within-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
 Purple lines represent the ratio between the effective reproduction numbers of the Delta and Omicron variants.
@@
 The original analysis detailing the data set previously reported shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
-Accounting for differences in epidemic growth rates, we find similar incubation-period distributions for both variants but a considerably shorter (0.5--0.9 days) mean generation interval for the Omicron variant.
-The mean forward generation interval of both variants further decreased between week 50 (13--19 December, 2021) and week 51 (20--26 December, 2021).
-Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.2-fold to 1.4-fold between early December and late January. 
-Neglecting differences in the generation-interval distributions can result in $\approx 10\%$ bias in estimates of the transmission advantage.
+Accounting for differences in epidemic growth rates, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant.
+Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January. 
+Neglecting differences in the generation-interval distributions can bias estimates of the transmission advantage.
 
@@
 These effects are also pertinent to all epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
+Our sensitivity analysis also shows that the assumptions about the correlation between incubation periods and generation intervals can also have important effects on the estimates of the generation-interval distributions (Supplementary Figure S1).
+
+A few studies have suggesed the the incubation period of the Omicron variant may be shorter than that of the Delta variant.
+The median estimates of the Omicron incubation period typically range between 3--4 days, consistent with earlier findings of \citep{backer2021omicron}. 
+However, these data were collected when the number of Omicron infections was growing rapidly \citep{jansen2021investigation,song2022serial}, suggesting that they may have been subject to similar biases.
+On the other hand, incubation-period estimates based on individuals who were exposed from the same event are likely more reliable (because they look forward in time):
+\cite{brandal2021outbreak} estimated the median incubation period of the Omicron variant to be 3 days among those who attended the same holiday party ($n=117$) on 26 November 2021 in Norway. 
+However, we cannot rule out the possibility that some of these attendees were infected prior to the party given that some individuals had COVID-like symptoms prior to the party with at least 96 of the attendees sharing offices; neglecting these factors can lead to underestimation of the mean incubation period.
+Systematic comparisons of data collection methods and epidemiological contexts are needed to properly assess the differences in incubation period distributions of the Delta and Omicron variants.
 
@@
 The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
+Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
 Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shorter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.
 
-While our study indicate that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how much their infectiousness profiles differ intrinsically.
-In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their intrinsic infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual intrinsic generation-interval distribution, which may have a longer mean).
+While our study indicate that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how much their infectiousness profiles differ biologically.
+In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
+In addition, the unmitigated generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating ``unmitigated'' or ``intrinsic'' generation-interval distributions of SARS-CoV-2 variants are expected to be a difficult problem as it requires early data during which interventions and awareness levels were minimal \citep{sender2021unmitigated}.
+Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual unmitigated generation-interval distribution, which may have a longer mean).
 
@@
 Lines represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
-For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during week 50.
+For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 and 51.
 }
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_compare_r_0.pdf}
+\includegraphics[width=\textwidth]{figure_compare_stratified.pdf}
 \caption{
-\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants while neglecting growth-rate differences.}
+\textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
 See Figure 4 in the main text for figure caption.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{figure_reproduction_advantage_between.pdf}
+\caption{
+\textbf{Estimated time-varying reproduction number advantages of the Omicron variant using between-household generation-interval distributions.}
+See Figure 5 in the main text for figure caption.
 }

### 2022-03-18 figure_compare_main.R

@@ -10,11 +10,12 @@ load("rdaout/fit_lognormal_base_comb_within.rda")
 load("rdaout/fit_lognormal_base_comb_between.rda")
 load("rdaout/fit_lognormal_comb_within_r.rda")
 load("rdaout/fit_lognormal_comb_between_r.rda")
-# load("rdaout/fit_lognormal_base.rda")
-# load("rdaout/fit_lognormal_r.rda")
-# load("rdaout/fit_lognormal_r_50_between.rda")
-# load("rdaout/fit_lognormal_r_51_between.rda")
-# load("rdaout/fit_lognormal_r_51.rda")
+
+filter(fit_lognormal_r_comb_nsgtf_within, param=="mean", r==-0.05)
+filter(fit_lognormal_r_comb_sgtf_within, param=="mean", round(r, 2)==0.15)
+
+filter(fit_lognormal_r_comb_nsgtf_between, param=="mean", r==-0.05)
+filter(fit_lognormal_r_comb_sgtf_between, param=="mean", round(r, 2)==0.15)
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
@@ -30,12 +31,16 @@ serialdata_comb_sgtf_within <- serialdata %>%
     n=sum(n)
   )
 
+t.test(rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n))
+
 serialdata_comb_nsgtf_within <- serialdata %>%
   filter(strain=="non-SGTF", household=="within") %>%
   group_by(serial, strain, household) %>%
   summarize(
     n=sum(n)
   )
+
+t.test(rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n))
 
 serialdata_comb_sgtf_between <- serialdata %>%
   filter(strain=="SGTF", household=="between") %>%
@@ -44,12 +49,16 @@ serialdata_comb_sgtf_between <- serialdata %>%
     n=sum(n)
   )
 
+t.test(rep(serialdata_comb_sgtf_between$serial, serialdata_comb_sgtf_between$n))
+
 serialdata_comb_nsgtf_between <- serialdata %>%
   filter(strain=="non-SGTF", household=="between") %>%
   group_by(serial, strain, household) %>%
   summarize(
     n=sum(n)
   )
+
[…]

### 2022-03-18 figure_compare_rho.R

@@ -4,15 +4,15 @@ library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(egg)
 library(gridExtra)
-load("rdaout/fit_lognormal_base_rho.rda")
+load("rdaout/fit_lognormal_base_comb_rho.rda")
 
-g1 <- ggplot(filter(fit_lognormal_base_rho_50_nsgtf_within, param=="mean"))  +
+g1 <- ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
   geom_line(aes(rho, est), lwd=1, col="black") +
   geom_line(aes(rho, lwr), lwd=1, col="black") +
   geom_line(aes(rho, upr), lwd=1, col="black") +
-  geom_line(data=filter(fit_lognormal_base_rho_50_sgtf_within, param=="mean", type=="Omicron"), aes(rho, est), lty=2, col="orange", lwd=1) +
-  geom_line(data=filter(fit_lognormal_base_rho_50_sgtf_within, param=="mean", type=="Omicron"), aes(rho, lwr), lty=2, col="orange", lwd=1) +
-  geom_line(data=filter(fit_lognormal_base_rho_50_sgtf_within, param=="mean", type=="Omicron"), aes(rho, upr), lty=2, col="orange", lwd=1) +
+  geom_line(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, est), lty=2, col="orange", lwd=1) +
+  geom_line(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, lwr), lty=2, col="orange", lwd=1) +
+  geom_line(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, upr), lty=2, col="orange", lwd=1) +
   scale_x_continuous("Lognormal correlation coefficient") +
   scale_y_continuous("Mean generation interval (days)") +
   theme(

### 2022-03-18 figure_compare_stratified.R

@@ -1,0 +1,95 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(gridExtra)
+source("serialfun.R")
+load("rdaout/fit_lognormal_r.rda")
+load("rdaout/fit_lognormal_r_50_between.rda")
+load("rdaout/fit_lognormal_r_51_between.rda")
+load("rdaout/fit_lognormal_r_51.rda")
+
+g1 <- ggplot(filter(fit_lognormal_r_50_nsgtf_within, param=="mean")) +
+  annotate("text", x=-Inf, y=Inf, hjust=-0.05, vjust=1.5, label="Week 50, within-household") +
+  geom_line(aes(r, est), lwd=1) +
+  geom_ribbon(aes(r, ymin=lwr, ymax=upr), alpha=0.2) + 
+  scale_x_continuous("Delta growth rate (1/day)") +
+  scale_y_continuous("Mean within-household\ngeneration interval (days)", limits=c(2.1, 4.5)) +
+  theme(
+    panel.grid = element_blank()
+  )
+
+g2 <- ggplot(filter(fit_lognormal_r_50_sgtf_within, param=="mean")) +
+  annotate("text", x=-Inf, y=Inf, hjust=-0.05, vjust=1.5, label="Week 50, within-household") +
+  geom_line(aes(r, est), lwd=1, col="orange", lty=2) +
+  geom_ribbon(aes(r, ymin=lwr, ymax=upr), alpha=0.2, fill="orange") + 
+  scale_x_continuous("Omicron growth rate (1/day)") +
+  scale_y_continuous("Mean within-household\ngeneration interval (days)", limits=c(2.1, 4.5)) +
+  theme(
+    panel.grid = element_blank()
+  )
+
+g3 <- ggplot(filter(fit_lognormal_r_51_nsgtf_within, param=="mean")) +
+  annotate("text", x=-Inf, y=Inf, hjust=-0.05, vjust=1.5, label="Week 51, within-household") +
+  geom_line(aes(r, est), lwd=1) +
+  geom_ribbon(aes(r, ymin=lwr, ymax=upr), alpha=0.2) + 
+  scale_x_continuous("Delta growth rate (1/day)") +
+  scale_y_continuous("Mean within-household\ngeneration interval (days)", limits=c(2.1, 4.5)) +
+  theme(
[…]

### 2022-03-18 figure_reproduction_advantage.R

@@ -9,7 +9,7 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(mgcv)
 library(mvtnorm)
 library(egg)
-load("rdaout/fit_lognormal_base.rda")
+load("rdaout/fit_lognormal_base_comb_within.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 
@@ -86,8 +86,8 @@ gfit2_Xp <- (gfit2_p2 - gfit2_p1) / 0.01/7
 set.seed(101)
 gfit1_sim <- rmvnorm(nsample, mean = coef(gfit1), sigma = vcov(gfit1))
 gfit2_sim <- rmvnorm(nsample, mean = coef(gfit2), sigma = vcov(gfit2))
-lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_nsgtf_within)[1:2], sigma=vcov(fit_lognormal_base_50_nsgtf_within))
-lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_sgtf_within)[1:2], sigma=vcov(fit_lognormal_base_50_sgtf_within))
+lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_nsgtf_within)[1:2], sigma=vcov(fit_lognormal_base_comb_nsgtf_within))
+lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_sgtf_within)[1:2], sigma=vcov(fit_lognormal_base_comb_sgtf_within))
 
 gfit1_post <- apply(gfit1_sim, 1, function(x) {gfit1_Xp %*% x})
 gfit2_post <- apply(gfit2_sim, 1, function(x) {gfit2_Xp %*% x})

### 2022-03-18 figure_reproduction_advantage_between.R

@@ -9,7 +9,7 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(mgcv)
 library(mvtnorm)
 library(egg)
-load("rdaout/fit_lognormal_base_50_between.rda")
+load("rdaout/fit_lognormal_base_comb_between.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 
@@ -86,8 +86,8 @@ gfit2_Xp <- (gfit2_p2 - gfit2_p1) / 0.01/7
 set.seed(101)
 gfit1_sim <- rmvnorm(nsample, mean = coef(gfit1), sigma = vcov(gfit1))
 gfit2_sim <- rmvnorm(nsample, mean = coef(gfit2), sigma = vcov(gfit2))
-lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_nsgtf_between)[1:2], sigma=vcov(fit_lognormal_base_50_nsgtf_between))
-lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_50_sgtf_between)[1:2], sigma=vcov(fit_lognormal_base_50_sgtf_between))
+lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_nsgtf_between)[1:2], sigma=vcov(fit_lognormal_base_comb_nsgtf_between))
+lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_sgtf_between)[1:2], sigma=vcov(fit_lognormal_base_comb_sgtf_between))
 
 gfit1_post <- apply(gfit1_sim, 1, function(x) {gfit1_Xp %*% x})
 gfit2_post <- apply(gfit2_sim, 1, function(x) {gfit2_Xp %*% x})
@@ -165,26 +165,6 @@ g1 <- ggplot(filter(advantagesumm, key %in% c("R_advantage", "R_delta", "R_omicr
     legend.position = "none"
   )
 
-advantagesumm2 <- advantagesumm %>%
-  select(-lwr, -upr) %>%
-  spread(key, median)
+gfinal <- ggarrange(g1, g2, nrow=1, labels=c("A", "B"), draw=FALSE)
 
-g3 <- ggplot(advantagesumm2) +
-  geom_path(aes(R_delta, R_advantage), lwd=1, arrow = arrow(length=unit(0.30,"cm"))) +
-  scale_x_log10("Delta reproduction number") +
[…]

### 2022-03-18 fit_lognormal_base_comb_rho.R

@@ -1,0 +1,95 @@
+library(dplyr)
+library(readxl)
+source("serialfun.R")
+source("fitfun.R")
+load("rdaout/calculate_incubation_mle.rda")
+load("rdaout/fit_lognormal_base_comb_within.rda")
+
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+rhovec <- seq(0.5, 0.85, length.out=11)
+
+serialdata <- read_xlsx("serial-netherlands.xlsx")
+
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
+data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
+
+fit_lognormal_base_rho_comb_nsgtf_within <- lapply(rhovec, function(rho) {
+  print(rho)
+  ff <- fitfun_lognormal(data=data_comb_nsgtf_within, 
+                         logmean_gen=coef(fit_lognormal_base_comb_nsgtf_within)[[1]], 
+                         logsd_gen=coef(fit_lognormal_base_comb_nsgtf_within)[[2]], 
+                         logmean_inc=logmean_inc_nsgtf, 
+                         logsd_inc=logsd_inc_nsgtf, 
+                         rho=rho, 
+                         r=r_nsgtf)
+  
+  mean <- exp(coef(ff)[[1]]+coef(ff)[[2]]^2/2)
+  
+  cc <- confint(ff, method="quad")
+  
+  tmp <- c(
+    exp(coef(ff)[[1]]) * exp(coef(ff)[[2]]^2/2),
+    exp(coef(ff)[[1]]) * exp(coef(ff)[[2]]^2/2) * coef(ff)[[2]]
+  )
+  
+  sd_mean <- sqrt(c(tmp %*% vcov(ff) %*% tmp))
+  
+  data.frame(
[…]

### 2022-03-18 omicron-generation.tex
update output

@@
 Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
-When we account for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
+When we account for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
 The differences in generation intervals may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
-Neglecting differences in the generation-interval distributions can further bias in estimates of transmission advantages of the Omicron variant.
+Neglecting differences in the generation-interval distributions can further bias in estimates of the reproduction advantage of the Omicron variant.
 
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction ratios ($> 4$ fold) in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate that the reproduction number ratio of 2.8 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+With our generation-interval estimates, we estimate that the reproduction number ratio of 2.6 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
 We also show that both growth-rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.

### 2022-03-19 omicron-generation.tex
almost done

--- 2022-03-18 omicron-generation.tex
@@
 
-Estimating the differences in the incubation-period and serial-interval distributions of SARS-CoV-2 variants is critical to understanding changes in their transmission dynamics and therefore controlling their outbreaks.
-However, these comparisons are inherently difficult due to their dependence on epidemic growth rates---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have shorter incubation periods.
-Here, we re-analyze incubation-period and serial-interval data collected by Backer \textit{et al.}, describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
+Estimating the differences in the incubation-period, serial-interval, and generation-interval distributions of SARS-CoV-2 variants is critical to understanding their transmission dynamics and outbreak control.
+However, estimating and comparing the timing of infection and transmission is inherently difficult due to the impact of epidemic dynamics---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have been infected recently.
+Here, we re-analyze incubation-period and serial-interval data describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
 Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
-When we account for growth-rate differences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.3 days; 95\% CI: 2.8--3.7 days) than for the Delta variant (4.2 days; 95\% CI: 4.0--4.4 days).
+When we account for the impact of growth-rate differences on inferences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
 The differences in generation intervals may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
-Neglecting differences in the generation-interval distributions can further bias in estimates of transmission advantages of the Omicron variant.
+As we show, incorporating differences in the generation-interval distributions can reduce biases in estimates of the reproduction advantage of the Omicron and other emerging variants.
 
@@
 
-Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the ongoing COVID-19 pandemic.
+Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the COVID-19 pandemic.
 Transmission advantages of invading variants are typically characterized by the ratios of reproduction numbers, $\Rinv/\Rres$, and the differences in growth rates, $\rinv-\rres$.
@@
 
-Estimating generation-interval is hard, largely due to difficulties in observing actual infection events.
-Many researchers primarily focus on comparisons of other transmission intervals, such as time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
-However, transmission-interval distributions are subject to dynamical effects which can bias estimation.
+Estimating the generation-intervaldistribution is challenging, in part, due to difficulties in observing actual infection events.
+Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
+Each of these transmission-interval distributions are subject to dynamical effects which, can bias estimation.
 For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
-Therefore, observed differences in transmission-interval distributions between variants can give a biased picture of true differences in their generation-interval distributions, especially if their growth rates differ.
+Therefore, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions, especially if their growth rates differ.
 
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The authors of the original article reported shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly omicron during the study period) and without (mostly delta), but did not consider growth-rate differences in their inference.
-Here, we discuss the epidemiological context in the Netherlands during the study period and take dynamics into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
+The study found shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) and without (mostly Delta), but did not consider growth-rate differences in their inference.
+Here, we take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
 
@@
 We analyze time series of reported COVID-19 cases (\url{https://data.rivm.nl/covid-19/}) and proportions of SARS-CoV-2 variants detected (\url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}) from the Netherlands between 29 November 2021 and 30 January 2022.
-Both data are publicly available on the National Institute for Public Health and the Environment (RIVM) website.
+Data sets are publicly available on the National Institute for Public Health and the Environment (RIVM) website.
 
-Serial interval data are taken from the supplementary material of \cite{backer2021omicron}.
+Serial interval data are taken from \cite{backer2021omicron} (see original article for details of data collection).
 The data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates or symptom onset dates.
-The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (weeks 50 and 51), and the type of transmission (within or between households).
-In the main text, we combine data from weeks 50 and 51 and instead present a stratified analysis in Supplementary Material.
+The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (weeks 50 and 51 of 2021), and the type of transmission (within or between households).
+In the main text, we combine data from weeks 50 and 51 of 2021 and present a stratified analysis in Supplementary Material.
 For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
-Incubation period data are not publicly available with the original article; instead, we rely on previous estimates by \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
-See original article for details of data collection.
+Incubation period data are not publicly available with the original article; instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
 
@@
 
-\cite{backer2021omicron} estimated the incubation periods from 513 individuals (258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022.
-They used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates, using a uniform distribution, and compares this to a known symptom-onset time.
+The incubation-period distributions from 513 individuals (258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022, were previously reported in \cite{backer2021omicron}.
+\cite{backer2021omicron} used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates, using a uniform distribution, and compares this to a known symptom-onset time.
 
-In practice, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward.
+In practice, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward \citep{park2021forward}.
 The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
@@
 
-In particular, when incidence of infection is growing exponentially, we are more likely to observe shorter backward incubation periods because there will be relatively more individuals who were infected recently.
+In particular, when incidence of infection is growing exponentially, we are more likely to observe backward incubation periods that are shorter than the corresponding forward incubation periods because there will be relatively more individuals who were infected recently.
 Assuming that incidence is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
@@
 
-The symptom-based generation-interval distribution describing \eref{serial} gives a biased picture because infectors who developed symptoms at the same time will have shorter incubation periods (when epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
+The generation-interval distribution for a symptom-based cohort describing \eref{serial} is biased because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
 This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
@@
 
-When epidemic is growing is exponentially, we have two opposing effects affecting the relationship between the mean serial and generation interval.
+When an epidemic is growing exponentially, there are two opposing effects affecting the relationship between the mean serial and generation interval.
 First, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
@@
 where $i(t)$ represents incidence of infection.
-Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for week 50---when epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
+Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for weeks 50 and 51 of 2021---when the epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
 Incidence of infection is approximated by shifting the smoothed case trajectories by one week to account for reporting delays.
@@
 The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
-Finally, we find that the growth-rate difference between the Delta and Omicron variants decreased over time.
+We note that the growth-rate difference between the Delta and Omicron variants decreased over time.
 Hereafter, we choose $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as our nominal growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
@@
 
-For a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022, \citep{backer2021omicron} found longer mean (backward) incubation period for the Delta variant than for the Omicron variant (\fref{incubation}A).
+Previous analysis of a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022 found longer mean (backward) incubation period for the Delta variant than for the Omicron variant \citep{backer2021omicron} (\fref{incubation}A).
 However, these measurements were done during a period when the incidence of Omicron was increasing while the increasing of Delta was decreasing (\fref{epidemic}).
@@
 On the other hand, the reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
-We estimate the reproduction number ratios stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
+We estimate the reproduction advantage stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
 However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
@@
 
-In both cases, the decrease in the reproduction number ratio coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant;
-a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the growth-rate difference (\fref{epidemic}D).
+In both cases, the decrease in the reproduction advantage coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant;
+a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the observed growth-rate difference (\fref{epidemic}D).
 
@@
 
-We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands.
-The original analysis detailing the data set previously reported shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
-Accounting for differences in epidemic growth rates, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant.
+We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands in late 2020 and early 2021.
+The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
+Accounting for differences in epidemic growth rates, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
 Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January. 
-Neglecting differences in the generation-interval distributions can bias estimates of the transmission advantage.
+Our updated generation-interval estimates can provide more accurate understanding of epidemic dynamics and control measures.
 
@@
 A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation-period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
-Here, we build on our previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
+Here, we build on previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
 Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
-These effects are also pertinent to all epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
+These effects are also pertinent to epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
 Our sensitivity analysis also shows that the assumptions about the correlation between incubation periods and generation intervals can also have important effects on the estimates of the generation-interval distributions (Supplementary Figure S1).
@@
 Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
-We tentatively hypothesize that these differences may be primarily driven by the network effect: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections \citep{park2020inferring,hart2022generation}.
+We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
 While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
@@
 Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
-Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and shorter clearance phase of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of the generation intervals.
+Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and faster within-host clearance of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of generation intervals.
 
-While our study indicate that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how much their infectiousness profiles differ biologically.
+While our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
 In addition, the unmitigated generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating ``unmitigated'' or ``intrinsic'' generation-interval distributions of SARS-CoV-2 variants are expected to be a difficult problem as it requires early data during which interventions and awareness levels were minimal \citep{sender2021unmitigated}.
-Nonetheless, our estimates of the realized generation-interval distributions better describe the epidemic dynamics, implicitly accounting for intervention and behavioral effects, and are therefore should be used to estimate current reproduction numbers (rather than a counterfactual unmitigated generation-interval distribution, which may have a longer mean).
+Nonetheless, our estimates of the realized generation-interval distributions better describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects.
+Therefore, our generation-interval estimates are expected to give more accurate estimates of reproduction numbers.
 
 Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
-In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction number ratios; however, the degree of ratio is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
-For example, \cite{pearson2021bounding} estimated a much higher reproduction ratios ($> 4$ fold) in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate that the reproduction number ratio of 2.8 assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction number ratios with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
-We also show that both growth-rate differences and the reproduction number ratios decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction number ratios remain constant \citep{park2021roles}.
+In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
+More generally, the bias in inferring the reproduction advantage an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
+For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant ($> 4$ fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
+With our generation-interval estimates, we estimate that the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+We also show that both growth-rate differences and the reproduction advantage decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction advantage remain constant \citep{park2021roles}.
 
@@
 Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
-Our study synthesizes previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters;
-however, uncertainty remains in the intrinsic infectiousness profiles of SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.
-Future studies extending our framework and combining more detailed epidemiological data will be critical to narrowing down these uncertainties.
+Our study synthesizes a previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters.
+Similar efforts will be critical to improve estimates of the infectiousness profiles of future SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.
 

### 2022-03-21 omicron-generation.tex
update

@@
 	\textbf\newline{
-		Epidemic dynamics shape inferences of the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants
+		Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants
 	}
@@
 When we account for the impact of growth-rate differences on inferences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
-The differences in generation intervals may reflect higher reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
-As we show, incorporating differences in the generation-interval distributions can reduce biases in estimates of the reproduction advantage of the Omicron and other emerging variants.
+The differences in generation intervals may reflect higher effective reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
+Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
 
@@
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The study found shorter serial intervals (3.5 vs 4.1 days) and incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) and without (mostly Delta), but did not consider growth-rate differences in their inference.
+The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) and without (mostly Delta), but did not consider growth-rate differences in their inference:
+during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
 Here, we take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
+We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.
 
@@
 
-To estimate the differences in growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
+In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
+To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
 We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
@@
 Backward incubation periods are measured from a cohort of individuals who developed symptoms at the same time. 
-We expect the backward incubation-period distribution to be sensitive to epidemic dynamics.
-The difference arises because forward incubation periods look forward from the reference point towards symptom development, which is an individual-level process, while backward incubation periods look backwards towards an infection event, which requires an interaction with an infectious individual.
+The backward incubation-period distribution is sensitive to epidemic dynamics:
+the difference arises because forward incubation periods look forward from the reference point towards symptom development, which is an individual-level process, while backward incubation periods look backwards towards an infection event, which requires an interaction with an infectious individual.
 
@@
 for illustrative purpose, we choose values of $r$ that represent the dynamics of Delta and Omicron infections during this period and repeat the analysis across plausible ranges of $r$ (discussed later in detail).
+
+While the derivation of the forward serial-interval distribution \eref{forwardserial} may be complex, its implementation is simple.
+The main difference between our model and prevous models that neglect dynamical effects \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation} is the exponential growth term $\exp(r \alpha_1)$ and the normalization term $\phi$---it is relatively straightforward to include these terms in existing models of serial intervals.
+\cite{ferretti2020quantifying, ferretti2020timing} also included this term in their analyses of serial-interval data, but only accounted for the epidemic growth effect (and not the decay effect).
 
@@
 
-To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant.
+%A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
+
+%\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint from population-level epidemic dynamics (rather than individual-level transmission data).
+
+To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
 Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.

### 2022-03-22 omicron-generation.tex
done

@@
 We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
-%% More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood.
-%% JD: clarified previous s. Don't think we need this, with the cite.
 Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.
@@
 
-The generation-interval distribution for a symptom-based cohort describing \eref{serial} is biased because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
+The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
 This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
@@
 \caption{
-\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands.}
+\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
 (A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
@@
 (C, D, G, H) Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta (C, G) and Omicron variants (G, H) for within-household (C, D) and between-household (G, H) transmission pairs.
-Lines represent median estimates.
+Lines represent maximum likelihood estimates.
 Shaded regions represent the corresponding 95\% confidence intervals.
@@
 On the other hand, the reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
-We estimate the reproduction advantage stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
-However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
-In this case, the reproduction ratio decreases from 2.38 (95\% CI: 2.13--2.67) to 1.43 (95\% CI: 1.17--1.75), corresponding to roughly 4--13\% bias.
+We estimate the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants) stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
+However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore the reproduction advantage (\fref{reproduction}B).
+In this case, the reproduction advantage decreases from 2.38 (95\% CI: 2.13--2.67) to 1.43 (95\% CI: 1.17--1.75), corresponding to roughly 4--13\% bias.
 Using between-household generation intervals also gives similar conclusions about changes and biases in the reproduction number estimates (Supplementary Figure S3).
@@
 
-%A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
-
-%\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint from population-level epidemic dynamics (rather than individual-level transmission data).
-
+A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
+\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
 To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
@@
 With our generation-interval estimates, we estimate that the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
-We also show that both growth-rate differences and the reproduction advantage decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction advantage remain constant \citep{park2021roles}.
+
+We considered two ways of measuring transmission advantanges: growth-rate differences and reproduction advantage.
+Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
+On the other hand, the growth-rate difference likely provides a better measure for transmission advantage for an ongoing epidemic as it provides information about which strain is growing faster.
+For example, when two strains have the same $\RR$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
+Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.
 
@@
 \textbf{Sensitivity of the estimates of the mean generation interval to the assumed values of the correlation coefficient of the lognormal distribution.}
-Lines represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
-For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 and 51.
+Lines and shaded regions represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
+For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 and 51 of 2021.
 }
@@
 \textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
-See Figure 4 in the main text for figure caption.
+Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-househol transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
+Lines represent maximum likelihood estimates.
+Shaded regions represent the corresponding 95\% confidence intervals.
 }
@@
 \textbf{Estimated time-varying reproduction number advantages of the Omicron variant using between-household generation-interval distributions.}
-See Figure 5 in the main text for figure caption.
+(A) Estimated instantaneous reproduction numbers and their ratios over time while accounting for differences in the generation-interval distributions.
+(B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
+The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
+The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the forward generation-interval distributions for between-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
+Purple lines represent the ratio between the effective reproduction numbers of the Delta and Omicron variants.
+Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
 }

### 2022-03-22 calculate_gauteng.R

@@ -1,0 +1,28 @@
+load("rdaout/fit_lognormal_base_comb_within.rda")
+r_diff <- 0.32
+r_delta <- -0.06
+r_omicron <- r_delta + r_diff
+
+gbar_delta_orig <- 6.4
+gkappa_delta_orig <- (5.4/6.4)^2
+
+## I have no idea what SD they assume..
+## maybe JD or SA might know...
+## TODO: ask and confirm
+gbar_omicron_orig <- 5.2
+gkappa_omicron_orig <- gkappa_delta_orig
+
+R_omicron_orig <- (1 + r_omicron * gbar_omicron_orig * gkappa_omicron_orig)^(1/gkappa_omicron_orig)
+R_delta_orig <- (1 + r_delta * gbar_delta_orig * gkappa_delta_orig)^(1/gkappa_delta_orig)
+
+R_omicron_new <- 1/integrate(function(x) {
+  dlnorm(x, coef(fit_lognormal_base_comb_sgtf_within)[[1]], coef(fit_lognormal_base_comb_sgtf_within)[[2]]) * exp(-r_omicron * x)
+}, 0, 100)[[1]]
+
+R_delta_new <- 1/integrate(function(x) {
+  dlnorm(x, coef(fit_lognormal_base_comb_nsgtf_within)[[1]], coef(fit_lognormal_base_comb_nsgtf_within)[[2]]) * exp(-r_delta * x)
+}, 0, 100)[[1]]
+
+R_omicron_orig/R_delta_orig
+
+R_omicron_new/R_delta_new

### 2022-03-22 figure_compare_main.R

@@ -1,7 +1,7 @@
 library(bbmle)
 library(dplyr)
 library(readxl)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(egg)
 library(gridExtra)
 source("serialfun.R")
@@ -119,21 +119,21 @@ g1 <- ggplot(serialdata_comb_nsgtf_within) +
   geom_point(data=serialdata_comb_sgtf_within, aes(serial+0.1, n/sum(n), col="Omicron", shape="Omicron"), size=3) +
   geom_line(data=serial_density_lognormal_nsgtf_within, aes(x, y, col="Delta", lty="Delta"), lwd=1) +
   geom_line(data=serial_density_lognormal_sgtf_within, aes(x, y, col="Omicron", lty="Omicron"), lwd=1) +
-  scale_x_continuous("Within-household serial intervals (days)", expand=c(0, 0), limits=c(-6, 16)) +
+  scale_x_continuous("Within-household\nserial intervals (days)", expand=c(0, 0), limits=c(-6, 16)) +
   scale_y_continuous("Density (1/day)", expand=c(0, 0), limits=c(0, 0.23)) +
   scale_color_manual("", values=c("black", "orange")) +
   scale_linetype_discrete("") +
   scale_shape_discrete("") +
   theme(
     panel.grid = element_blank(),
-    legend.position = c(0.8, 0.85),
+    legend.position = c(0.75, 0.85),
     legend.background = element_rect(fill=NA)
   )
 
 g2 <- ggplot(genden_nsgtf_within) +
   geom_line(aes(time, density), lwd=1) +
   geom_line(data=genden_sgtf_within, aes(time, density), lwd=1, lty=2, col="Orange") +
-  scale_x_continuous("Within-household generation intervals (days)", expand=c(0, 0), limits=c(0, 14)) +
+  scale_x_continuous("Within-household\ngeneration intervals (days)", expand=c(0, 0), limits=c(0, 14)) +
   scale_y_continuous("Density (1/day)", expand=c(0, 0), limits=c(0, 0.38)) +
   scale_color_manual("", values=c("black", "orange")) +
   scale_linetype_discrete("") +
@@ -165,21 +165,21 @@ g5 <- ggplot(serialdata_comb_nsgtf_between) +
[…]

### 2022-03-22 figure_compare_r_0.R

@@ -1,126 +1,0 @@
-library(bbmle)
-library(dplyr)
-library(readxl)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
-library(egg)
-library(gridExtra)
-source("serialfun.R")
-load("rdaout/fit_lognormal_base_r_0.rda")
-
-mean_nsgtf <- exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[1]]+coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]^2/2)
-
-tmp_nsgtf <- c(
-  exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]^2/2),
-  exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]^2/2) * coef(fit_lognormal_base_r_0_50_nsgtf_within)[[2]]
-)
-
-sd_mean_nsgtf <- sqrt(c(tmp_nsgtf %*% vcov(fit_lognormal_base_r_0_50_nsgtf_within) %*% tmp_nsgtf))
-
-mean_sgtf <- exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[1]]+coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]^2/2)
-
-tmp_sgtf <- c(
-  exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]^2/2),
-  exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[1]]) * exp(coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]^2/2) * coef(fit_lognormal_base_r_0_50_sgtf_within)[[2]]
-)
-
-sd_mean_sgtf <- sqrt(c(tmp_sgtf %*% vcov(fit_lognormal_base_r_0_50_sgtf_within) %*% tmp_sgtf))
-
-c(mean_nsgtf - 1.96 * sd_mean_nsgtf, mean_nsgtf, mean_nsgtf + 1.96 * sd_mean_nsgtf)
[…]

### 2022-03-22 figure_compare_rho.R

@@ -7,16 +7,14 @@ library(gridExtra)
 load("rdaout/fit_lognormal_base_comb_rho.rda")
 
 g1 <- ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
+  geom_ribbon(aes(rho, ymin=lwr, ymax=upr), alpha=0.2) +
   geom_line(aes(rho, est), lwd=1, col="black") +
-  geom_line(aes(rho, lwr), lwd=1, col="black") +
-  geom_line(aes(rho, upr), lwd=1, col="black") +
+  geom_ribbon(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, ymin=lwr, ymax=upr), fill="orange", alpha=0.2) +
   geom_line(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, est), lty=2, col="orange", lwd=1) +
-  geom_line(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, lwr), lty=2, col="orange", lwd=1) +
-  geom_line(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, upr), lty=2, col="orange", lwd=1) +
   scale_x_continuous("Lognormal correlation coefficient") +
   scale_y_continuous("Mean generation interval (days)") +
   theme(
     panel.grid = element_blank()
   )
 
-ggsave("figure_compare_rho.pdf", g1, width=12, heigh=8)
+ggsave("figure_compare_rho.pdf", g1, width=6, heigh=4)

### 2022-03-22 figure_epidemic.R

@@ -3,7 +3,7 @@ library(readxl)
 library(dplyr)
 library(lubridate)
 library(zoo)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=13))
 library(mgcv)
 library(mvtnorm)
 library(egg)
@@ -80,7 +80,9 @@ g1 <- ggplot(cases_all) +
   scale_y_log10("Number of reported cases", limits=c(9000, NA),
                 breaks=c(1e4, 2e4, 4e4, 8e4)) +
   theme(
-    panel.grid = element_blank()
+    panel.grid = element_blank(),
+    axis.text.x = element_text(angle=45, hjust=1),
+    axis.title.x = element_blank()
   )
 
 g2 <- ggplot(variant3) +
@@ -96,7 +98,9 @@ g2 <- ggplot(variant3) +
   scale_color_manual("", values=c("black", "orange")) +
   theme(
     panel.grid = element_blank(),
-    legend.position = "right"
+    legend.position = "right",
+    axis.text.x = element_text(angle=45, hjust=1),
+    axis.title.x = element_blank()
   )
 
 gfit1 <- gam(log(delta_cases)~s(time, bs="cs"), data=variant4, method="REML")
@@ -164,7 +168,9 @@ g3 <- ggplot(variant4) +
   scale_y_log10("Number of estimated cases") +
   theme(
     panel.grid = element_blank(),
-    legend.position = "none"
+    legend.position = "none",
+    axis.text.x = element_text(angle=45, hjust=1),
+    axis.title.x = element_blank()
   )
 
 g4 <- ggplot(all_growth) +
@@ -179,11 +185,13 @@ g4 <- ggplot(all_growth) +
   scale_fill_manual(values=c("purple", "black", "orange")) +
   theme(
     panel.grid = element_blank(),
-    legend.title = element_blank()
+    legend.title = element_blank(),
+    axis.text.x = element_text(angle=45, hjust=1),
+    axis.title.x = element_blank()
   )
 
 all_growth %>% filter(type=="Delta")
 
 gtot <- ggarrange(g1, g2, g3, g4, nrow=2, labels=c("A", "B", "C", "D"), draw=FALSE)
 
-ggsave("figure_epidemic.pdf", gtot, width=12, height=8)
+ggsave("figure_epidemic.pdf", gtot, width=8, height=6)

### 2022-03-22 figure_incubation.R

@@ -1,5 +1,5 @@
 library(dplyr)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(egg)
 source("sample_incubation.R")
 

### 2022-03-22 figure_reproduction_advantage.R

@@ -5,7 +5,7 @@ library(tidyr)
 library(dplyr)
 library(lubridate)
 library(zoo)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(mgcv)
 library(mvtnorm)
 library(egg)
@@ -144,7 +144,9 @@ g2 <- ggplot(filter(advantagesumm, key %in% c("R_advantage_naive", "R_delta", "R
   scale_fill_manual(values=c("purple", "black", "orange"), labels=c("Advantage", "Delta", "Omicron")) +
   theme(
     panel.grid = element_blank(),
-    legend.title = element_blank()
+    legend.title = element_blank(),
+    axis.text.x = element_text(angle=45, hjust=1),
+    axis.title.x = element_blank()
   )
 
 g1 <- ggplot(filter(advantagesumm, key %in% c("R_advantage", "R_delta", "R_omicron"))) +
@@ -161,7 +163,9 @@ g1 <- ggplot(filter(advantagesumm, key %in% c("R_advantage", "R_delta", "R_omicr
   theme(
     panel.grid = element_blank(),
     legend.title = element_blank(),
-    legend.position = "none"
+    legend.position = "none",
+    axis.text.x = element_text(angle=45, hjust=1),
+    axis.title.x = element_blank()
   )
 
 advantagesumm2 <- advantagesumm %>%
@@ -194,7 +198,7 @@ g5 <- ggplot(advantagesumm2) +
 
 gfinal <- ggarrange(g1, g2, nrow=1, labels=c("A", "B"), draw=FALSE)
 
-ggsave("figure_reproduction_advantage.pdf", gfinal, width=8, height=3)
+ggsave("figure_reproduction_advantage.pdf", gfinal, width=8.1, height=3)
 
 gsupp <- ggarrange(g5, g3, g4, nrow=1, labels=c("A", "B", "C"), draw=FALSE)
 

### 2022-03-22 fit_lognormal_base_r_0.R

@@ -1,74 +1,0 @@
-library(dplyr)
-library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-
-r_nsgtf <- 0
-r_sgtf <- 0
-rho <- 0.75
-
-moment_0_nsgtf <- integrate(function(z) {
-  dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
-}, 0, Inf)[[1]]
-
-moment_1_nsgtf <- integrate(function(z) {
-  z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
-}, -100, 100)[[1]]
-
-moment_2_nsgtf <- integrate(function(z) {
-  z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)
-}, -100, 100)[[1]]
-
-moment_0_sgtf <- integrate(function(z) {
-  dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * z)
-}, 0, Inf)[[1]]
-
-moment_1_sgtf <- integrate(function(z) {
-  z * dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * z)
-}, -100, 100)[[1]]
-
-moment_2_sgtf <- integrate(function(z) {
-  z^2 * dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r_sgtf * z)
-}, -100, 100)[[1]]
-
-mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
-var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
-
-logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
-logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
-
-mean_inc_sgtf <- moment_1_sgtf/moment_0_sgtf
-var_inc_sgtf <- moment_2_sgtf/moment_0_sgtf - mean_inc_sgtf^2
-
[…]

### 2022-03-22 omicron-generation.tex
spelling check

@@
 
-Estimating the generation-intervaldistribution is challenging, in part, due to difficulties in observing actual infection events.
+Estimating the generation-interval distribution is challenging, in part, due to difficulties in observing actual infection events.
 Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
@@
 While the derivation of the forward serial-interval distribution \eref{forwardserial} may be complex, its implementation is simple.
-The main difference between our model and prevous models that neglect dynamical effects \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation} is the exponential growth term $\exp(r \alpha_1)$ and the normalization term $\phi$---it is relatively straightforward to include these terms in existing models of serial intervals.
+The main difference between our model and previous models that neglect dynamical effects \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation} is the exponential growth term $\exp(r \alpha_1)$ and the normalization term $\phi$---it is relatively straightforward to include these terms in existing models of serial intervals.
 \cite{ferretti2020quantifying, ferretti2020timing} also included this term in their analyses of serial-interval data, but only accounted for the epidemic growth effect (and not the decay effect).
@@
 
-Similar pictures arise for between-household transmission pairs, but the differencse in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
+Similar pictures arise for between-household transmission pairs, but the differences in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
 Consistent with the original study, which also reported shorter mean serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean generation intervals for between-household Delta pairs.
-While the difference in mean generation intervals is larger, there is greater uncertainty in their mean estimates (\fref{serial}F): 2.9 days (95\% CI: 2.5--3.3 days) for the Omicron variant and 3.5 days (95\% CI: 3.2--3.8 dyas) for the Delta variant.
+While the difference in mean generation intervals is larger, there is greater uncertainty in their mean estimates (\fref{serial}F): 2.9 days (95\% CI: 2.5--3.3 days) for the Omicron variant and 3.5 days (95\% CI: 3.2--3.8 days) for the Delta variant.
 Once again, these patterns are robust across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants (\fref{serial}G--H).
@@
 
-A few studies have suggesed the the incubation period of the Omicron variant may be shorter than that of the Delta variant.
+A few studies have suggested the the incubation period of the Omicron variant may be shorter than that of the Delta variant.
 The median estimates of the Omicron incubation period typically range between 3--4 days, consistent with earlier findings of \citep{backer2021omicron}. 
@@
 A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
-\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
+\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
 To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
@@
 
-We considered two ways of measuring transmission advantanges: growth-rate differences and reproduction advantage.
+We considered two ways of measuring transmission advantages: growth-rate differences and reproduction advantage.
 Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
@@
 \textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
-Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-househol transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
+Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-household transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
 Lines represent maximum likelihood estimates.

### 2022-03-22 omicron-generation.tex
final update

@@
 Each of these transmission-interval distributions are subject to dynamical effects which, can bias estimation.
-For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently, shortening their incubation periods---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
+For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
 Therefore, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions, especially if their growth rates differ.
@@
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) and without (mostly Delta), but did not consider growth-rate differences in their inference:
+The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider growth-rate differences in their inference:
 during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
@@
 \end{align}
-where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$.
+where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$, and $b_I(x)$ represents the backward incubation-period distribution.
 Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time, \fref{diagram}B):

### 2022-03-22 omicron-generation.tex
adding a sentence about the delta method

@@
 Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
-We then calculate the mean forward generation interval using \eref{meangen} and approximate the 95\% confidence interval using the Delta method.
+We then calculate the mean forward generation interval using \eref{meangen}.
+The 95\% confidence intervals are calculated by taking the estimated variance-covariance matrix for the log-mean and -standard deviation parameters of the log normal distributions and calculating the corresponding variance-covariance for the overall mean using Taylor expansion---this method is also known as the Delta method \citep{oehlert1992note}.
 We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.

### 2022-03-25 omicron-generation.tex
editing descriptions

@@
 
-Estimating the generation-interval distribution is challenging, in part, due to difficulties in observing actual infection events.
+Estimating the generation-interval distribution is challenging, in part due to difficulties in observing actual infection events.
 Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
@@
 The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
-We expect this forward incubation-period distribution $f_I(\tau)$ to remain constant over the course of an epidemic and provide reliable estimates of the distribution across individuals. 
+We expect this forward incubation-period distribution $f_I(\tau)$ to remain relatively constant over the course of an epidemic, although biases can arise in \emph{observing} incubation periods, based on public or medical awareness of the disease.
 Backward incubation periods are measured from a cohort of individuals who developed symptoms at the same time. 
@@
 
-For a given growth rate $r$, the corresponding forward incubation-period distributions can be calculated by inverting \eref{backward}:
+For a given growth rate $r$, the corresponding forward incubation-period distributions can be calculated by inverting \eref{backward}, taking into account that $f_I$ is a distribution:
 \begin{equation}
@@
 To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified lognormal distribution falls between the lower and upper credible limits matches 95\% \citep{park2020reconciling}.
+First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified lognormal distribution falls between the lower and upper credible limits is 95\% \citep{park2020reconciling}.
 We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.
@@
 
-\subsection{Estimating instantaneous reproduction number}
+\subsection{Estimating instantaneous reproduction numbers}
 
@@
 We note that the growth-rate difference between the Delta and Omicron variants decreased over time.
-Hereafter, we choose $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as our nominal growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
+Hereafter, we use $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as representative growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
 We then evaluate the growth-rate effects across $r=-0.1/\mathrm{day}$--$0.0/\mathrm{day}$ for the Delta variant and $r=0.1/\mathrm{day}$--$0.2/\mathrm{day}$ for the Omicron variant as a sensitivity analysis.
@@
 Thus, dynamical bias would be expected to lead to shorter observed (backward) incubation periods in Omicron, and longer observed incubation periods in Delta. 
-When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions (\fref{incubation}B)---for illustrative purposes, we assume $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively.
-Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates with unclear differences between two variants (\fref{incubation}C--D).
+When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions (\fref{incubation}B).
+Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates (\fref{incubation}C--D).
 % \jd{Do we need the ff.~s.? What does it add to the Fig?}
 % \swp{I think people like to see actual values...}
+% \jd{I trimmed some other stuff around here; take a look at a diff if you can 2022 Mar 25 (Fri).}
 For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$ (\fref{incubation}C), while
@@
 The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
-Accounting for differences in epidemic growth rates, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
+Accounting for differences in epidemic growth rates, however, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
 Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January. 
-Our updated generation-interval estimates can provide more accurate understanding of epidemic dynamics and control measures.
+Improving generation-interval estimates by taking dynamical effects into account may improve understanding of epidemic dynamics and control measures.
 
@@
 A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation-period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
-Here, we build on previous work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
-Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
-These effects are also pertinent to epidemiological inferences of past events from a cohort of infected individuals who experienced the succeeding event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
+Here, we build on newer work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
+%% Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
+\jd{Not seeing what this s.\ is saying.
+It's important any time we want to calculate or compare intervals.
+What's the time when it's less important?
+When we're comparing the same variant at the same time and place to itself?}
+These effects are also pertinent to epidemiological inferences of past events from a cohort of infected individuals who experienced a later event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
 Our sensitivity analysis also shows that the assumptions about the correlation between incubation periods and generation intervals can also have important effects on the estimates of the generation-interval distributions (Supplementary Figure S1).
@@
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-In addition, the unmitigated generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating ``unmitigated'' or ``intrinsic'' generation-interval distributions of SARS-CoV-2 variants are expected to be a difficult problem as it requires early data during which interventions and awareness levels were minimal \citep{sender2021unmitigated}.
-Nonetheless, our estimates of the realized generation-interval distributions better describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects.
-Therefore, our generation-interval estimates are expected to give more accurate estimates of reproduction numbers.
+In addition, the ``intrinsic'' generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic (or ``unmitigated'') generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
+Nonetheless, our estimates of the realized generation-interval distributions better describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
+and can therefore be expected to improve estimates of effective reproduction numbers.
 
@@
 Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
-On the other hand, the growth-rate difference likely provides a better measure for transmission advantage for an ongoing epidemic as it provides information about which strain is growing faster.
+On the other hand, the growth-rate difference is easier to estimate in real time is also more directly relevant to short-term dynamics. 
 For example, when two strains have the same $\RR$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.

### 2022-04-10 omicron-generation.tex
edit
lots of edits

@@
 	\textbf\newline{
-		Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants
+		Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2
 	}
@@
 Estimating the differences in the incubation-period, serial-interval, and generation-interval distributions of SARS-CoV-2 variants is critical to understanding their transmission dynamics and outbreak control.
-However, estimating and comparing the timing of infection and transmission is inherently difficult due to the impact of epidemic dynamics---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have been infected recently.
+However, the impact of epidemic dynamics is often neglected in estimating the timing of infection and transmission---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have been infected recently.
 Here, we re-analyze incubation-period and serial-interval data describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
 Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
-When we account for the impact of growth-rate differences on inferences of two variants, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
-The differences in generation intervals may reflect higher effective reproduction numbers of the Omicron variant, which can drive faster susceptible depletion among close contacts.
+When we account for growth-rate differences of two variants during the study period, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
+The differences in realized generation intervals may be driven by the network effect---higher effective reproduction numbers of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (and therefore shortening realized generation intervals).
 Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
@@
 
-Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the COVID-19 pandemic.
+Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the COVID-19 pandemic \citep{park2021roles}.
 Transmission advantages of invading variants are typically characterized by the ratios of reproduction numbers, $\Rinv/\Rres$, and the differences in growth rates, $\rinv-\rres$.
@@
 Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
-Each of these transmission-interval distributions are subject to dynamical effects which, can bias estimation.
-For example, when the epidemic is growing, a cohort of infectors who developed symptoms at the same time are more likely to have been infected recently---in other words, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
+Each of these transmission-interval distributions are subject to dynamical effects, which can cause transmission-interval distributions to systematically differ from the corresponding generation-interval distribution.
+For example, when the epidemic is growing, there will be more recent infections. 
+Therefore, we are more likely to observe recently infected individuals among a cohort of infectors who developed symptoms at the same---we refer to this effect as the ``dynamical bias''.
+In this case, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
 Therefore, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions, especially if their growth rates differ.
@@
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider growth-rate differences in their inference:
+The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
 during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
@@
 
-Serial interval data are taken from \cite{backer2021omicron} (see original article for details of data collection).
-The data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates or symptom onset dates.
-The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (weeks 50 and 51 of 2021), and the type of transmission (within or between households).
-In the main text, we combine data from weeks 50 and 51 of 2021 and present a stratified analysis in Supplementary Material.
+Serial interval data are taken from \cite{backer2021omicron}.
+Infector-infectee pairs were identified through contact tracing, and their symptom onset dates were reported through a national surveillance database. 
+Serial intervals were then calculated by taking the difference between symptom onset dates of the infector and the infectee.
+In order to ensure independence between serial intervals, one infectee was chosen at random for each infector in the original analysis.
+See original article for additional details of data collection.
+
+Publicly available data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates or symptom onset dates.
+The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (week 50, 13--19 December 2021, and week 51, 20--26 December 2021), and the type of transmission (within or between households).
+In the main text, we combine data from weeks 50 and 51 of 2021 (13--26 December) and present a stratified analysis in Supplementary Material.
 For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
@@
 To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
+We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data.
+For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
+instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
+We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making the uncertainty negligible.
+
 We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
+More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} package).
+We also assume normally distributed errors around the logged weekly cases.
+In principle, using negative-binomial likelihoods with a log link function is more appropriate for modeling count data, especially when data are overdispersed \citep{ver2007quasi};
+here, we use Gaussian likelihoods on logged cases because the inferred numbers of cases caused by each variant are not counts.
 Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.
@@
 We calculate time-varying growth rates from each parameter set and use equi-tailed quantiles to generate 95\% confidence limits. 
+% \swp{Sam is calling this ``simultaneous confidence intervals'' but not sure what he means. Need to ask again.}
 
@@
 The incubation-period distributions from 513 individuals (258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022, were previously reported in \cite{backer2021omicron}.
-\cite{backer2021omicron} used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates, using a uniform distribution, and compares this to a known symptom-onset time.
+\cite{backer2021omicron} used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates.
+In particular, the methods of \cite{backer2020incubation} assume that the infection time is uniformly distributed across exposure dates and compares the inferred infection time to a known symptom-onset time to calculate the incubation period for each individual.
+While this method may accurately estimate the infection time, and therefore the incubation period, of each individual, dynamical biases can still affect the distribution of the inferred incubation periods from this cohort.
 
-In practice, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward \citep{park2021forward}.
+More specifically, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward \citep{park2021forward}.
 The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
-We expect this forward incubation-period distribution $f_I(\tau)$ to remain relatively constant over the course of an epidemic, although biases can arise in \emph{observing} incubation periods, based on public or medical awareness of the disease.
+We expect this forward incubation-period distribution $f_I(\tau)$ to remain relatively constant over the course of an epidemic of one given variant, although biases can arise in \emph{observing} incubation periods, based on public or medical awareness of the disease.
 Backward incubation periods are measured from a cohort of individuals who developed symptoms at the same time. 
 The backward incubation-period distribution is sensitive to epidemic dynamics:
-the difference arises because forward incubation periods look forward from the reference point towards symptom development, which is an individual-level process, while backward incubation periods look backwards towards an infection event, which requires an interaction with an infectious individual.
+the difference between the forward and backward distribution arises because forward incubation periods look forward from the reference point towards symptom development, which is an individual-level process, while backward incubation periods look backwards towards an infection event, which requires an interaction with an infectious individual.
 
 In particular, when incidence of infection is growing exponentially, we are more likely to observe backward incubation periods that are shorter than the corresponding forward incubation periods because there will be relatively more individuals who were infected recently.
-Assuming that incidence is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
+Assuming that incidence of infection is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
 \begin{equation}
@@
 \end{equation}
-The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution without taking growth rates into account.
+Therefore, the backward incubation-period distribution $b_I(\tau)$ gives a biased estimate of the corresponding forward distribution $f_I(\tau)$.
+The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution.
 
-For a given growth rate $r$, the corresponding forward incubation-period distributions can be calculated by inverting \eref{backward}, taking into account that $f_I$ is a distribution:
+Assuming a constant growth rate $r$, the corresponding forward incubation-period distributions can be calculated by inverting \eref{backward}, taking into account that $f_I$ is a probability distribution and therefore needs to be normalised to integrate to 1:
 \begin{equation}
@@
 instead we take the backward incubation-period distributions $b_I(x)$ estimated by \cite{backer2021omicron}, which was originally assumed to follow a Weibull distribution, and apply \eref{forward}.
-In particular, \cite{backer2021omicron} estimated the scale and shape parameters of the Weibull distribution to be 4.93 (95\% CI: 4.51--5.37) and 1.83 (95\% CI: 1.59--2.08), respectively, for the non-SGTF cases, and 3.60 (95\% CI: 3.23--3.98) and 1.50 ((95\% CI: 1.32--1.70), respectively, for SGTF cases.
+In particular, \cite{backer2021omicron} estimated the scale and shape parameters of the Weibull distribution to be 4.93 (95\% CI: 4.51--5.37) and 1.83 (95\% CI: 1.59--2.08), respectively, for the Delta cases, and 3.60 (95\% CI: 3.23--3.98) and 1.50 ((95\% CI: 1.32--1.70), respectively, for Omicron cases.
 
@@
 In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
-(B) Generation intervals are better measured from the cohort of infectors who are infected at the same time because their incubation periods will be unbiased.
+(B) Generation intervals for the cohort of infectors who are infected at the same time will have unbiased incubation periods.
 \label{fig:diagram}
@@
 
-The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
-This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
+The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased (compared to the generation-interval distribution for an infection-based cohort) because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
+This generation-interval distribution for a symptom-based cohort depends on the backward incubation-period distribution:
 \begin{align}
@@
 \end{align}
+Previous analyses of serial-interval distributions typically assumed that the incubation periods and generation intervals are independent \citep{ganyani2020estimating}; in this case, the generation-interval distribution for the symptom-based and infection-based cohorts are identical.
 
@@
 Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
+Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed them to conclude that the mean generation and serial intervals are identical.
 For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time, \fref{diagram}B), and drop the subscript $\mathrm{inf}$.
 
-We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
-The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient on a log scale $\rho$.
-Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
-Assuming that the incidence of infection is changing exponentially at rate $r$, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be calculated by integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ \citep{park2021forward}:
+Assuming that the incidence of infection is changing exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged across $t$ \citep{park2021forward}.
+Then, we can focus on the forward serial-interval distribution at $t=0$, which in turn allows us to reparameterize the incubation-period and generation-interval distributions in terms of the infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$.
+Under this parameterization, for a given length of a serial interval $\tau$, we can rewrite the incubation period of the infector as $-\alpha_1$; the generation interval as $\alpha_2 - \alpha_1$; and the incubation period of the infectee as $\tau - \alpha_2$.
+Then, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be expressed in terms of three distributions (\eref{serial}): the backward incubation-period distribution of the infector $b_I(-\alpha_1)$, the forward generation-interval distribution conditional on a known value of the incubation period $x$, $f_{G|I}(\alpha_2-\alpha_1|-\alpha_1)$, and the forward incubation-period distribution of the infectee $f_I(\tau - \alpha_2)$.
+Integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ and rewriting the backward incubation-period distribution $b_I(-\alpha_1)$ in terms of the forward distribution, we obtain \citep{park2021forward}:
 \begin{equation}
@@
 \end{equation}
-where $\phi$ is a normalization constant chosen so that $\int f_S(x) \dx = 1$;
-$-\alpha_1$ corresponds to the incubation period of the infector;
-$\alpha_2 - \alpha_1$ corresponds to the generation interval;
-and $\tau - \alpha_2$ corresponds to the incubation period of the infectee.
-As discussed earlier in Section 2.3, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
-As we show later in the results section, the exponential growth rate changes over the study period, including weeks 50 and 51;
+where $\phi$ is a normalization constant chosen so that $\int f_S(x) \dx = 1$.
+As discussed earlier, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
+As we show later in the results section, the exponential growth rate changes over the study period, including weeks 50 and 51 (13--26 December 2021);
 for illustrative purpose, we choose values of $r$ that represent the dynamics of Delta and Omicron infections during this period and repeat the analysis across plausible ranges of $r$ (discussed later in detail).
@@
 
+We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
+The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient on a log scale $\rho$.
+Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
 For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
@@
 where $i(t)$ represents incidence of infection.
-Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for weeks 50 and 51 of 2021---when the epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
+Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for weeks 50 and 51 of 2021 (13--26 December)---when the epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
 Incidence of infection is approximated by shifting the smoothed case trajectories by one week to account for reporting delays.
@@
 The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
+These changes in growth rates coincide with the introduction of lockdown on 19 December 2021 \citep{netherlandslockdown} and its relaxation beginning 15 January 2022 \citep{netherlandsrelax,netherlandsopen}.
 We note that the growth-rate difference between the Delta and Omicron variants decreased over time.
@@
 Thus, dynamical bias would be expected to lead to shorter observed (backward) incubation periods in Omicron, and longer observed incubation periods in Delta. 
-When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions (\fref{incubation}B).
+When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions with a mean of 4.1 days (\fref{incubation}B);
+in this case, the difference between the mean backward and forward incubation periods correspond to $-22\%$ and $7\%$ bias for the Omicron and Delta variants, respectively.
 Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates (\fref{incubation}C--D).
-% \jd{Do we need the ff.~s.? What does it add to the Fig?}
-% \swp{I think people like to see actual values...}
-% \jd{I trimmed some other stuff around here; take a look at a diff if you can 2022 Mar 25 (Fri).}
-For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$ (\fref{incubation}C), while
-the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$ (\fref{incubation}D).
+For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$ (\fref{incubation}C),
+while the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$ (\fref{incubation}D).
 
@@
 Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
-For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13-–19 December, 2021).
+For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13--26 December, 2021).
 For within-household transmission pairs (\fref{serial}A), the Omicron variant has shorter mean serial interval (3.1 days; 95\% CI: 2.9--3.3 days) than that of the Delta variant (3.7 days; 95\% CI: 3.5--3.8 days).
@@
 Across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants, we estimate robust differences in their mean generation intervals (\fref{serial}C--D).
-Assuming lower values of the correlation between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
+Assuming lower values of the correlation $\rho$ between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
+In particular, the generation-interval estimates of the Omicron variant are more sensitive to the assumed values of $\rho$ due to faster changes in incidence of infection---for example, changing $\rho$ from 0.85 to 0.5 changes the mean generation-interval estimates for the Omicron variant from 3.1 days (95\% CI: 2.8--3.3 days) to 2.7 days (95\% CI: 2.5--2.9 days).
 
@@
 
-In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (weeks 50 and 51).
-While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50.
-We also estimate a reduction in the mean forward generation intervals from week 50 to week 51 (especially for the Delta variant).
+In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (13--19 December 2021 and 20--26 December 2021).
+While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50 (13--19 December 2021).
+We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant.
 
@@
 
-Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.73 (95\% CI: 1.59--1.89) to 1.14 (95\% CI: 1.00--1.32) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
-On the other hand, the reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
-We estimate the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants) stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
+Accounting for differences in the generation-interval distributions, we estimate that the instantaneous reproduction number of the Omicron variant decreased from 1.73 (95\% CI: 1.59--1.89) to 1.14 (95\% CI: 1.00--1.32) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
+On the other hand, the instantaneous reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
+We estimate the reproduction advantage (i.e., the ratio between the instantaneous reproduction numbers of the Omicron and Delta variants) stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
 However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore the reproduction advantage (\fref{reproduction}B).
@@
 \caption{
-\textbf{Estimated time-varying reproduction number advantages of the Omicron variant.}
+\textbf{Estimated instantaneous reproduction number advantages of the Omicron variant.}
 (A) Estimated instantaneous reproduction numbers and their ratios over time while accounting for differences in the generation-interval distributions.
@@
 
-We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands in late 2020 and early 2021.
+We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands in late 2021 and early 2022.
 The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
@@
 A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation-period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
-Here, we build on newer work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depends on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
-%% Accounting for the growth-rate effects is especially important when comparing serial and generation intervals of different variants or from different time periods.
-\jd{Not seeing what this s.\ is saying.
-It's important any time we want to calculate or compare intervals.
-What's the time when it's less important?
-When we're comparing the same variant at the same time and place to itself?}
+We build on newer work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depend on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
 These effects are also pertinent to epidemiological inferences of past events from a cohort of infected individuals who experienced a later event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
 Our sensitivity analysis also shows that the assumptions about the correlation between incubation periods and generation intervals can also have important effects on the estimates of the generation-interval distributions (Supplementary Figure S1).
+
+This study presents a simple method for accounting for dynamical biases in inferring incubation-period distributions based on epidemic growth rates.
+In practice, it is easier to measure the backward incubation-period distribution in practice because we can directly observe symptom onset. 
+Therefore, the observed incubation-period distributions are generally expected to be biased, and similar kinds of corrections will be necessary to accurately estimate the incubation-period distribution.
+We note that making these kinds of corrections will also depend on data availability, model complexity, and other epidemiological covariates affecting incubation periods, such as vaccine statuses. 
+Accounting for different sources of biases is critical to accurately estimating incubation-period distributions (and other epidemiological distributions alike) but will necessarily increase uncertainties in the estimates.
+On the other hand, it is still possible to characterize the forward incubation-period distributions without making growth-rate-based corrections through a careful cohorting of individuals with similar infection times when detailed information about infection time is available---we were not able to explore this in our analysis because we relied on publicly available information, which do not contain individual-level information, such as exposure or symptom onset dates.
 
@@
 \cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
-To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
 Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
@@
 While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
-The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
+The network effect may also explain a decrease in the mean generation interval between week 50 and 51 (13--26 December 2021), especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
 Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
@@
 
+We assume a constant growth rate for each variant throughout our analysis.
+During the study period, growth rates of both the Delta and Omicron variants changed slowly, and therefore our constant-growth-rate assumption provides a reasonable approximation for their dynamics across two weeks.
+However, such assumption might not be applicable when growth rates are changing rapidly (e.g., due to an introduction of stingent control measures) or if the sampling window is too wide.
+Extending our framework to account for time-varying growth rates is relatively simple when inferring the forward incubation-period distribution from the corresponding backward distribution---we can simply replace $r$ with $r(t)$ in \eref{forward} because the backward incubation-period distribution is a weighted average of the forward incubation-period distributions and the number of individuals in each cohort (i.e., individuals who were infected at the same time).
+However, such extensions are more complicated for linking generation- and serial-interval distributions because the forward serial-interval distribution also depends on the cohort reproduction number---for example, if a certain cohort of infectors had higher reproduction number (e.g., because they were infected before control measures were observed), we are more likely to observe transmission that started from this cohort (see \citep{park2021forward} for mathematical details).
+Assuming a constant growth rate allows us to simplify this complexity because the reproduction number also stays constant during the exponential phase.
+Future studies should
+
 Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
@@
 Lines and shaded regions represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
-For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 and 51 of 2021.
+For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 (13--19 December) and 51 (20--26 December) of 2021.
 }
@@
 \textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
-Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-household transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
+Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-household transmission) and the week of infectors' symptom onset (week 50, 13--19 December 2021, vs week 51, 20--26 December 2021,).
 Lines represent maximum likelihood estimates.

### 2022-04-10 figure_epidemic.R

@@ -71,8 +71,12 @@ variant4 <- variant3 %>%
   )
 
 g1 <- ggplot(cases_all) +
+  geom_rect(xmin=as.Date("2021-12-13"), xmax=as.Date("2021-12-19"), ymin=-Inf, ymax=Inf, alpha=0.1, fill="gray") +
+  geom_rect(xmin=as.Date("2021-12-20"), xmax=as.Date("2021-12-26"), ymin=-Inf, ymax=Inf, alpha=0.02, fill=4) +
   geom_point(aes(Date_of_publication, cases)) +
   geom_line(aes(Date_of_publication, rollmean), lty=1) +
+  annotate("text", x=as.Date("2021-12-15")+1, y=75000, label=c("Week\n50"), family="Times", size=3) +
+  annotate("text", x=as.Date("2021-12-22")+1, y=75000, label=c("Week\n51"), family="Times", size=3) +
   scale_x_date("Date", limits=c(as.Date("2021-11-28"), as.Date("2022-01-30")), 
                breaks=variant4$date,
                labels=c("Nov 28", "Dec 5", "Dec 12", "Dec 19", "Dec 26", "Jan 2", "Jan 9", "Jan 16",
@@ -175,12 +179,16 @@ g3 <- ggplot(variant4) +
 
 g4 <- ggplot(all_growth) +
   geom_hline(yintercept=0, lty=2) +
+  geom_vline(xintercept=4, lty=2) +
+  geom_vline(xintercept=8-1/7, lty=2) +
+  annotate("text", x=4, y=0.29, hjust=-0.05, label="Lockdown\nintroduced", family="Times") +
+  annotate("text", x=8-1/7, y=0.29, hjust=-0.05, label="Lockdown\nrelaxed", family="Times") +
   geom_ribbon(aes(time, ymin=lwr, ymax=upr, fill=type), alpha=0.2) +
   geom_line(aes(time, median, col=type), lwd=0.8) +
   scale_x_continuous("Date", breaks=1:10,
                      labels=c("Nov 28", "Dec 5", "Dec 12", "Dec 19", "Dec 26", "Jan 2", "Jan 9", "Jan 16",
                               "Jan 23", "Jan 30")) +
-  scale_y_continuous("Growth rate (1/day)") +
+  scale_y_continuous("Growth rate (1/day)", limits=c(NA, 0.3)) +
   scale_color_manual(values=c("purple", "black", "orange")) +
[…]

### 2022-04-10 figure_incubation.R

@@ -79,6 +79,14 @@ corrected_sgtf <- tibble(
   density=density0/sum(density0*0.1)
 )
 
+sum(observed_sgtf$time * observed_sgtf$density)/sum(observed_sgtf$density)
+
+sum(observed_nsgtf$time * observed_nsgtf$density)/sum(observed_nsgtf$density)
+
+sum(corrected_sgtf$time * corrected_sgtf$density)/sum(corrected_sgtf$density)
+
+sum(corrected_nsgtf$time * corrected_nsgtf$density)/sum(corrected_nsgtf$density)
+
 g1 <- ggplot(observed_nsgtf) +
   geom_line(aes(time, density, col="Delta", lty="Delta"), lwd=1) +
   geom_line(data=observed_sgtf, aes(time, density, col="Omicron", lty="Omicron"), lwd=1) +

### 2022-04-18 doc/omicron-generation-report.tex
move everything

\documentclass[12pt]{article}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

\usepackage{graphicx}
\usepackage{xspace}
%\usepackage{adjustbox}

\newcommand{\comment}{\showcomment}
%% \newcommand{\comment}{\nocomment}

\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\bmb}[1]{\comment{blue}{BMB}{#1}}
\newcommand{\djde}[1]{\comment{red}{DJDE}{#1}}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc} % make sure fancy dashes etc. don't get dropped

\usepackage{lineno}
\linenumbers

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{unsrt}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\mathrm{c}}}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\Rintrinsic}{\ensuremath{{\mathcal R}_{\textrm{\tiny intrinsic}}}\xspace}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\usepackage{lettrine}

\newcommand{\dropcapfont}{\fontfamily{lmss}\bfseries\fontsize{26pt}{28pt}\selectfont}
\newcommand{\dropcap}[1]{\lettrine[lines=2,lraise=0.05,findent=0.1em, nindent=0em]{{\dropcapfont{#1}}}{}}

\begin{document}

\begin{flushleft}{
	\Large
	\textbf\newline{
		Epidemic dynamics and network structures affect differences in incubation-period and generation-interval distributions of SARS-CoV-2 variants
	}
}
\bigskip

\section*{Summary}

\begin{itemize}
  \item Neglecting growth rate differences can exaggerate or diminish the observed differences in epidemiological distributions (e.g., incubation period distribution and generation-interval distribution) of SARS-CoV-2 variants
  \item Accounting for growth rate differences yield similar incubation period distribution estimates for Delta and Omicron variants
  \item Despite similaries in the incubation period distributions, we generally estimate shorter mean generation intervals for the Omicron variant (2.7--3.9 days) than for the Delta variant (3.9--4.5 days).
  \item Higher reproduction number of the Omicron variant can lead to a faster depletion of susceptible contacts, esepcially within a household, resulting in shorter realized generation intervals. Simulations suggest that the estimated differences in the mean generation interval are consistent with differences driven by susceptible depletion effects in small networks.
\end{itemize}

\end{flushleft}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_epidemic.pdf}
\caption{
\textbf{Epidemic dynamics on the Delta and Omicron variants in Netherlands.}
(A) Daily numbers of reported COVID-19 cases in Netherlands (points).
The solid line represents the 7-day moving average.
Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
(B) Proportion of SARS-CoV-2 variants detected from Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
(C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth rate differences (purple).
Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
Growth rates are estimated by taking the derivative of the generalized additive model estimates.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_incubation.pdf}
\caption{
\textbf{Observed and corrected differences in incubation period distributions of Delta and Omicron variants.}
(A) Maximum likelihood estimates of the observed (backward) incubation periods of the Delta (black) and Omicron (orange) variants by \cite{backer2021omicron}.
(B) Corrected estimates of the forward incubation periods for different assumptions about the growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare.pdf}
\caption{
\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
(A) Observed forward serial intervals from household transmission pairs in Netherlands for the Delta (black points) and Omicron (orange trianlges) variants \citep{backer2021omicron}.
Serial intervals are calculated from infectors who developed symptoms between 13 and 19 December 2021.
The black sold line and the orange dashed lines represent the maximum likelihood fits for forward serial-interval distributions of Delta ($r=-0.05/\textrm{day}$) and Omicron ($r=0.15/\textrm{day}$) variants, respectively. 
Bivariate lognormal distributions with correlation coefficient of 0.75 are used to model relationships between incubation and generation intervals.
(B) Sensitivity of the mean generation-interval estimates to assumed growth rates of the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
Middle lines represent median estimates, and lower and upper lines represent corresponding 95\% confidence intervals.
(C--D) Maximum likelihood estimaes for the bivariate relationship between incubation and generation intervals of the Omicron (C) and Delta (D) variants.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_reproduction_advantage.pdf}
\caption{
\textbf{Estimated time-varying reproduction number advantages of the Omicron variant.}
The effective reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (Figure 1C) and taking the maximum likelihood estimates of the generation-interval distributions based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
Purple lines illustrate the ratio between the effective reproduction numbers of the Delta and Omicron variants.
Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_household.pdf}
\caption{
\textbf{Effects of network structures on realized generation intervals.}
We simulate transmission from 2000 fully connected networks with varying sizes (4--20), reproduction number ratios (1.5--2.5), and over-dispersion parameters ($k=0.1$--$10$) and compare mean generation intervals.
Egocentric simulations (solid lines) only consider onward transmission from a single, primary infector in each network.
Local simulations (dashed lines) allow infection to continue to spread within a network until infections can no longer spread (due to herd immunity, stochastic fadeouts, and/or a lack of susceptible individuals).
Lines represents mean across 2000 networks, and shaded regions represent corresponding 95\% confidence intervals.
The Delta variant (black) is assumed to have effective reproduction number of 0.8.
The Omicron variant (orange) is assumed to have effective reproduction number between 1.2 and 2.
Dotted horizontal lines represent the maximum likelihood estimates of the generation-interval distributions based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant.
}
\end{figure}

\pagebreak

\section*{Supplementary Materials}

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_he.pdf}
\caption{
\textbf{Estimated relationship between forward incubation and generation-interval distributions of Delta and Omicron variants.}
We model relationships between incubation and generation intervals using a shifted gamma distribution as suggested by \cite{he2020temporal} as a sensitivity analysis.
We obtain similar mean generation-interval estimates, but the predited bivariate relationships (C--D) are considerably different.
This model also gives overall worse fits to the observed serial-interval distribution.
See Figure 3 in the main text for details.
}
\end{figure}

\bibliography{omicron.bib}

\end{document}

### 2022-04-18 analysis/calculate_gauteng.R

@@ -1,4 +1,4 @@
-load("rdaout/fit_lognormal_base_comb_within.rda")
+load("../rdaout/fit_lognormal_base_comb_within.rda")
 r_diff <- 0.32
 r_delta <- -0.06
 r_omicron <- r_delta + r_diff

### 2022-04-18 analysis/calculate_incubation_mle.R

@@ -1,4 +1,4 @@
-source("sample_incubation.R")
+source("../R/sample_incubation.R")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
@@ -39,4 +39,4 @@ var_inc_sgtf <- moment_2_sgtf/moment_0_sgtf - mean_inc_sgtf^2
 logsd_inc_sgtf <- sqrt(log(var_inc_sgtf/mean_inc_sgtf^2 + 1))
 logmean_inc_sgtf <- log(mean_inc_sgtf/exp(logsd_inc_sgtf^2/2))
 
-save("logsd_inc_nsgtf", "logmean_inc_nsgtf", "logsd_inc_sgtf", "logmean_inc_sgtf", file="calculate_incubation_mle.rda")
+save("logsd_inc_nsgtf", "logmean_inc_nsgtf", "logsd_inc_sgtf", "logmean_inc_sgtf", file="../rdaout/calculate_incubation_mle.rda")

### 2022-04-18 analysis/fit_lognormal_base.R

@@ -1,9 +1,9 @@
 library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("rdaout/calculate_incubation_mle.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+load("../rdaout/calculate_incubation_mle.rda")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
@@ -11,24 +11,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_comb_sgtf_within <- serialdata %>%
-  filter(strain=="SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
+serialdata_50_sgtf_within <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="within")
 
-serialdata_comb_nsgtf_within <- serialdata %>%
-  filter(strain=="non-SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
+serialdata_50_nsgtf_within <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="within")
 
-data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
-data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
+data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
+data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
 
-fit_lognormal_base_comb_sgtf_within <- fitfun_lognormal(data=data_comb_sgtf_within, 
+fit_lognormal_base_50_sgtf_within <- fitfun_lognormal(data=data_50_sgtf_within, 
                                                       logmean_gen=logmean_inc_sgtf, 
                                                       logsd_gen=logsd_inc_sgtf, 
                                                       logmean_inc=logmean_inc_sgtf, 
@@ -36,7 +28,7 @@ fit_lognormal_base_comb_sgtf_within <- fitfun_lognormal(data=data_comb_sgtf_with
                                                       rho=rho, 
                                                       r=r_sgtf)
 
[…]

### 2022-04-18 analysis/fit_lognormal_base_50_between.R

@@ -1,8 +1,9 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("calculate_incubation_mle.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+load("../rdaout/calculate_incubation_mle.rda")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
@@ -35,4 +36,4 @@ fit_lognormal_base_50_nsgtf_between <- fitfun_lognormal(data=data_50_nsgtf_betwe
                                                        rho=rho, 
                                                        r=r_nsgtf)
 
-save("fit_lognormal_base_50_sgtf_between", "fit_lognormal_base_50_nsgtf_between", file="fit_lognormal_base_50_between.rda")
+save("fit_lognormal_base_50_sgtf_between", "fit_lognormal_base_50_nsgtf_between", file="../rdaout/fit_lognormal_base_50_between.rda")

### 2022-04-18 analysis/fit_lognormal_base_comb_between.R

@@ -1,9 +1,9 @@
 library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("rdaout/calculate_incubation_mle.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+load("../rdaout/calculate_incubation_mle.rda")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
@@ -44,4 +44,4 @@ fit_lognormal_base_comb_nsgtf_between <- fitfun_lognormal(data=data_comb_nsgtf_b
                                                        rho=rho, 
                                                        r=r_nsgtf)
 
-save("fit_lognormal_base_comb_sgtf_between", "fit_lognormal_base_comb_nsgtf_between", file="fit_lognormal_base_comb_between.rda")
+save("fit_lognormal_base_comb_sgtf_between", "fit_lognormal_base_comb_nsgtf_between", file="../rdaout/fit_lognormal_base_comb_between.rda")

### 2022-04-18 analysis/fit_lognormal_base_comb_rho.R

@@ -1,29 +1,39 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("rdaout/calculate_incubation_mle.rda")
-load("rdaout/fit_lognormal_base.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+load("../rdaout/calculate_incubation_mle.rda")
+load("../rdaout/fit_lognormal_base_comb_within.rda")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
-rhovec <- seq(0.5, 0.9, length.out=11)
+rhovec <- seq(0.5, 0.85, length.out=11)
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
 
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
 
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
+data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
 
-fit_lognormal_base_rho_50_nsgtf_within <- lapply(rhovec, function(rho) {
-  ff <- fitfun_lognormal(data=data_50_nsgtf_within, 
-                         logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
-                         logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
+fit_lognormal_base_rho_comb_nsgtf_within <- lapply(rhovec, function(rho) {
[…]

### 2022-04-18 analysis/fit_lognormal_base_comb_within.R

@@ -1,8 +1,9 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("calculate_incubation_mle.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+load("../rdaout/calculate_incubation_mle.rda")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
@@ -10,16 +11,24 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
 
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
 
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
+data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
 
-fit_lognormal_base_50_sgtf_within <- fitfun_lognormal(data=data_50_sgtf_within, 
+fit_lognormal_base_comb_sgtf_within <- fitfun_lognormal(data=data_comb_sgtf_within, 
                                                       logmean_gen=logmean_inc_sgtf, 
                                                       logsd_gen=logsd_inc_sgtf, 
                                                       logmean_inc=logmean_inc_sgtf, 
@@ -27,7 +36,7 @@ fit_lognormal_base_50_sgtf_within <- fitfun_lognormal(data=data_50_sgtf_within,
                                                       rho=rho, 
                                                       r=r_sgtf)
 
[…]

### 2022-04-18 analysis/fit_lognormal_base_rho.R

@@ -1,38 +1,30 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("rdaout/calculate_incubation_mle.rda")
-load("rdaout/fit_lognormal_base_comb_within.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+load("../rdaout/calculate_incubation_mle.rda")
+load("../rdaout/fit_lognormal_base.rda")
 
 r_nsgtf <- -0.05
 r_sgtf <- 0.15
-rhovec <- seq(0.5, 0.85, length.out=11)
+rhovec <- seq(0.5, 0.9, length.out=11)
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_comb_sgtf_within <- serialdata %>%
-  filter(strain=="SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
+serialdata_50_sgtf_within <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="within")
 
-serialdata_comb_nsgtf_within <- serialdata %>%
-  filter(strain=="non-SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
+serialdata_50_nsgtf_within <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="within")
 
-data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
-data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
+data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
+data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
 
-fit_lognormal_base_rho_comb_nsgtf_within <- lapply(rhovec, function(rho) {
-  print(rho)
-  ff <- fitfun_lognormal(data=data_comb_nsgtf_within, 
-                         logmean_gen=coef(fit_lognormal_base_comb_nsgtf_within)[[1]], 
-                         logsd_gen=coef(fit_lognormal_base_comb_nsgtf_within)[[2]], 
[…]

### 2022-04-18 analysis/fit_lognormal_comb_between_r.R

@@ -1,9 +1,10 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("fit_lognormal_base.rda")
+source("../R/Rserialfun.R")
+source("../R/Rfitfun.R")
+source("../R/Rsample_incubation.R")
+load("../rdaout/fit_lognormal_base_comb_between.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -12,16 +13,24 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_between <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="between")
+serialdata_comb_sgtf_between <- serialdata %>%
+  filter(strain=="SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
 
-serialdata_50_nsgtf_between <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="between")
+serialdata_comb_nsgtf_between <- serialdata %>%
+  filter(strain=="non-SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
 
-data_50_sgtf_between <- rep(serialdata_50_sgtf_between$serial, serialdata_50_sgtf_between$n)
-data_50_nsgtf_between <- rep(serialdata_50_nsgtf_between$serial, serialdata_50_nsgtf_between$n)
+data_comb_sgtf_between <- rep(serialdata_comb_sgtf_between$serial, serialdata_comb_sgtf_between$n)
+data_comb_nsgtf_between <- rep(serialdata_comb_nsgtf_between$serial, serialdata_comb_nsgtf_between$n)
 
-fit_lognormal_r_50_nsgtf_between <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_comb_nsgtf_between <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
[…]

### 2022-04-18 analysis/fit_lognormal_comb_within_r.R

@@ -1,10 +1,10 @@
 library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("rdaout/fit_lognormal_base_comb_within.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+source("../R/sample_incubation.R")
+load("../rdaout/fit_lognormal_base_comb_within.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -130,4 +130,4 @@ fit_lognormal_r_comb_sgtf_within <- lapply(r_sgtf, function(r) {
 }) %>%
   bind_rows
 
-save("fit_lognormal_r_comb_sgtf_within", "fit_lognormal_r_comb_nsgtf_within", file="fit_lognormal_comb_within_r.rda")
+save("fit_lognormal_r_comb_sgtf_within", "fit_lognormal_r_comb_nsgtf_within", file="../rdaout/fit_lognormal_comb_within_r.rda")

### 2022-04-18 analysis/fit_lognormal_r.R

@@ -1,10 +1,10 @@
 library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("fit_lognormal_base.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+source("../R/sample_incubation.R")
+load("../rdaout/fit_lognormal_base.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -13,16 +13,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_51_sgtf_within <- serialdata %>%
-  filter(week==51, strain=="SGTF", household=="within")
+serialdata_50_sgtf_within <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="within")
 
-serialdata_51_nsgtf_within <- serialdata %>%
-  filter(week==51, strain=="non-SGTF", household=="within")
+serialdata_50_nsgtf_within <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="within")
 
-data_51_sgtf_within <- rep(serialdata_51_sgtf_within$serial, serialdata_51_sgtf_within$n)
-data_51_nsgtf_within <- rep(serialdata_51_nsgtf_within$serial, serialdata_51_nsgtf_within$n)
+data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
+data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
 
-fit_lognormal_r_51_nsgtf_within <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
@@ -42,13 +42,13 @@ fit_lognormal_r_51_nsgtf_within <- lapply(r_nsgtf, function(r) {
   logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
[…]

### 2022-04-18 analysis/fit_lognormal_r_50_between.R

@@ -1,10 +1,10 @@
 library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("rdaout/fit_lognormal_base_comb_between.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+source("../R/sample_incubation.R")
+load("../rdaout/fit_lognormal_base.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -13,24 +13,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_comb_sgtf_between <- serialdata %>%
-  filter(strain=="SGTF", household=="between") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
+serialdata_50_sgtf_between <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="between")
 
-serialdata_comb_nsgtf_between <- serialdata %>%
-  filter(strain=="non-SGTF", household=="between") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
+serialdata_50_nsgtf_between <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="between")
 
-data_comb_sgtf_between <- rep(serialdata_comb_sgtf_between$serial, serialdata_comb_sgtf_between$n)
-data_comb_nsgtf_between <- rep(serialdata_comb_nsgtf_between$serial, serialdata_comb_nsgtf_between$n)
+data_50_sgtf_between <- rep(serialdata_50_sgtf_between$serial, serialdata_50_sgtf_between$n)
+data_50_nsgtf_between <- rep(serialdata_50_nsgtf_between$serial, serialdata_50_nsgtf_between$n)
 
-fit_lognormal_r_comb_nsgtf_between <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_50_nsgtf_between <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
[…]

### 2022-04-18 analysis/fit_lognormal_r_51.R

@@ -1,9 +1,10 @@
+library(bbmle)
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("fit_lognormal_base.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+source("../R/sample_incubation.R")
+load("../rdaout/fit_lognormal_base.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -12,16 +13,16 @@ rho <- 0.75
 
 serialdata <- read_xlsx("serial-netherlands.xlsx")
 
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
+serialdata_51_sgtf_within <- serialdata %>%
+  filter(week==51, strain=="SGTF", household=="within")
 
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
+serialdata_51_nsgtf_within <- serialdata %>%
+  filter(week==51, strain=="non-SGTF", household=="within")
 
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
+data_51_sgtf_within <- rep(serialdata_51_sgtf_within$serial, serialdata_51_sgtf_within$n)
+data_51_nsgtf_within <- rep(serialdata_51_nsgtf_within$serial, serialdata_51_nsgtf_within$n)
 
-fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_51_nsgtf_within <- lapply(r_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
@@ -41,13 +42,13 @@ fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
   logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
[…]

### 2022-04-18 analysis/fit_lognormal_r_51_between.R

@@ -1,9 +1,9 @@
 library(dplyr)
 library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("fit_lognormal_base.rda")
+source("../R/serialfun.R")
+source("../R/fitfun.R")
+source("../R/sample_incubation.R")
+load("../rdaout/fit_lognormal_base.rda")
 
 r_nsgtf <- seq(-0.1, 0, length.out=11)
 r_sgtf <- seq(0.1, 0.2, length.out=11)
@@ -139,4 +139,4 @@ fit_lognormal_r_51_sgtf_between <- lapply(r_sgtf, function(r) {
 }) %>%
   bind_rows
 
-save("fit_lognormal_r_51_sgtf_between", "fit_lognormal_r_51_nsgtf_between", file="fit_lognormal_r_51_between.rda")
+save("fit_lognormal_r_51_sgtf_between", "fit_lognormal_r_51_nsgtf_between", file="../rdaout/fit_lognormal_r_51_between.rda")

### 2022-04-18 figure/figure_reproduction_advantage_new.R

@@ -1,0 +1,216 @@
+library(vroom)
+library(readxl)
+library(tidyr)
+library(dplyr)
+library(lubridate)
+library(zoo)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(mgcv)
+library(mvtnorm)
+library(egg)
+load("rdaout/fit_lognormal_base.rda")
+
+gbar_delta_orig <- 6.4
+gkappa_delta_orig <- (5.4/6.4)^2
+
+weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
+
+nsample <- 1000
+
+cases <- vroom("COVID-19_aantallen_gemeente_per_dag.csv")
+variant <- read_xlsx("variant-netherlands.xlsx")
+
+cases_all <- cases %>%
+  group_by(Date_of_publication) %>%
+  summarize(
+    cases=sum(Total_reported)
+  ) %>%
+  arrange(Date_of_publication) %>%
+  mutate(
+    rollmean=rollmean(cases, 7, fill=NA)
+  ) %>%
+  filter(Date_of_publication >= "2021-11-22", Date_of_publication <= "2022-01-30")
+
+cases_all_weekly <- cases_all %>%
+  mutate(
+    group=cut(Date_of_publication, weekbreak+1)
+  ) %>%
+  group_by(group) %>%
+  summarize(
+    cases=sum(cases),
+    date=max(as.Date(Date_of_publication))
+  )
+
+variant2 <- data.frame(
+  omicron=unlist(variant[9,-c(1:2, 13)]),
+  delta=unlist(variant[5,-c(1:2, 13)]),
+  total=unlist(variant[1,-c(1:2, 13)]),
+  week=c(4, 3, 2, 1, 52, 51, 50, 49, 48, 47),
+  year=c(2022, 2022, 2022, 2022, 2021, 2021, 2021, 2021, 2021, 2021),
+  date=rev(cases_all_weekly$date)
+)
+
+variant3 <- variant2 %>%
+  group_by(year, week) %>%
+  mutate(
+    omicron_prop=omicron/total,
+    omicron_prop_lwr=binom.test(omicron, total)[[4]][1],
+    omicron_prop_upr=binom.test(omicron, total)[[4]][2],
+    delta_prop=delta/total,
+    delta_prop_lwr=binom.test(delta, total)[[4]][1],
+    delta_prop_upr=binom.test(delta, total)[[4]][2]
+  ) %>%
+  merge(
[…]

### 2022-04-18 fit_he_base.R

@@ -1,39 +1,0 @@
-library(bbmle)
-library(dplyr)
-library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-load("calculate_incubation_mle.rda")
-
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
-rho <- 0.75
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
-
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
-
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
-
-fit_he_base_50_sgtf_within <- fitfun_he(data=data_50_sgtf_within, 
-                                        inf_shape=5, 
-                                        inf_mean=5, 
-                                        inf_shift=5,
-                                        logmean_inc=logmean_inc_sgtf, 
-                                        logsd_inc=logsd_inc_sgtf, 
-                                        r=r_sgtf)
-
-fit_he_base_50_nsgtf_within <- fitfun_he(data=data_50_nsgtf_within, 
-                                         inf_shape=5, 
-                                         inf_mean=5, 
-                                         inf_shift=5,
-                                         logmean_inc=logmean_inc_nsgtf, 
-                                         logsd_inc=logsd_inc_nsgtf, 
-                                         r=r_nsgtf)
-
-save("fit_he_base_50_sgtf_within", "fit_he_base_50_nsgtf_within", file="fit_he_base.rda")

### 2022-04-18 fit_he_r.R

@@ -1,110 +1,0 @@
-library(dplyr)
-library(readxl)
-source("serialfun.R")
-source("fitfun.R")
-source("sample_incubation.R")
-load("fit_he_base.rda")
-
-r_nsgtf <- seq(-0.1, 0, length.out=11)
-r_sgtf <- seq(0.1, 0.2, length.out=11)
-
-rho <- 0.75
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
-
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
-
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
-
-fit_he_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
-  print(r)
-  moment_0_nsgtf <- integrate(function(z) {
-    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
-  }, 0, 1000)[[1]]
-  
-  moment_1_nsgtf <- integrate(function(z) {
-    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
-  }, 0, 1000)[[1]]
-  
-  moment_2_nsgtf <- integrate(function(z) {
-    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
-  }, 0, 1000)[[1]]
-  
-  mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
-  var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
-  
-  logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
-  logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
-  
-  ff <- fitfun_he(data=data_50_nsgtf_within, 
[…]

### 2022-04-18 simfun.R

@@ -1,102 +1,0 @@
-sample2 <- function(x, size) {
-  if(length(x)==1) {
-    rep(x, size)
-  } else{
-    sample(x, size, replace=TRUE)
-  }
-}
-
-simfun_egocentric <- function(R=1.7,
-                              k=0.1,
-                              meanlog=1.2,
-                              sdlog=0.5,
-                              household=4) {
-  z <- 0
-  
-  while (z == 0) {
-    z <- rnbinom(1, mu=R, size=k)
-  }
-  
-  gen <- sort(rlnorm(z, meanlog, sdlog))
-  
-  if (z < household) {
-    return(gen)
-  } else {
-    return(gen[1:household])
-  }
-}
-
-simfun_house <- function(R=1.7,
-                         k=0.1,
-                         meanlog=1.2,
-                         sdlog=0.5,
-                         household=4) {
-  V <- 1:household
-  
-  initial_infected <- 1
-  
-  imax <- household
-  
-  queue_v <- queue_t <- queue_infector <- NA
-  
-  queue_v <- initial_infected 
-  queue_t <- 0
-  
-  t_infected <- rep(NA, household)
-  t_infected[initial_infected] <- 0
-  
-  t_gillespie <- NULL
-  c_infected <- 0
-  
-  done <- rep(FALSE, household)
-  infected_by <- rep(NA, household)
-  
-  stop <- FALSE
-  
-  while (!stop) {
-    j.index <- which.min(queue_t)
-    j <- queue_v[j.index]
-    
-    infected_by[j] <- queue_infector[j.index]
-    t_infected[j] <- queue_t[j.index]
-    
-    t <- queue_t[j.index]; t_gillespie <- c(t_gillespie, t)
-    
-    c_infected <- c_infected +1
-    
-    n <- V[V != j]
-    
-    z <- rnbinom(1, mu=R, size=k)
-    
-    if (j==1) {
-      ## force the first infection to have some infections to start an outbreak
-      while (z == 0) {
-        z <- rnbinom(1, mu=R, size=k)
-      }
-    }
-    
-    if (z > 0) {
-      queue_v <- c(queue_v, sample2(n, z))
-      queue_infector <- c(queue_infector, rep(j, z))
-    }
-    
-    gen <- rlnorm(z, meanlog, sdlog)
-    
-    if (z > 0) {
-      queue_t <- c(queue_t, t + gen)
-    }
-    
[…]

### 2022-04-18 simulate_generation_household.R

@@ -1,52 +1,0 @@
-source("simfun.R")
-load("fit_lognormal_base.rda")
-
-nsim <- 2000
-
-kvec <- c(0.1, 1, 10)
-householdvec <- c(4, 8, 12, 16, 20)
-ratiovec <- c(1.5, 2, 2.5)
-pardata <- expand.grid(kvec, householdvec, ratiovec)
-
-R_delta <- 0.8
-
-logmean <- coef(fit_lognormal_base_50_nsgtf_within)[1]
-logsd <- coef(fit_lognormal_base_50_nsgtf_within)[2]
-
-reslist <- vector('list', nrow(pardata))
-
-set.seed(101)
-for (i in 1:nrow(pardata)) {
-  print(i)
-  delta_egocentric <- unlist(replicate(nsim, simfun_egocentric(R=R_delta, k=pardata[i,1], logmean, logsd, household=pardata[i,2])))
-  delta_house <- unlist(replicate(nsim, simfun_house(R=R_delta, k=pardata[i,1], logmean, logsd, household=pardata[i,2])))
-  
-  omicron_egocentric <- unlist(replicate(nsim, simfun_egocentric(R=R_delta*pardata[i,3], k=pardata[i,1], logmean, logsd, household=pardata[i,2])))
-  omicron_house <- unlist(replicate(nsim, simfun_house(R=R_delta*pardata[i,3], k=pardata[i,1], logmean, logsd, household=pardata[i,2])))
-  
-  reslist[[i]] <- data.frame(
-    mean=c(mean(delta_egocentric), mean(delta_house), mean(omicron_egocentric), mean(omicron_house)),
-    lwr=c(
-      t.test(delta_egocentric)[[4]][[1]],
-      t.test(delta_house)[[4]][[1]],
-      t.test(omicron_egocentric)[[4]][[1]],
-      t.test(omicron_house)[[4]][[1]]
-    ),
-    upr=c(
-      t.test(delta_egocentric)[[4]][[2]],
-      t.test(delta_house)[[4]][[2]],
-      t.test(omicron_egocentric)[[4]][[2]],
-      t.test(omicron_house)[[4]][[2]]
-    ),
-    strain=c("Delta", "Delta", "Omicron", "Omicron"),
-    type=c("Egocentric", "Household", "Egocentric", "Household"),
-    k=pardata[i,1],
-    household=pardata[i,2],
-    ratio=pardata[i,3]
-  )
-}
-
-simulate_generation_household <- reslist %>%
[…]

### 2022-04-22 figure/figure_compare_rho.R
take a brak

@@ -4,9 +4,12 @@ library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(egg)
 library(gridExtra)
+
+library(shellpipes)
+startGraphics(width=6, heigh=4)
 load("rdaout/fit_lognormal_base_comb_rho.rda")
 
-g1 <- ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
+print(ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
   geom_ribbon(aes(rho, ymin=lwr, ymax=upr), alpha=0.2) +
   geom_line(aes(rho, est), lwd=1, col="black") +
   geom_ribbon(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, ymin=lwr, ymax=upr), fill="orange", alpha=0.2) +
@@ -16,5 +19,5 @@ g1 <- ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
   theme(
     panel.grid = element_blank()
   )
+)
 
-ggsave("figure_compare_rho.pdf", g1, width=6, heigh=4)

### 2022-04-22 figure/figure_compare_stratified.R

@@ -4,7 +4,7 @@ library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(egg)
 library(gridExtra)
-source("serialfun.R")
+source("R/serialfun.R")
 load("rdaout/fit_lognormal_r.rda")
 load("rdaout/fit_lognormal_r_50_between.rda")
 load("rdaout/fit_lognormal_r_51_between.rda")
@@ -92,4 +92,4 @@ g8 <- ggplot(filter(fit_lognormal_r_51_sgtf_between, param=="mean")) +
 
 gcomb <- ggarrange(g1, g2, g3, g4, g5, g6, g7, g8, nrow=2, draw=FALSE)
 
-ggsave("figure_compare_stratified.pdf", gcomb, width=12, height=6)
+saveGG(gcomb, width=12, height=6)

### 2022-04-22 figure/figure_diagram.R

@@ -1,5 +1,7 @@
 library(ggplot2); theme_set(theme_bw(base_family="Times"))
 library(egg)
+library(shellpipes)
+startGraphics()
 
 size <- 4
 stroke <- 1
@@ -140,5 +142,4 @@ g2 <- ggplot() +
   )
 
 gcomb <- ggarrange(g1, g2, nrow=2, draw=FALSE, labels=c("A", "B"))
-
-ggsave("figure_diagram.pdf", gcomb, width=9, height=10)
+saveGG(gcomb)

### 2022-04-22 figure/figure_reproduction_advantage.R

@@ -9,14 +9,16 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(mgcv)
 library(mvtnorm)
 library(egg)
+
+library(shellpipes)
 load("rdaout/fit_lognormal_base_comb_within.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 
 nsample <- 1000
 
-cases <- vroom("COVID-19_aantallen_gemeente_per_dag.csv")
-variant <- read_xlsx("variant-netherlands.xlsx")
+cases <- vroom("data/COVID-19_aantallen_gemeente_per_dag.csv")
+variant <- read_xlsx("data/variant-netherlands.xlsx")
 
 cases_all <- cases %>%
   group_by(Date_of_publication) %>%
@@ -198,7 +200,7 @@ g5 <- ggplot(advantagesumm2) +
 
 gfinal <- ggarrange(g1, g2, nrow=1, labels=c("A", "B"), draw=FALSE)
 
-ggsave("figure_reproduction_advantage.pdf", gfinal, width=8.1, height=3)
+saveGG(gfinal, width=8.1, height=3)
 
 gsupp <- ggarrange(g5, g3, g4, nrow=1, labels=c("A", "B", "C"), draw=FALSE)
 

### 2022-04-22 figure/figure_reproduction_advantage_between.R

@@ -9,14 +9,16 @@ library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(mgcv)
 library(mvtnorm)
 library(egg)
+
+library(shellpipes)
 load("rdaout/fit_lognormal_base_comb_between.rda")
 
 weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
 
 nsample <- 1000
 
-cases <- vroom("COVID-19_aantallen_gemeente_per_dag.csv")
-variant <- read_xlsx("variant-netherlands.xlsx")
+cases <- vroom("data/COVID-19_aantallen_gemeente_per_dag.csv")
+variant <- read_xlsx("data/variant-netherlands.xlsx")
 
 cases_all <- cases %>%
   group_by(Date_of_publication) %>%
@@ -167,4 +169,4 @@ g1 <- ggplot(filter(advantagesumm, key %in% c("R_advantage", "R_delta", "R_omicr
 
 gfinal <- ggarrange(g1, g2, nrow=1, labels=c("A", "B"), draw=FALSE)
 
-ggsave("figure_reproduction_advantage_between.pdf", gfinal, width=8, height=3)
+saveGG(gfinal, width=8, height=3)

### 2022-04-22 omicron-generation.tex
one more s
weekend shutdown (xiangshan)

@@
 \begin{figure}[!tp]
-\includegraphics[width=\textwidth]{figure_diagram.pdf}
+\includegraphics[width=\textwidth]{figure/figure_diagram.ggp.pdf}
 \caption{
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figure_epidemic.pdf}
+\includegraphics[width=\textwidth]{figure/figure_epidemic.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_incubation.pdf}
+\includegraphics[width=\textwidth]{figure/figure_incubation.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_compare_main.pdf}
+\includegraphics[width=\textwidth]{figure/figure_compare_main.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_reproduction_advantage.pdf}
+\includegraphics[width=\textwidth]{figure/figure_reproduction_advantage.ggp.pdf}
 \caption{
@@
 Assuming a constant growth rate allows us to simplify this complexity because the reproduction number also stays constant during the exponential phase.
-Future studies should
+Extending our framework to account for time-varying growth rates can provide more accurate tools for inferring epidemiological distributions.
 
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_compare_rho.pdf}
+\includegraphics[width=\textwidth]{figure/figure_compare_rho.Rout.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_compare_stratified.pdf}
+\includegraphics[width=\textwidth]{figure/figure_compare_stratified.ggp.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure_reproduction_advantage_between.pdf}
+\includegraphics[width=\textwidth]{figure/figure_reproduction_advantage_between.pdf}
 \caption{

### 2022-05-06 omicron-generation.tex
refactoring

@@
 For example, when the epidemic is growing, there will be more recent infections. 
-Therefore, we are more likely to observe recently infected individuals among a cohort of infectors who developed symptoms at the same---we refer to this effect as the ``dynamical bias''.
+Therefore, we are more likely to observe recently infected individuals among a cohort of infectors who developed symptoms at the same time---we refer to this effect as the ``dynamical bias''.
 In this case, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
@@
 
-Publicly available data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates or symptom onset dates.
-The data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (week 50, 13--19 December 2021, and week 51, 20--26 December 2021), and the type of transmission (within or between households).
+Publicly available data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates, symptom onset dates, or age.
+The original article presented serial-interval estimates stratified by the vaccination status in supplementary materials, but stratified data are not publicly available; we rely on publicly available data to keep the analysis simple and to focus on the qualitative effects of dynamical biases.
+The aggregated data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (week 50, 13--19 December 2021, and week 51, 20--26 December 2021), and the type of transmission (within- or between-household transmission).
 In the main text, we combine data from weeks 50 and 51 of 2021 (13--26 December) and present a stratified analysis in Supplementary Material.
@@
 
-The incubation-period distributions from 513 individuals (258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022, were previously reported in \cite{backer2021omicron}.
+The incubation-period distributions from 513 individuals (consisting of 258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022, were previously reported in \cite{backer2021omicron}.
+We note that these data sets cover a wider range of periods and therefore are different from the serial-interval data.
 \cite{backer2021omicron} used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates.
@@
 Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
-Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed them to conclude that the mean generation and serial intervals are identical.
+Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed the authors to conclude that the mean generation and serial intervals are identical.
 For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time, \fref{diagram}B), and drop the subscript $\mathrm{inf}$.
@@
 However, these measurements were done during a period when the incidence of Omicron was increasing while the increasing of Delta was decreasing (\fref{epidemic}).
-Thus, dynamical bias would be expected to lead to shorter observed (backward) incubation periods in Omicron, and longer observed incubation periods in Delta. 
-When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions with a mean of 4.1 days (\fref{incubation}B);
+% Thus, dynamical bias would be expected to lead to shorter observed (backward) incubation periods in Omicron, and longer observed incubation periods in Delta. 
+When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions with a mean of 4.1 days (95\% CI: 3.8--4.4 days) for the Delta variant and 4.2 days (95\% CI: 3.6--4.9 days) for the Omicron variant \fref{incubation}B);
 in this case, the difference between the mean backward and forward incubation periods correspond to $-22\%$ and $7\%$ bias for the Omicron and Delta variants, respectively.
@@
 while the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$ (\fref{incubation}D).
+We estimate wider uncertanties for the Omicron variant because higher absolute growth rate values lead to larger differences between the backward and forward incubation periods.
 
@@
 This study presents a simple method for accounting for dynamical biases in inferring incubation-period distributions based on epidemic growth rates.
-In practice, it is easier to measure the backward incubation-period distribution in practice because we can directly observe symptom onset. 
+In practice, it is easier to measure the backward incubation-period distribution because we can directly observe symptom onset. 
 Therefore, the observed incubation-period distributions are generally expected to be biased, and similar kinds of corrections will be necessary to accurately estimate the incubation-period distribution.
@@
 
-A few studies have suggested the the incubation period of the Omicron variant may be shorter than that of the Delta variant.
+A few studies have suggested that the incubation period of the Omicron variant may be shorter than that of the Delta variant.
 The median estimates of the Omicron incubation period typically range between 3--4 days, consistent with earlier findings of \citep{backer2021omicron}. 
@@
 We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
+Previous simulations showed that network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
 While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
-The network effect may also explain a decrease in the mean generation interval between week 50 and 51 (13--26 December 2021), especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
 Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
-Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and faster within-host clearance of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of generation intervals.
+Other factors, such as more stringent contact tracing measures against the Omicron variant in the Netherlands \citep{backer2021omicron} faster within-host clearance of the Omicron variant \citep{hay2022viral}, and viral kinetics of reinfection and breakthrough infections, also likely contributed to shortening of generation intervals.
 
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant ($> 4$ fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate that the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+With our generation-interval estimates, we estimate the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
 
@@
 Assuming a constant growth rate allows us to simplify this complexity because the reproduction number also stays constant during the exponential phase.
-Extending our framework to account for time-varying growth rates can provide more accurate tools for inferring epidemiological distributions.
+Extending our framework to account for time-varying growth rates can provide more accurate tools for inferring epidemiological delay distributions.
+
+Finally, we do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
+In general, epidemic growth rates likely differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
+We are not able to perform stratified analyses because individiaul-level information was not publicly available.
+Therefore, while we estimate unclear differences in incubation-period distributions between Delta and Omicron infections, controlling for other covariates, such as age and immune history, may help better characterize differences in Delta and Omicron infections.
 

### 2022-05-06 analysis/fit_lognormal_base_comb_between.R

@@ -1,47 +1,0 @@
-library(bbmle)
-library(dplyr)
-library(readxl)
-source("../R/serialfun.R")
-source("../R/fitfun.R")
-load("../rdaout/calculate_incubation_mle.rda")
-
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
-rho <- 0.75
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_comb_sgtf_between <- serialdata %>%
-  filter(strain=="SGTF", household=="between") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
-
-serialdata_comb_nsgtf_between <- serialdata %>%
-  filter(strain=="non-SGTF", household=="between") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
-
-data_comb_sgtf_between <- rep(serialdata_comb_sgtf_between$serial, serialdata_comb_sgtf_between$n)
-data_comb_nsgtf_between <- rep(serialdata_comb_nsgtf_between$serial, serialdata_comb_nsgtf_between$n)
-
-fit_lognormal_base_comb_sgtf_between <- fitfun_lognormal(data=data_comb_sgtf_between, 
-                                                      logmean_gen=logmean_inc_sgtf, 
-                                                      logsd_gen=logsd_inc_sgtf, 
-                                                      logmean_inc=logmean_inc_sgtf, 
-                                                      logsd_inc=logsd_inc_sgtf, 
-                                                      rho=rho, 
-                                                      r=r_sgtf)
-
-fit_lognormal_base_comb_nsgtf_between <- fitfun_lognormal(data=data_comb_nsgtf_between, 
-                                                       logmean_gen=logmean_inc_nsgtf, 
-                                                       logsd_gen=logsd_inc_nsgtf, 
-                                                       logmean_inc=logmean_inc_nsgtf, 
-                                                       logsd_inc=logsd_inc_nsgtf, 
-                                                       rho=rho, 
-                                                       r=r_nsgtf)
-
-save("fit_lognormal_base_comb_sgtf_between", "fit_lognormal_base_comb_nsgtf_between", file="../rdaout/fit_lognormal_base_comb_between.rda")

### 2022-05-06 figure/figure_reproduction_advantage.R

@@ -108,7 +108,6 @@ advantagedata <- lapply(1:nsample, function(x) {
   R_omicron <- tail(i2, -n)/sapply(1:(length(i2)-n), function(y) sum(i2[y:(y+n-1)] * rev(gen2)))
   R_omicron_naive <- tail(i2, -n)/sapply(1:(length(i2)-n), function(y) sum(i2[y:(y+n-1)] * rev(gen1)))
   
-  
   data.frame(
     R_delta=R_delta,
     R_omicron=R_omicron,

### 2022-05-06 scripts/baseparam.R

@@ -1,0 +1,8 @@
+r_nsgtf <- -0.05
+r_sgtf <- 0.15
+
+rvec_nsgtf <- seq(-0.1, 0, length.out=11)
+rvec_sgtf <- seq(0.1, 0.2, length.out=11)
+
+rho <- 0.75
+rhovec <- seq(0.5, 0.85, length.out=11)

### 2022-05-06 scripts/fit_lognormal_base.R

@@ -1,0 +1,26 @@
+library(bbmle)
+library(dplyr)
+library(readxl)
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/calculate_incubation_mle.rda")
+
+fit_lognormal_base_50_sgtf_within <- fitfun_lognormal(data=data_50_sgtf_within, 
+                                                      logmean_gen=logmean_inc_sgtf, 
+                                                      logsd_gen=logsd_inc_sgtf, 
+                                                      logmean_inc=logmean_inc_sgtf, 
+                                                      logsd_inc=logsd_inc_sgtf, 
+                                                      rho=rho, 
+                                                      r=r_sgtf)
+
+fit_lognormal_base_50_nsgtf_within <- fitfun_lognormal(data=data_50_nsgtf_within, 
+                                                       logmean_gen=logmean_inc_nsgtf, 
+                                                       logsd_gen=logsd_inc_nsgtf, 
+                                                       logmean_inc=logmean_inc_nsgtf, 
+                                                       logsd_inc=logsd_inc_nsgtf, 
+                                                       rho=rho, 
+                                                       r=r_nsgtf)
+
+save("fit_lognormal_base_50_sgtf_within", "fit_lognormal_base_50_nsgtf_within", file="../rdaout/fit_lognormal_base.rda")

### 2022-05-06 scripts/fit_lognormal_base_50_between.R

@@ -1,24 +1,9 @@
 library(bbmle)
-library(dplyr)
-library(readxl)
-source("../R/serialfun.R")
-source("../R/fitfun.R")
-load("../rdaout/calculate_incubation_mle.rda")
-
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
-rho <- 0.75
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_50_sgtf_between <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="between")
-
-serialdata_50_nsgtf_between <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="between")
-
-data_50_sgtf_between <- rep(serialdata_50_sgtf_between$serial, serialdata_50_sgtf_between$n)
-data_50_nsgtf_between <- rep(serialdata_50_nsgtf_between$serial, serialdata_50_nsgtf_between$n)
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/calculate_incubation_mle.rda")
 
 fit_lognormal_base_50_sgtf_between <- fitfun_lognormal(data=data_50_sgtf_between, 
                                                       logmean_gen=logmean_inc_sgtf, 

### 2022-05-06 scripts/fit_lognormal_base_comb_between.R

@@ -1,26 +1,11 @@
 library(bbmle)
-library(dplyr)
-library(readxl)
-source("../R/serialfun.R")
-source("../R/fitfun.R")
-load("../rdaout/calculate_incubation_mle.rda")
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/calculate_incubation_mle.rda")
 
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
-rho <- 0.75
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
-
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
-
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
-
-fit_lognormal_base_50_sgtf_within <- fitfun_lognormal(data=data_50_sgtf_within, 
+fit_lognormal_base_comb_sgtf_between <- fitfun_lognormal(data=data_comb_sgtf_between, 
                                                       logmean_gen=logmean_inc_sgtf, 
                                                       logsd_gen=logsd_inc_sgtf, 
                                                       logmean_inc=logmean_inc_sgtf, 
@@ -28,7 +13,7 @@ fit_lognormal_base_50_sgtf_within <- fitfun_lognormal(data=data_50_sgtf_within,
                                                       rho=rho, 
                                                       r=r_sgtf)
 
-fit_lognormal_base_50_nsgtf_within <- fitfun_lognormal(data=data_50_nsgtf_within, 
+fit_lognormal_base_comb_nsgtf_between <- fitfun_lognormal(data=data_comb_nsgtf_between, 
                                                        logmean_gen=logmean_inc_nsgtf, 
                                                        logsd_gen=logsd_inc_nsgtf, 
                                                        logmean_inc=logmean_inc_nsgtf, 
@@ -36,4 +21,4 @@ fit_lognormal_base_50_nsgtf_within <- fitfun_lognormal(data=data_50_nsgtf_within
                                                        rho=rho, 
[…]

### 2022-05-06 scripts/fit_lognormal_base_comb_rho.R

@@ -1,33 +1,10 @@
 library(bbmle)
-library(dplyr)
-library(readxl)
-source("../R/serialfun.R")
-source("../R/fitfun.R")
-load("../rdaout/calculate_incubation_mle.rda")
-load("../rdaout/fit_lognormal_base_comb_within.rda")
-
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
-rhovec <- seq(0.5, 0.85, length.out=11)
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_comb_sgtf_within <- serialdata %>%
-  filter(strain=="SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
-
-serialdata_comb_nsgtf_within <- serialdata %>%
-  filter(strain=="non-SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
-  summarize(
-    n=sum(n)
-  )
-
-data_comb_sgtf_within <- rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n)
-data_comb_nsgtf_within <- rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n)
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/calculate_incubation_mle.rda")
+load("rdaout/fit_lognormal_base_comb_within.rda")
 
 fit_lognormal_base_rho_comb_nsgtf_within <- lapply(rhovec, function(rho) {
   print(rho)

### 2022-05-06 scripts/fit_lognormal_r.R

@@ -1,28 +1,12 @@
 library(bbmle)
-library(dplyr)
-library(readxl)
-source("../R/serialfun.R")
-source("../R/fitfun.R")
-source("../R/sample_incubation.R")
-load("../rdaout/fit_lognormal_base.rda")
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("R/sample_incubation.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/fit_lognormal_base.rda")
 
-r_nsgtf <- seq(-0.1, 0, length.out=11)
-r_sgtf <- seq(0.1, 0.2, length.out=11)
-
-rho <- 0.75
-
-serialdata <- read_xlsx("serial-netherlands.xlsx")
-
-serialdata_50_sgtf_within <- serialdata %>%
-  filter(week==50, strain=="SGTF", household=="within")
-
-serialdata_50_nsgtf_within <- serialdata %>%
-  filter(week==50, strain=="non-SGTF", household=="within")
-
-data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
-data_50_nsgtf_within <- rep(serialdata_50_nsgtf_within$serial, serialdata_50_nsgtf_within$n)
-
-fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
+fit_lognormal_r_50_nsgtf_within <- lapply(rvec_nsgtf, function(r) {
   print(r)
   moment_0_nsgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
@@ -72,7 +56,7 @@ fit_lognormal_r_50_nsgtf_within <- lapply(r_nsgtf, function(r) {
 }) %>%
   bind_rows
 
-fit_lognormal_r_50_sgtf_within <- lapply(r_sgtf, function(r) {
+fit_lognormal_r_50_sgtf_within <- lapply(rvec_sgtf, function(r) {
   print(r)
   moment_0_sgtf <- integrate(function(z) {
     dweibull(z, shape=backward_shape_sgtf, scale=backward_scale_sgtf) * exp(r * z)

### 2022-05-06 scripts/serialdata.R

@@ -1,0 +1,71 @@
+library(dplyr)
+library(readxl)
+
+serialdata <- read_xlsx("data/serial-netherlands.xlsx")
+
+serialdata_50_sgtf_within <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="within")
+
+serialdata_50_nsgtf_within <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="within")
+
+serialdata_50_sgtf_between <- serialdata %>%
+  filter(week==50, strain=="SGTF", household=="between")
+
+serialdata_50_nsgtf_between <- serialdata %>%
+  filter(week==50, strain=="non-SGTF", household=="between")
+
+serialdata_51_sgtf_within <- serialdata %>%
+  filter(week==51, strain=="SGTF", household=="within")
+
+serialdata_51_nsgtf_within <- serialdata %>%
+  filter(week==51, strain=="non-SGTF", household=="within")
+
+serialdata_51_sgtf_between <- serialdata %>%
+  filter(week==51, strain=="SGTF", household=="between")
+
+serialdata_51_nsgtf_between <- serialdata %>%
+  filter(week==51, strain=="non-SGTF", household=="between")
+
+serialdata_comb_sgtf_within <- serialdata %>%
+  filter(strain=="SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_within <- serialdata %>%
+  filter(strain=="non-SGTF", household=="within") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_sgtf_between <- serialdata %>%
+  filter(strain=="SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+serialdata_comb_nsgtf_between <- serialdata %>%
+  filter(strain=="non-SGTF", household=="between") %>%
+  group_by(serial, strain, household) %>%
+  summarize(
+    n=sum(n)
+  )
+
+data_50_sgtf_within <- rep(serialdata_50_sgtf_within$serial, serialdata_50_sgtf_within$n)
[…]

### 2022-05-20 omicron-generation.tex
refactoring a whole lot

@@
 %\usepackage{adjustbox}
+
+\usepackage{grffile}
 
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figure/figure_epidemic.pdf}
+\includegraphics[width=\textwidth]{figure/figure_epidemic.ggp.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_incubation.pdf}
+\includegraphics[width=\textwidth]{figure/figure_incubation.ggp.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_compare_main.pdf}
+\includegraphics[width=\textwidth]{figure/figure_compare_main.ggp.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_compare_rho.Rout.pdf}
+\includegraphics[width=\textwidth]{figure/figure_compare_rho.ggp.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_reproduction_advantage_between.pdf}
+\includegraphics[width=\textwidth]{figure/figure_reproduction_advantage_between.ggp.pdf}
 \caption{

### 2022-05-20 figure/advantage.R

@@ -1,0 +1,56 @@
+library(bbmle)
+library(tidyr)
+library(dplyr)
+library(mvtnorm)
+library(shellpipes)
+rpcall("advantage.Rout advantage.R cases.rda rdaout/fit_lognormal_base_comb_within.rda")
+
+sourceFiles()
+loadEnvironments()
+
+set.seed(101)
+lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_nsgtf_within)[1:2], sigma=vcov(fit_lognormal_base_comb_nsgtf_within))
+lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_sgtf_within)[1:2], sigma=vcov(fit_lognormal_base_comb_sgtf_within))
+
+advantagedata <- lapply(1:nsample, function(x) {
+  i1 <- c(exp(gfit1_p1 %*% c(gfit1_sim[x,]))/7)
+  i2 <- c(exp(gfit2_p1 %*% c(gfit2_sim[x,]))/7)
+  
+  gen1 <- diff(plnorm(seq(0, 14, by=0.5), lognormal_sim1[x,1], lognormal_sim1[x,2]))
+  gen2 <- diff(plnorm(seq(0, 14, by=0.5), lognormal_sim2[x,1], lognormal_sim2[x,2]))
+  
+  n <- length(gen1)
+  
+  R_delta <- tail(i1, -n)/sapply(1:(length(i1)-n), function(y) sum(i1[y:(y+n-1)] * rev(gen1)))
+  R_omicron <- tail(i2, -n)/sapply(1:(length(i2)-n), function(y) sum(i2[y:(y+n-1)] * rev(gen2)))
+  R_omicron_naive <- tail(i2, -n)/sapply(1:(length(i2)-n), function(y) sum(i2[y:(y+n-1)] * rev(gen1)))
+  
+  data.frame(
+    R_delta=R_delta,
+    R_omicron=R_omicron,
+    R_omicron_naive=R_omicron_naive,
+    R_advantage=R_omicron/R_delta,
+    R_advantage_naive=R_omicron_naive/R_delta,
+    r_delta=tail(gfit1_post[,x], -n),
+    r_omicron=tail(gfit2_post[,x], -n),
+    r_advantage=tail(advantage_post[,x], -n),
+    sim=x,
+    time=tail(seq(1, 10, by=1/14), -n)
+  )
+}) %>%
+  bind_rows
+
[…]

### 2022-05-20 figure/advantage_between.R

@@ -1,0 +1,56 @@
+library(bbmle)
+library(tidyr)
+library(dplyr)
+library(mvtnorm)
+library(shellpipes)
+rpcall("advantage.Rout advantage.R cases.rda rdaout/fit_lognormal_base_comb_between.rda")
+
+sourceFiles()
+loadEnvironments()
+
+set.seed(101)
+lognormal_sim1 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_nsgtf_between)[1:2], sigma=vcov(fit_lognormal_base_comb_nsgtf_between))
+lognormal_sim2 <- rmvnorm(nsample, mean=coef(fit_lognormal_base_comb_sgtf_between)[1:2], sigma=vcov(fit_lognormal_base_comb_sgtf_between))
+
+advantagedata <- lapply(1:nsample, function(x) {
+  i1 <- c(exp(gfit1_p1 %*% c(gfit1_sim[x,]))/7)
+  i2 <- c(exp(gfit2_p1 %*% c(gfit2_sim[x,]))/7)
+  
+  gen1 <- diff(plnorm(seq(0, 14, by=0.5), lognormal_sim1[x,1], lognormal_sim1[x,2]))
+  gen2 <- diff(plnorm(seq(0, 14, by=0.5), lognormal_sim2[x,1], lognormal_sim2[x,2]))
+  
+  n <- length(gen1)
+  
+  R_delta <- tail(i1, -n)/sapply(1:(length(i1)-n), function(y) sum(i1[y:(y+n-1)] * rev(gen1)))
+  R_omicron <- tail(i2, -n)/sapply(1:(length(i2)-n), function(y) sum(i2[y:(y+n-1)] * rev(gen2)))
+  R_omicron_naive <- tail(i2, -n)/sapply(1:(length(i2)-n), function(y) sum(i2[y:(y+n-1)] * rev(gen1)))
+  
+  data.frame(
+    R_delta=R_delta,
+    R_omicron=R_omicron,
+    R_omicron_naive=R_omicron_naive,
+    R_advantage=R_omicron/R_delta,
+    R_advantage_naive=R_omicron_naive/R_delta,
+    r_delta=tail(gfit1_post[,x], -n),
+    r_omicron=tail(gfit2_post[,x], -n),
+    r_advantage=tail(advantage_post[,x], -n),
+    sim=x,
+    time=tail(seq(1, 10, by=1/14), -n)
+  )
+}) %>%
+  bind_rows
+
[…]

### 2022-05-20 figure/cases.R

@@ -1,0 +1,130 @@
+library(vroom)
+library(readxl)
+library(dplyr)
+library(lubridate)
+library(zoo)
+library(mgcv)
+library(mvtnorm)
+library(shellpipes)
+## Files passed to shellpipes by make (or run this line interactively)
+rpcall("cases.Rout cases.R")
+
+sourceFiles()
+loadEnvironments()
+
+datevec <- seq(as.Date("2021-11-28"), as.Date("2022-01-30"), by=1)
+weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
+
+nsample <- 1000
+
+cases <- vroom("data/COVID-19_aantallen_gemeente_per_dag.csv")
+variant <- read_xlsx("data/variant-netherlands.xlsx")
+
+cases_all <- cases %>%
+  group_by(Date_of_publication) %>%
+  summarize(
+    cases=sum(Total_reported)
+  ) %>%
+  arrange(Date_of_publication) %>%
+  mutate(
+    rollmean=rollmean(cases, 7, fill=NA)
+  ) %>%
+  filter(Date_of_publication >= "2021-11-22", Date_of_publication <= "2022-01-30")
+
+cases_all_weekly <- cases_all %>%
+  mutate(
+    group=cut(Date_of_publication, weekbreak+1)
+  ) %>%
+  group_by(group) %>%
+  summarize(
+    cases=sum(cases),
+    date=max(as.Date(Date_of_publication))
+  )
+
+variant2 <- data.frame(
+  omicron=unlist(variant[9,-c(1:2, 13)]),
+  delta=unlist(variant[5,-c(1:2, 13)]),
+  total=unlist(variant[1,-c(1:2, 13)]),
+  week=c(4, 3, 2, 1, 52, 51, 50, 49, 48, 47),
+  year=c(2022, 2022, 2022, 2022, 2021, 2021, 2021, 2021, 2021, 2021),
+  date=rev(cases_all_weekly$date)
+)
+
+variant3 <- variant2 %>%
+  group_by(year, week) %>%
+  mutate(
+    omicron_prop=omicron/total,
+    omicron_prop_lwr=binom.test(omicron, total)[[4]][1],
+    omicron_prop_upr=binom.test(omicron, total)[[4]][2],
+    delta_prop=delta/total,
[…]

### 2022-05-20 figure/figure_compare_main.R

@@ -4,12 +4,12 @@ library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(egg)
 library(gridExtra)
-source("R/serialfun.R")
-load("rdaout/calculate_incubation_mle.rda")
-load("rdaout/fit_lognormal_base_comb_within.rda")
-load("rdaout/fit_lognormal_base_comb_between.rda")
-load("rdaout/fit_lognormal_comb_within_r.rda")
-load("rdaout/fit_lognormal_comb_between_r.rda")
+library(shellpipes)
+## Files passed to shellpipes by make (or run this line interactively)
+rpcall("figure_compare_main.Rout figure_compare_main.R R/serialfun.R scripts/serialdata.R rdaout/calculate_incubation_mle.rda rdaout/fit_lognormal_base_comb_within.rda rdaout/fit_lognormal_base_comb_between.rda rdaout/fit_lognormal_comb_within_r.rda rdaout/fit_lognormal_comb_between_r.rda")
+
+sourceFiles()
+loadEnvironments()
 
 filter(fit_lognormal_r_comb_nsgtf_within, param=="mean", r==-0.05)
 filter(fit_lognormal_r_comb_sgtf_within, param=="mean", round(r, 2)==0.15)
@@ -17,48 +17,18 @@ filter(fit_lognormal_r_comb_sgtf_within, param=="mean", round(r, 2)==0.15)
 filter(fit_lognormal_r_comb_nsgtf_between, param=="mean", r==-0.05)
 filter(fit_lognormal_r_comb_sgtf_between, param=="mean", round(r, 2)==0.15)
 
-serialdata <- read_xlsx("data/serial-netherlands.xlsx")
+xmin <- -6
+xmax <- 16
+sstep <- 0.2
+xvec <- seq(xmin, xmax, by=sstep)
 
-xvec <- seq(-6, 16, by=0.2)
+# t.test(rep(serialdata_comb_sgtf_within$serial, serialdata_comb_sgtf_within$n))
 
-ivec <- seq(0, 16, by=0.2)
-gvec <- seq(0, 16, by=0.2)
+# t.test(rep(serialdata_comb_nsgtf_within$serial, serialdata_comb_nsgtf_within$n))
 
-serialdata_comb_sgtf_within <- serialdata %>%
-  filter(strain=="SGTF", household=="within") %>%
-  group_by(serial, strain, household) %>%
[…]

### 2022-05-20 figure/figure_compare_rho.R

@@ -6,10 +6,12 @@ library(egg)
 library(gridExtra)
 
 library(shellpipes)
-startGraphics(width=6, heigh=4)
-load("rdaout/fit_lognormal_base_comb_rho.rda")
+rpcall("figure_compare_rho.Rout figure_compare_rho.R advantage.rda rdaout/fit_lognormal_base_comb_rho.rda")
 
-print(ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
+sourceFiles()
+loadEnvironments()
+
+g1 <- ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
   geom_ribbon(aes(rho, ymin=lwr, ymax=upr), alpha=0.2) +
   geom_line(aes(rho, est), lwd=1, col="black") +
   geom_ribbon(data=filter(fit_lognormal_base_rho_comb_sgtf_within, param=="mean", type=="Omicron"), aes(rho, ymin=lwr, ymax=upr), fill="orange", alpha=0.2) +
@@ -19,5 +21,5 @@ print(ggplot(filter(fit_lognormal_base_rho_comb_nsgtf_within, param=="mean"))  +
   theme(
     panel.grid = element_blank()
   )
-)
 
+saveGG(g1, width=6, heigh=4)

### 2022-05-20 figure/figure_compare_stratified.R

@@ -4,11 +4,12 @@ library(readxl)
 library(ggplot2); theme_set(theme_bw(base_family = "Times"))
 library(egg)
 library(gridExtra)
-source("R/serialfun.R")
-load("rdaout/fit_lognormal_r.rda")
-load("rdaout/fit_lognormal_r_50_between.rda")
-load("rdaout/fit_lognormal_r_51_between.rda")
-load("rdaout/fit_lognormal_r_51.rda")
+
+library(shellpipes)
+rpcall("figure_compare_stratified.Rout figure_compare_stratified.R R/serialfun.R rdaout/fit_lognormal_r_50_within.rda rdaout/fit_lognormal_r_50_between.rda rdaout/fit_lognormal_r_51_within.rda rdaout/fit_lognormal_r_51_between.rda")
+
+sourceFiles()
+loadEnvironments()
 
 g1 <- ggplot(filter(fit_lognormal_r_50_nsgtf_within, param=="mean")) +
   annotate("text", x=-Inf, y=Inf, hjust=-0.05, vjust=1.5, label="Week 50, within-household") +

### 2022-05-20 figure/figure_diagram.R

@@ -142,4 +142,4 @@ g2 <- ggplot() +
   )
 
 gcomb <- ggarrange(g1, g2, nrow=2, draw=FALSE, labels=c("A", "B"))
-saveGG(gcomb)
+saveGG(gcomb, width=9, height=10)

### 2022-05-20 figure/figure_epidemic.R

@@ -1,74 +1,10 @@
-library(vroom)
-library(readxl)
-library(dplyr)
-library(lubridate)
-library(zoo)
 library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=13))
-library(mgcv)
-library(mvtnorm)
 library(egg)
+library(shellpipes)
+rpcall("figure_epidemic.Rout figure_epidemic.R cases.rda")
 
-datevec <- seq(as.Date("2021-11-28"), as.Date("2022-01-30"), by=1)
-weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
-
-nsample <- 1000
-
-cases <- vroom("data/COVID-19_aantallen_gemeente_per_dag.csv")
-variant <- read_xlsx("data/variant-netherlands.xlsx")
-
-cases_all <- cases %>%
-  group_by(Date_of_publication) %>%
-  summarize(
-    cases=sum(Total_reported)
-  ) %>%
-  arrange(Date_of_publication) %>%
-  mutate(
-    rollmean=rollmean(cases, 7, fill=NA)
-  ) %>%
-  filter(Date_of_publication >= "2021-11-22", Date_of_publication <= "2022-01-30")
-
-cases_all_weekly <- cases_all %>%
-  mutate(
-    group=cut(Date_of_publication, weekbreak+1)
-  ) %>%
-  group_by(group) %>%
-  summarize(
-    cases=sum(cases),
-    date=max(as.Date(Date_of_publication))
-  )
-
-variant2 <- data.frame(
-  omicron=unlist(variant[9,-c(1:2, 13)]),
-  delta=unlist(variant[5,-c(1:2, 13)]),
-  total=unlist(variant[1,-c(1:2, 13)]),
-  week=c(4, 3, 2, 1, 52, 51, 50, 49, 48, 47),
-  year=c(2022, 2022, 2022, 2022, 2021, 2021, 2021, 2021, 2021, 2021),
-  date=rev(cases_all_weekly$date)
-)
-
-variant3 <- variant2 %>%
-  group_by(year, week) %>%
-  mutate(
-    omicron_prop=omicron/total,
-    omicron_prop_lwr=binom.test(omicron, total)[[4]][1],
-    omicron_prop_upr=binom.test(omicron, total)[[4]][2],
-    delta_prop=delta/total,
[…]

### 2022-05-20 figure/figure_incubation.R

@@ -1,91 +1,10 @@
-library(dplyr)
 library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(egg)
-source("R/sample_incubation.R")
+library(shellpipes)
+rpcall("figure_incubation.Rout figure_incubation.R rdaout/calculate_incubation_r.rda")
 
-rvec_nsgtf <- seq(-0.1, 0, length.out=11)
-rvec_sgtf <- seq(0.1, 0.2, length.out=11)
-
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
-
-corrected_mean_nsgtf <- lapply(1:length(rvec_nsgtf), function(x) {
-  numer <- sapply(1:nsample, function(y) {
-    integrate(function(z) {
-      z * dweibull(z, shape=backward_shape_nsgtf_sample[y], scale=backward_scale_nsgtf_sample[y]) * exp(rvec_nsgtf[x] * z)
-    }, 0, Inf)[[1]]
-  })
-  
-  denom <- sapply(1:nsample, function(y) {
-    integrate(function(z) {
-      dweibull(z, shape=backward_shape_nsgtf_sample[y], scale=backward_scale_nsgtf_sample[y]) * exp(rvec_nsgtf[x] * z)
-    }, 0, Inf)[[1]]
-  })
-  
-  mean <- numer/denom
-  
-  data.frame(
-    r=rvec_nsgtf[x],
-    median=median(mean),
-    lwr=quantile(mean, c(0.025)),
-    upr=quantile(mean, c(0.975))
-  )
-}) %>%
-  bind_rows
-
-corrected_mean_sgtf <- lapply(1:length(rvec_sgtf), function(x) {
-  numer <- sapply(1:nsample, function(y) {
-    integrate(function(z) {
-      z * dweibull(z, shape=backward_shape_sgtf_sample[y], scale=backward_scale_sgtf_sample[y]) * exp(rvec_sgtf[x] * z)
-    }, 0, 1e3)[[1]]
-  })
-  
-  denom <- sapply(1:nsample, function(y) {
-    integrate(function(z) {
-      dweibull(z, shape=backward_shape_sgtf_sample[y], scale=backward_scale_sgtf_sample[y]) * exp(rvec_sgtf[x] * z)
-    }, 0, 1e3)[[1]]
-  })
-  
-  mean <- numer/denom
-  
-  data.frame(
-    r=rvec_sgtf[x],
[…]

### 2022-05-20 figure/figure_reproduction_advantage.R

@@ -1,136 +1,12 @@
-library(bbmle)
-library(vroom)
-library(readxl)
-library(tidyr)
 library(dplyr)
-library(lubridate)
-library(zoo)
 library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
-library(mgcv)
-library(mvtnorm)
 library(egg)
 
 library(shellpipes)
-load("rdaout/fit_lognormal_base_comb_within.rda")
+rpcall("figure_reproduction_advantage.Rout figure_reproduction_advantage.R advantage.rda")
 
-weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
-
-nsample <- 1000
-
-cases <- vroom("data/COVID-19_aantallen_gemeente_per_dag.csv")
-variant <- read_xlsx("data/variant-netherlands.xlsx")
-
-cases_all <- cases %>%
-  group_by(Date_of_publication) %>%
-  summarize(
-    cases=sum(Total_reported)
-  ) %>%
-  arrange(Date_of_publication) %>%
-  mutate(
-    rollmean=rollmean(cases, 7, fill=NA)
-  ) %>%
-  filter(Date_of_publication >= "2021-11-22", Date_of_publication <= "2022-01-30")
-
-cases_all_weekly <- cases_all %>%
-  mutate(
-    group=cut(Date_of_publication, weekbreak+1)
-  ) %>%
-  group_by(group) %>%
-  summarize(
-    cases=sum(cases),
-    date=max(as.Date(Date_of_publication))
-  )
-
-variant2 <- data.frame(
-  omicron=unlist(variant[9,-c(1:2, 13)]),
-  delta=unlist(variant[5,-c(1:2, 13)]),
-  total=unlist(variant[1,-c(1:2, 13)]),
-  week=c(4, 3, 2, 1, 52, 51, 50, 49, 48, 47),
-  year=c(2022, 2022, 2022, 2022, 2021, 2021, 2021, 2021, 2021, 2021),
-  date=rev(cases_all_weekly$date)
-)
-
-variant3 <- variant2 %>%
-  group_by(year, week) %>%
-  mutate(
-    omicron_prop=omicron/total,
-    omicron_prop_lwr=binom.test(omicron, total)[[4]][1],
[…]

### 2022-05-20 figure/figure_reproduction_advantage_between.R

@@ -1,138 +1,12 @@
-library(bbmle)
-library(vroom)
-library(readxl)
-library(tidyr)
 library(dplyr)
-library(lubridate)
-library(zoo)
-library(ggplot2); theme_set(theme_bw(base_family = "Times"))
-library(mgcv)
-library(mvtnorm)
+library(ggplot2); theme_set(theme_bw(base_family = "Times", base_size=14))
 library(egg)
 
 library(shellpipes)
-load("rdaout/fit_lognormal_base_comb_between.rda")
+rpcall("figure_reproduction_advantage_between.Rout figure_reproduction_advantage_between.R advantage_between.rda")
 
-weekbreak <- as.Date("2021-12-19") + 7 * (-4:6)
-
-nsample <- 1000
-
-cases <- vroom("data/COVID-19_aantallen_gemeente_per_dag.csv")
-variant <- read_xlsx("data/variant-netherlands.xlsx")
-
-cases_all <- cases %>%
-  group_by(Date_of_publication) %>%
-  summarize(
-    cases=sum(Total_reported)
-  ) %>%
-  arrange(Date_of_publication) %>%
-  mutate(
-    rollmean=rollmean(cases, 7, fill=NA)
-  ) %>%
-  filter(Date_of_publication >= "2021-11-22", Date_of_publication <= "2022-01-30")
-
-cases_all_weekly <- cases_all %>%
-  mutate(
-    group=cut(Date_of_publication, weekbreak+1)
-  ) %>%
-  group_by(group) %>%
-  summarize(
-    cases=sum(cases),
-    date=max(as.Date(Date_of_publication))
-  )
-
-variant2 <- data.frame(
-  omicron=unlist(variant[9,-c(1:2, 13)]),
-  delta=unlist(variant[5,-c(1:2, 13)]),
-  total=unlist(variant[1,-c(1:2, 13)]),
-  week=c(4, 3, 2, 1, 52, 51, 50, 49, 48, 47),
-  year=c(2022, 2022, 2022, 2022, 2021, 2021, 2021, 2021, 2021, 2021),
-  date=rev(cases_all_weekly$date)
-)
-
-variant3 <- variant2 %>%
-  group_by(year, week) %>%
-  mutate(
-    omicron_prop=omicron/total,
-    omicron_prop_lwr=binom.test(omicron, total)[[4]][1],
[…]

### 2022-05-20 scripts/calculate_incubation_mle.R

@@ -1,7 +1,5 @@
-source("../R/sample_incubation.R")
-
-r_nsgtf <- -0.05
-r_sgtf <- 0.15
+source("baseparam.R")
+source("sample_incubation.R")
 
 moment_0_nsgtf <- integrate(function(z) {
   dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r_nsgtf * z)

### 2022-05-20 scripts/calculate_incubation_r.R

@@ -1,0 +1,85 @@
+library(dplyr)
+source("baseparam.R")
+source("sample_incubation.R")
+
+corrected_mean_nsgtf <- lapply(1:length(rvec_nsgtf), function(x) {
+  numer <- sapply(1:nsample, function(y) {
+    integrate(function(z) {
+      z * dweibull(z, shape=backward_shape_nsgtf_sample[y], scale=backward_scale_nsgtf_sample[y]) * exp(rvec_nsgtf[x] * z)
+    }, 0, Inf)[[1]]
+  })
+  
+  denom <- sapply(1:nsample, function(y) {
+    integrate(function(z) {
+      dweibull(z, shape=backward_shape_nsgtf_sample[y], scale=backward_scale_nsgtf_sample[y]) * exp(rvec_nsgtf[x] * z)
+    }, 0, Inf)[[1]]
+  })
+  
+  mean <- numer/denom
+  
+  data.frame(
+    r=rvec_nsgtf[x],
+    median=median(mean),
+    lwr=quantile(mean, c(0.025)),
+    upr=quantile(mean, c(0.975))
+  )
+}) %>%
+  bind_rows
+
+corrected_mean_sgtf <- lapply(1:length(rvec_sgtf), function(x) {
+  numer <- sapply(1:nsample, function(y) {
+    integrate(function(z) {
+      z * dweibull(z, shape=backward_shape_sgtf_sample[y], scale=backward_scale_sgtf_sample[y]) * exp(rvec_sgtf[x] * z)
+    }, 0, 1e3)[[1]]
+  })
+  
+  denom <- sapply(1:nsample, function(y) {
+    integrate(function(z) {
+      dweibull(z, shape=backward_shape_sgtf_sample[y], scale=backward_scale_sgtf_sample[y]) * exp(rvec_sgtf[x] * z)
+    }, 0, 1e3)[[1]]
+  })
+  
+  mean <- numer/denom
+  
+  data.frame(
+    r=rvec_sgtf[x],
+    median=median(mean),
+    lwr=quantile(mean, c(0.025)),
+    upr=quantile(mean, c(0.975))
+  )
+}) %>%
+  bind_rows
+
+observed_nsgtf <- tibble(
+  time=seq(0, 14, by=0.1),
+  density=dweibull(time, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf)
+)
+
+observed_sgtf <- tibble(
+  time=seq(0, 14, by=0.1),
[…]

### 2022-05-20 scripts/fit_lognormal_r_50_within.R

@@ -1,0 +1,109 @@
+library(bbmle)
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("sample_incubation.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/fit_lognormal_base.rda")
+
+fit_lognormal_r_50_nsgtf_within <- lapply(rvec_nsgtf, function(r) {
+  print(r)
+  moment_0_nsgtf <- integrate(function(z) {
+    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_1_nsgtf <- integrate(function(z) {
+    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_2_nsgtf <- integrate(function(z) {
+    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
+  var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
+  
+  logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
+  logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
+  
+  ff <- fitfun_lognormal(data=data_50_nsgtf_within, 
+                         logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
+                         logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
+                         logmean_inc=logmean_inc_nsgtf, 
+                         logsd_inc=logsd_inc_nsgtf, 
+                         rho=rho, 
+                         r=r)
+  
+  mean <- exp(coef(ff)[[1]]+coef(ff)[[2]]^2/2)
+  
+  cc <- confint(ff, method="quad")
+  
+  tmp <- c(
+    exp(coef(ff)[[1]]) * exp(coef(ff)[[2]]^2/2),
[…]

### 2022-05-20 scripts/fit_lognormal_r_51_within.R

@@ -1,0 +1,113 @@
+library(bbmle)
+source("R/serialfun.R")
+source("R/fitfun.R")
+source("sample_incubation.R")
+source("serialdata.R")
+source("baseparam.R")
+load("rdaout/fit_lognormal_base.rda")
+
+fit_lognormal_r_51_nsgtf_within <- lapply(r_nsgtf, function(r) {
+  print(r)
+  moment_0_nsgtf <- integrate(function(z) {
+    dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_1_nsgtf <- integrate(function(z) {
+    z * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  moment_2_nsgtf <- integrate(function(z) {
+    z^2 * dweibull(z, shape=backward_shape_nsgtf, scale=backward_scale_nsgtf) * exp(r * z)
+  }, 0, 1000)[[1]]
+  
+  mean_inc_nsgtf <- moment_1_nsgtf/moment_0_nsgtf
+  var_inc_nsgtf <- moment_2_nsgtf/moment_0_nsgtf - mean_inc_nsgtf^2
+  
+  logsd_inc_nsgtf <- sqrt(log(var_inc_nsgtf/mean_inc_nsgtf^2 + 1))
+  logmean_inc_nsgtf <- log(mean_inc_nsgtf/exp(logsd_inc_nsgtf^2/2))
+  
+  ff <- fitfun_lognormal(data=data_51_nsgtf_within, 
+                         logmean_gen=coef(fit_lognormal_base_50_nsgtf_within)[[1]], 
+                         logsd_gen=coef(fit_lognormal_base_50_nsgtf_within)[[2]], 
+                   logmean_inc=logmean_inc_nsgtf, 
+                   logsd_inc=logsd_inc_nsgtf, 
+                   rho=rho, 
+                   r=r)
+  
+  mean <- exp(coef(ff)[[1]]+coef(ff)[[2]]^2/2)
+  
+  cc <- confint(ff, method="quad")
+  
+  tmp <- c(
+    exp(coef(ff)[[1]]) * exp(coef(ff)[[2]]^2/2),
[…]

### 2022-05-20 omicron-generation.tex
adding authors

@@
 }
+\newline
+\\
+Sang Woo Park\textsuperscript{1,*},
+Kaiyuan Sun\textsuperscript{2},
+Sam Abbott\textsuperscript{3},
+Ron Sender\textsuperscript{4},
+Yinon Bar-on\textsuperscript{4},
+Joshua Weitz\textsuperscript{5,6,7},
+Sebastian Funk\textsuperscript{3,8}, 
+Bryan T. Grenfell\textsuperscript{1,9},
+Jantien A Backer\textsuperscript{10},
+Jacco Wallinga\textsuperscript{10,11},
+Cecile Viboud\textsuperscript{2},
+Jonathan Dushoff\textsuperscript{12,13,14}
+\\
+\bigskip
+\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
+\\
+\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
+\\
+\textbf{3} Centre for the Mathematical Modelling of Infectious Diseases, London School of Hygiene \& Tropical Medicine, London, UK
+\\
+\textbf{4} Weizmann Institute of Science, Rehovot, Israel
+\\
+\textbf{5} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
+\\
+\textbf{6} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
+\\
+\textbf{7} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
+\\
+\textbf{8} Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, Keppel Street, London, UK
+\\
+\textbf{9} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
+\\
+\textbf{10} Center for Infectious Disease Control, WHO COVID-19 reference laboratory, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
+\\
+\textbf{11} Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
+\\
+\textbf{12} Department of Biology, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{13} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
+\\
+\textbf{14} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\\
+\bigskip
+
+*Corresponding author: swp2@princeton.edu
 \bigskip

### 2022-05-20 omicron-generation.tex
think i'm done

@@
 during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
-Here, we take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
+We take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
 We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.
@@
 instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
-We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making the uncertainty negligible.
+We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty negligible.
 
@@
 We calculate time-varying growth rates from each parameter set and use equi-tailed quantiles to generate 95\% confidence limits. 
-% \swp{Sam is calling this ``simultaneous confidence intervals'' but not sure what he means. Need to ask again.}
+We note that this method of calculating confidence intervals gives point-wise confidence intervals, meaning that the confidence intervals give 95\% coverage for the set of estimates at each time point;
+these intervals are narrower than simultaneous confidence intervals, which give 95\% coverage for the set of estimated curvatures across the time period \citep{simpson2017}.
 
@@
 While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50 (13--19 December 2021).
-We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant.
+We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant;
+this decrease in the mean generation interval is likely associated with the lockdown.
 
@@
 The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
-Accounting for differences in epidemic growth rates, however, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
+Accounting for differences in epidemic growth rates, however, we find similar incubation-period distributions for both variants but a shorter (by 0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
 Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January.
@@
 
-We primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
+There are several limitations to our analysis.
+First, we primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
 In doing so, we implicitly assume that the delay between infection and reports is fixed.
@@
 
-Finally, we do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
+We do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
 In general, epidemic growth rates likely differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
@@
 Therefore, while we estimate unclear differences in incubation-period distributions between Delta and Omicron infections, controlling for other covariates, such as age and immune history, may help better characterize differences in Delta and Omicron infections.
+
+Finally, there are several sources of biases in serial interval data that we did not consider.
+For example, the direction of transmission is particularly difficult to establish for SARS-CoV-2 due to pre-symptomatic transmission.
+Other sources of information, such as exposure history and positive test results, can help resolve uncertainties but are imperfect.
+Serial interval data also depend on the ability of infected individuals to accurately recall when their symptoms started.
+Future studies may explore how these biases affect the inference of generation intervals from serial intervals.
 

### 2022-05-22 omicron-generation.tex
done

@@
 However, the impact of epidemic dynamics is often neglected in estimating the timing of infection and transmission---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have been infected recently.
-Here, we re-analyze incubation-period and serial-interval data describing transmissions of the Delta and Omicron variants from Netherlands at the end of December 2021.
+Here, we re-analyze incubation-period and serial-interval data describing transmissions of the Delta and Omicron variants from the Netherlands at the end of December 2021.
 Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
 When we account for growth-rate differences of two variants during the study period, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
-The differences in realized generation intervals may be driven by the network effect---higher effective reproduction numbers of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (and therefore shortening realized generation intervals).
+The differences in realized generation intervals may be driven by the network effect---higher effective reproduction numbers of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (therefore shortening realized generation intervals).
 Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
@@
 To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
-We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data.
+We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets.
 For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
@@
 
-We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
-More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} package).
+We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
+More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
 We also assume normally distributed errors around the logged weekly cases.
@@
 
-We use our estimates of the generation-interval distributions to infer instantaneous reproduction numbers \Rt\ of the Delta and Omicron variant, as well as the ratio between two reproduction numbers.
+We use our estimates of the generation-interval distributions to infer instantaneous reproduction numbers \Rt\ of the Delta and Omicron variant, as well as the ratio between the two reproduction numbers.
 Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
@@
 Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
-The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022.
+The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022 (\fref{epidemic}D).
 The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
@@
 In both cases, the decrease in the reproduction advantage coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant;
-a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the observed growth-rate difference (\fref{epidemic}D).
+a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the observed growth-rate difference shown earlier (\fref{epidemic}D).
 
@@
 
-This study presents a simple method for accounting for dynamical biases in inferring incubation-period distributions based on epidemic growth rates.
+This study presents a simple method for accounting for dynamical biases when inferring incubation-period distributions based on epidemic growth rates.
 In practice, it is easier to measure the backward incubation-period distribution because we can directly observe symptom onset. 
@@
 In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
-More generally, the bias in inferring the reproduction advantage an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
+More generally, the bias in inferring the reproduction advantage of an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
 For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant ($> 4$ fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
@@
 Future studies may explore how these biases affect the inference of generation intervals from serial intervals.
+While comparisons of incubation-period and serial-interval distributions can shed insight on pathogen dynamics, both distributions typically do not account for the dynamics of asymptomatic infections; neglecting these differences can further bias estimates of transmissibility of a pathogen \citep{park2020time}.
 
@@
 Our study synthesizes a previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters.
-Similar efforts will be critical to improve estimates of the infectiousness profiles of future SARS-CoV-2 variants, especially among asymptomatically infected individuals \citep{park2020time}.
+Similar efforts will be critical to improve estimates of the infectiousness profiles of future SARS-CoV-2 variants, especially among asymptomatically infected individuals.
+These conclusions are relevant for other emerging and endemic pathogens in general.
 

### 2022-05-30 omicron-generation.tex
almost done
omicron

@@
 \\
-\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
+\textbf{2} Division of International Epidemiology and Population Studies, Fogarty International Center, 20 National Institutes of Health, Bethesda, MD, USA
 \\
@@
 \\
-\textbf{10} Center for Infectious Disease Control, WHO COVID-19 reference laboratory, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
+\textbf{10} Center for Infectious Disease Control, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
 \\
@@
 When we account for growth-rate differences of two variants during the study period, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
-The differences in realized generation intervals may be driven by the network effect---higher effective reproduction numbers of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (therefore shortening realized generation intervals).
+The differences in realized generation intervals may be driven by the network effect--- higher effective transmissibility of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (therefore shortening realized generation intervals).
 Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
@@
 Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
-The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
+The study found shorter mean within-household serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
 during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
@@
 To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
-We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets.
+We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets (typically around 2 days).
 For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
 instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
-We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty negligible.
+We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty due to sample size negligible.
 
@@
 For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
-We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation.
+We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation---this is also known as the method of moments.
 Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
@@
 Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
-Growth rates are estimated by taking the derivative of the generalized additive model estimates.
+Growth rates are estimated by taking the derivative of the generalized additive model estimates of logged number of cases.
 \label{fig:epidemic}
@@
 Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
-We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
+We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
 Previous simulations showed that network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
@@
 Other factors, such as more stringent contact tracing measures against the Omicron variant in the Netherlands \citep{backer2021omicron} faster within-host clearance of the Omicron variant \citep{hay2022viral}, and viral kinetics of reinfection and breakthrough infections, also likely contributed to shortening of generation intervals.
+If shorter generation intervals of the Omicron variant represents an increased proportion of presymptomatic transmission, control measures that target symptomatic individuals can become less effective.
 
@@
 We do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
-In general, epidemic growth rates likely differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
+In general, epidemic growth rates may differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
 We are not able to perform stratified analyses because individiaul-level information was not publicly available.
@@
 These conclusions are relevant for other emerging and endemic pathogens in general.
+
+\section*{Data availability}
+
+All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/omicron-generation}).
+
+\section*{Acknowledgements}
+
+We thank Ron Milo for providing helpful comments on the manuscript.
+This project has received funding from European Union's Horizon 2020 research and innovation programme - project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (JW and JB, grant agreement number 101003688).
+The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S.
+Department of Health and Human Services.
 

### 2022-06-04 omicron-generation.tex
update

@@
 \\
-\textbf{4} Weizmann Institute of Science, Rehovot, Israel
+\textbf{4} Department of Plant and Environmental Sciences, Weizmann Institute of Science, Rehovot, Israel
 \\
@@
 
-Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
+Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the B.1.617.2 (Delta) and B.1.1.529 (Omicron) variants from the Netherlands between 13 and 26 December 2021.
 The study found shorter mean within-household serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
-during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
+during this period, the incidence of Omicron cases were increasing, whereas the incidence of Delta cases were decreasing.
 We take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
@@
 For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
-Incubation period data are not publicly available with the original article; instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
+Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron 165 and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
+however, the data are not publicly available with the original article. 
+Instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
 
@@
 In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
-To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
+To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases infected with each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
 We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets (typically around 2 days).
@@
 instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
-We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty due to sample size negligible.
+We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty due to sample size negligible;
+we note that the estimates of the proportion of each variant can be still sensitive to non-randomness of the sampling.
 
 We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
-More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
+More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
 We also assume normally distributed errors around the logged weekly cases.
 In principle, using negative-binomial likelihoods with a log link function is more appropriate for modeling count data, especially when data are overdispersed \citep{ver2007quasi};
-here, we use Gaussian likelihoods on logged cases because the inferred numbers of cases caused by each variant are not counts.
-Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.
+here, we use Gaussian likelihoods on logged cases because the inferred numbers of cases infected with each variant are not counts.
+Finally, we take the derivative of the predicted logged numbers of cases infected with each variant to obtain time-varying growth rate estimates.
 
@@
 \begin{equation}
-b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx}.
+b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx},
 \label{eq:backward}
 \end{equation}
+where the denominator is a normalization constant so that $b_I(\tau)$ integrates to 1.
 Therefore, the backward incubation-period distribution $b_I(\tau)$ gives a biased estimate of the corresponding forward distribution $f_I(\tau)$.
@@
 \textbf{Schematic diagrams of serial and generation intervals from symptom- and infection-based infector cohorts.}
-(A) Serial intervals are typically measured from the cohort of infectors who develop symptoms at the same time. 
+(A) Forward serial intervals are measured from the cohort of infectors who develop symptoms at the same time. 
 In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
@@
 
-When an epidemic is growing exponentially, there are two opposing effects affecting the relationship between the mean serial and generation interval.
-First, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
-Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}}] > \mathbb E[\tau_{g, \mathrm{symp}}]$.
-Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
+In summary, when an epidemic is growing exponentially, there are two opposing effects affecting the relationship between the mean forward serial and generation interval.
+First, infectors who developed symptoms at the same time are more likely to have shorter (backward) incubation periods than the corresponding forward incubation periods of their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
+Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}}] > \mathbb E[\tau_{g, \mathrm{symp}}]$ due to correlations between incubation periods and generation intervals.
+Therefore, the difference between the mean forward serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean forward serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
 Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed the authors to conclude that the mean generation and serial intervals are identical.
@@
 Under this parameterization, for a given length of a serial interval $\tau$, we can rewrite the incubation period of the infector as $-\alpha_1$; the generation interval as $\alpha_2 - \alpha_1$; and the incubation period of the infectee as $\tau - \alpha_2$.
-Then, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be expressed in terms of three distributions (\eref{serial}): the backward incubation-period distribution of the infector $b_I(-\alpha_1)$, the forward generation-interval distribution conditional on a known value of the incubation period $x$, $f_{G|I}(\alpha_2-\alpha_1|-\alpha_1)$, and the forward incubation-period distribution of the infectee $f_I(\tau - \alpha_2)$.
+Then, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be expressed in terms of three distributions (\eref{serial}): the backward incubation-period distribution of the infector $b_I(-\alpha_1)$ (taken from \eref{backward}), the forward generation-interval distribution conditional on a known value of the incubation period $x$, $f_{G|I}(\alpha_2-\alpha_1|-\alpha_1)$, and the forward incubation-period distribution of the infectee $f_I(\tau - \alpha_2)$.
 Integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ and rewriting the backward incubation-period distribution $b_I(-\alpha_1)$ in terms of the forward distribution, we obtain \citep{park2021forward}:
@@
 For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
-We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation---this is also known as the method of moments.
+We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation (also known as the method of moments); we note that we are unable to directly fit a lognormal distribution to the forward incubation-period distribution because we are relying on existing estimates rather than raw data.
 Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
@@
 Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
-The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022 (\fref{epidemic}D).
-The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
+The number of COVID-19 cases infected with the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022 (\fref{epidemic}D).
+The number of COVID-19 cases infected with the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
 These changes in growth rates coincide with the introduction of lockdown on 19 December 2021 \citep{netherlandslockdown} and its relaxation beginning 15 January 2022 \citep{netherlandsrelax,netherlandsopen}.
@@
 (B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
-(C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
+(C) Weekly numbers of COVID-19 cases infected with the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
 Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
@@
 In particular, the generation-interval estimates of the Omicron variant are more sensitive to the assumed values of $\rho$ due to faster changes in incidence of infection---for example, changing $\rho$ from 0.85 to 0.5 changes the mean generation-interval estimates for the Omicron variant from 3.1 days (95\% CI: 2.8--3.3 days) to 2.7 days (95\% CI: 2.5--2.9 days).
+We cover a wide range of $\rho$ to consider the possibility that our original assumpton ($\rho=0.75$) may under or over-estimate the true $\rho$.
 

### 2022-06-04 figure/figure_epidemic.R

@@ -6,8 +6,8 @@ sourceFiles()
 loadEnvironments()
 
 g1 <- ggplot(cases_all) +
-  geom_rect(xmin=as.Date("2021-12-13"), xmax=as.Date("2021-12-19"), ymin=-Inf, ymax=Inf, alpha=0.1, fill="gray") +
-  geom_rect(xmin=as.Date("2021-12-20"), xmax=as.Date("2021-12-26"), ymin=-Inf, ymax=Inf, alpha=0.02, fill=4) +
+  annotate("rect", xmin=as.Date("2021-12-13"), xmax=as.Date("2021-12-19"), ymin=0, ymax=Inf, alpha=0.5, fill="gray") +
+  annotate("rect", xmin=as.Date("2021-12-20"), xmax=as.Date("2021-12-26"), ymin=0, ymax=Inf, alpha=0.5, fill=4) +
   geom_point(aes(Date_of_publication, cases)) +
   geom_line(aes(Date_of_publication, rollmean), lty=1) +
   annotate("text", x=as.Date("2021-12-15")+1, y=75000, label=c("Week\n50"), family="Times", size=3) +
@@ -25,6 +25,8 @@ g1 <- ggplot(cases_all) +
   )
 
 g2 <- ggplot(variant3) +
+  annotate("rect", xmin=as.Date("2021-12-13"), xmax=as.Date("2021-12-19"), ymin=-Inf, ymax=Inf, alpha=0.5, fill="gray") +
+  annotate("rect", xmin=as.Date("2021-12-20"), xmax=as.Date("2021-12-26"), ymin=-Inf, ymax=Inf, alpha=0.5, fill=4) +
   geom_point(aes(date, omicron_prop, col="Omicron", shape="Omicron")) +
   geom_errorbar(aes(date, ymin=omicron_prop_lwr, ymax=omicron_prop_upr, col="Omicron"), width=0) +
   geom_point(aes(date, delta_prop, col="Delta", shape="Delta")) +
@@ -43,6 +45,8 @@ g2 <- ggplot(variant3) +
   )
 
 g3 <- ggplot(variant4) +
+  annotate("rect", xmin=as.Date("2021-12-13"), xmax=as.Date("2021-12-19"), ymin=0, ymax=Inf, alpha=0.5, fill="gray") +
[…]

### 2022-06-07 omicron-generation.tex
modified:   omicron-generation.tex

@@
 When we account for growth-rate differences of two variants during the study period, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
-The differences in realized generation intervals may be driven by the network effect--- higher effective transmissibility of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (therefore shortening realized generation intervals).
+We further note that the differences in estimated generation intervals may be driven by the ``network effect''---higher effective transmissibility of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (therefore shortening realized generation intervals).
 Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
@@
 Transmission advantages of invading variants are typically characterized by the ratios of reproduction numbers, $\Rinv/\Rres$, and the differences in growth rates, $\rinv-\rres$.
-These two quantities are linked by the generation-interval distributions of the resident and invading variants.
+These quantities are linked by the generation-interval distributions of the resident and invading variants.
 For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical ($\Rinv = \Rres > 1$).
@@
 Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
-Each of these transmission-interval distributions are subject to dynamical effects, which can cause transmission-interval distributions to systematically differ from the corresponding generation-interval distribution.
-For example, when the epidemic is growing, there will be more recent infections. 
-Therefore, we are more likely to observe recently infected individuals among a cohort of infectors who developed symptoms at the same time---we refer to this effect as the ``dynamical bias''.
+Each of these transmission-interval distributions can be subject to dynamical effects, which can cause transmission-interval distributions to systematically differ from the corresponding generation-interval distribution.
+
+For example, when the epidemic is growing, there will be more recent infections, and we are therefore more likely to observe recently infected individuals among a cohort of infectors who developed symptoms at the same time.
 In this case, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
-Therefore, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions, especially if their growth rates differ.
+We refer to such effects of growth rate on expected intervals as ``dynamical bias''
+Because of dynamical bias, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions when their growth rates differ.
 
@@
 For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
-Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron 165 and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
+Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron \jd{What?: 165} and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
 however, the data are not publicly available with the original article. 
 Instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
+\jd{Is this weird now that Backer is an author? I wouldn't like it if I were a reviewer.}
 
@@
 In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
-To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases infected with each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
-We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets (typically around 2 days).
-For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
+\jd{I understand why we show the differences, but I think we're doing calculations with the separate growth rates, not the differences and should talk less about differences here.}
+To estimate the growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases infected with each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
+We note that the proportion of Delta and Omicron variants detected is reported with the date of sampling, whereas the case data are reported with the date of report, meaning that there is some delay between the two data sets (typically around 2 days).
+For simplicity, we do not account for this delay in our growth-rate estimates; 
 instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
-We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty due to sample size negligible;
-we note that the estimates of the proportion of each variant can be still sensitive to non-randomness of the sampling.
+We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making uncertainty due to sample size small;
+we note however that this estimate is also sensitive to  the assumption that sampling is random.
 
 We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
-More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
-We also assume normally distributed errors around the logged weekly cases.
-In principle, using negative-binomial likelihoods with a log link function is more appropriate for modeling count data, especially when data are overdispersed \citep{ver2007quasi};
-here, we use Gaussian likelihoods on logged cases because the inferred numbers of cases infected with each variant are not counts.
+More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a penalized cubic spline fitted with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
+\jd{I am not sure if it's good to have the s() statement without the rest of the gam() statement. Should we leave it out?}
+We use Gaussian likelihoods to fit to logged cases in part for convenience, and in part because the inferred numbers of cases infected with each variant are not whole numbers.
+In principle, it might be preferable to explicitly model the process of sampling for strain testing, and then using negative-binomial likelihoods for case numbers \citep{ver2007quasi}, but our main purpose here is simply to roughly estimate growth rates with reasonable uncertainties.
 Finally, we take the derivative of the predicted logged numbers of cases infected with each variant to obtain time-varying growth rate estimates.
@@
 We note that this method of calculating confidence intervals gives point-wise confidence intervals, meaning that the confidence intervals give 95\% coverage for the set of estimates at each time point;
-these intervals are narrower than simultaneous confidence intervals, which give 95\% coverage for the set of estimated curvatures across the time period \citep{simpson2017}.
+these intervals are narrower than simultaneous confidence intervals, which give 95\% coverage for the set of estimated time series across the whole time period \citep{simpson2017}.
 
@@
 The incubation-period distributions from 513 individuals (consisting of 258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022, were previously reported in \cite{backer2021omicron}.
-We note that these data sets cover a wider range of periods and therefore are different from the serial-interval data.
+These data cover a wider time period than the serial-interval data.
 \cite{backer2021omicron} used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates.
 In particular, the methods of \cite{backer2020incubation} assume that the infection time is uniformly distributed across exposure dates and compares the inferred infection time to a known symptom-onset time to calculate the incubation period for each individual.
-While this method may accurately estimate the infection time, and therefore the incubation period, of each individual, dynamical biases can still affect the distribution of the inferred incubation periods from this cohort.
+Even if this method can accurately estimate the infection time, and therefore the incubation period, of each individual, dynamical biases can still affect this sort of cohort-based estimation of incubation period.
 
 More specifically, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward \citep{park2021forward}.
-The forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
-We expect this forward incubation-period distribution $f_I(\tau)$ to remain relatively constant over the course of an epidemic of one given variant, although biases can arise in \emph{observing} incubation periods, based on public or medical awareness of the disease.
+Forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
+We expect the forward incubation-period distribution $f_I(\tau)$ to remain relatively constant over the course of an epidemic of one given variant, although biases can arise in \emph{observing} incubation periods, based on public or medical awareness of the disease.
 Backward incubation periods are measured from a cohort of individuals who developed symptoms at the same time. 
@@
 
+\jd{See above about whether we have the data or not.}
 We also model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
@@
 (B) Generation intervals for the cohort of infectors who are infected at the same time will have unbiased incubation periods.
+\jd{Awkward; in what sense do GIs ``have'' incubation periods?
+maybe: are not biased by dynamical effects on incubation periods?}
 \label{fig:diagram}
@@
 
-The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased (compared to the generation-interval distribution for an infection-based cohort) because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
+The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased (compared to the generation-interval distribution for an infection-based cohort) because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and are therefore likely to transmit earlier (\fref{diagram}A).
 This generation-interval distribution for a symptom-based cohort depends on the backward incubation-period distribution:
@@
 
-Assuming that the incidence of infection is changing exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged across $t$ \citep{park2021forward}.
+Assuming that the incidence of infection is changing exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged through time \citep{park2021forward}.
+\jd{Not really. It changes when susceptible numbers decrease rapidly. Can we just say ``relatively constant''? Should we explain about susceptibles?}
 Then, we can focus on the forward serial-interval distribution at $t=0$, which in turn allows us to reparameterize the incubation-period and generation-interval distributions in terms of the infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$.
@@
 As discussed earlier, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
-As we show later in the results section, the exponential growth rate changes over the study period, including weeks 50 and 51 (13--26 December 2021);
-for illustrative purpose, we choose values of $r$ that represent the dynamics of Delta and Omicron infections during this period and repeat the analysis across plausible ranges of $r$ (discussed later in detail).
+As we show in Results, the exponential growth rate changes over the study period, including weeks 50 and 51 (13--26 December 2021);
+for illustrative purposes, we choose representative values of $r$ that for Delta and Omicron during this period and also explore across plausible ranges of $r$ (see below).
 
@@
 We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
-The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient on a log scale $\rho$.
-Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
+The joint distribution is parameterized by log-scale means, $\mu_I$ and $\mu_G$, log-scale variances, $\sigma_I^2$ and $\sigma_G^2$, and the log-scale correlation coefficient $\rho$.
+\jd{Changed because log mean sounds like the log of the mean, which could be misleading; people often say mean log, but then the rest of the sentence gets awkward.}
+Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log-scale mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log-scale variance of $\sigma_G^2 (1 - \rho^2)$.
 For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
@@
 We then calculate the mean forward generation interval using \eref{meangen}.
-The 95\% confidence intervals are calculated by taking the estimated variance-covariance matrix for the log-mean and -standard deviation parameters of the log normal distributions and calculating the corresponding variance-covariance for the overall mean using Taylor expansion---this method is also known as the Delta method \citep{oehlert1992note}.
+The 95\% confidence intervals are calculated by taking the estimated variance-covariance matrix for our mean and standard deviation parameters and calculating the corresponding variance-covariance for the overall mean using Taylor expansion---this method is also known as the Delta method \citep{oehlert1992note}.
 We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.
@@
 Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
-% \jd{We should think about this, too -- is it really just the forward if it's \Rt?}
-% \swp{Not sure what you mean... during exponential growth, forward = intrinsic...}
+\jd{We should think about this, too -- is it really just the forward if it's \Rt?}
+\swp{Not sure what you mean... during exponential growth, forward = intrinsic...}
+\jd{I mean, should we just say forward instead of intrinsic? And then skip the ``approximate'' just below?}
 \begin{equation}

### 2022-06-07 omicron-generation.tex
modified:   omicron-generation.tex

@@
 \caption{
-\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
+\textbf{Epidemic dynamics of the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
 (A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
 The solid line represents the 7-day moving average.
-Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
-(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
+Data are publicly available at \url{https://data.rivm.nl/covid-19/}.
+(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available at \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
 (C) Weekly numbers of COVID-19 cases infected with the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
@@
 Previous analysis of a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022 found longer mean (backward) incubation period for the Delta variant than for the Omicron variant \citep{backer2021omicron} (\fref{incubation}A).
-However, these measurements were done during a period when the incidence of Omicron was increasing while the increasing of Delta was decreasing (\fref{epidemic}).
-% Thus, dynamical bias would be expected to lead to shorter observed (backward) incubation periods in Omicron, and longer observed incubation periods in Delta. 
-When we account for these growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions with a mean of 4.1 days (95\% CI: 3.8--4.4 days) for the Delta variant and 4.2 days (95\% CI: 3.6--4.9 days) for the Omicron variant \fref{incubation}B);
-in this case, the difference between the mean backward and forward incubation periods correspond to $-22\%$ and $7\%$ bias for the Omicron and Delta variants, respectively.
+However, when we account for growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions with a mean of 4.1 days (95\% CI: 3.8--4.4 days) for the Delta variant and 4.2 days (95\% CI: 3.6--4.9 days) for the Omicron variant \fref{incubation}B).
+In this case, the difference between the mean backward and forward incubation periods correspond to $-22\%$ and $7\%$ bias for the Omicron and Delta variants, respectively.
 Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates (\fref{incubation}C--D).
-For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0/\mathrm{days}$ (\fref{incubation}C),
+For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0.0/\mathrm{days}$ (\fref{incubation}C),
 while the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$ (\fref{incubation}D).
-We estimate wider uncertanties for the Omicron variant because higher absolute growth rate values lead to larger differences between the backward and forward incubation periods.
+Wider confidence intervals for the Omicron variant are driven by greater uncertainties from the dynamical correction, which is larger for Omicron because of higher absolute growth rates.
 
@@
 
-Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
+We can use these estimates of the forward incubation-period distributions to estimate the forward generation-interval distributions.
 For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13--26 December, 2021).
@@
 3.0 days (95\% CI: 2.7--3.2 days) for the Omicron variant and 3.8 days (95\% CI: 3.7--4.0 days) for the Delta variant.
-Across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants, we estimate robust differences in their mean generation intervals (\fref{serial}C--D).
+Our estimate of this difference in these mean generation intervals is robust across plausible ranges of assumptions about the growth rates of the variants (\fref{serial}C--D).
 Assuming lower values of the correlation $\rho$ between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
 In particular, the generation-interval estimates of the Omicron variant are more sensitive to the assumed values of $\rho$ due to faster changes in incidence of infection---for example, changing $\rho$ from 0.85 to 0.5 changes the mean generation-interval estimates for the Omicron variant from 3.1 days (95\% CI: 2.8--3.3 days) to 2.7 days (95\% CI: 2.5--2.9 days).
-We cover a wide range of $\rho$ to consider the possibility that our original assumpton ($\rho=0.75$) may under or over-estimate the true $\rho$.
+We explore a wide range of $\rho$ to consider the possibility that our original assumption ($\rho=0.75$) may under or over-estimate the true $\rho$.
 
+\jd{How do we decide things are unclear? Overlapping CIs is generally too conservative.}
 Similar pictures arise for between-household transmission pairs, but the differences in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
@@
 In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (13--19 December 2021 and 20--26 December 2021).
-While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50 (13--19 December 2021).
+While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all strata, except for within-household transmission pairs during week 50 (13--19 December 2021).
 We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant;
@@
 
-This study presents a simple method for accounting for dynamical biases when inferring incubation-period distributions based on epidemic growth rates.
+This study presents a method for accounting for dynamical biases when inferring incubation-period distributions based on epidemic growth rates.
+\jd{I'm a bit worried about the ff.~s.
+Backward and forward intervals are measured at the same time, it's all about how they are collated (and what info is available.
+This relates a bit to the point about not having access to Backer's data.}
 In practice, it is easier to measure the backward incubation-period distribution because we can directly observe symptom onset. 
@@
 In addition, the ``intrinsic'' generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic (or ``unmitigated'') generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
-Nonetheless, our estimates of the realized generation-interval distributions better describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
+Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
 and can therefore be expected to improve estimates of effective reproduction numbers.
@@
 More generally, the bias in inferring the reproduction advantage of an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
-For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant ($> 4$ fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+\jd{Need to be a bit more explicit about how we ``match'' 4-fold while estimating 2.6.}
 
@@
 Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
-On the other hand, the growth-rate difference is easier to estimate in real time is also more directly relevant to short-term dynamics. 
-For example, when two strains have the same $\RR$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
+On the other hand, the growth-rate difference is easier to estimate in real time, and is also more directly relevant to short-term dynamics. 
+For example, when two strains have the same $\RR>1$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
+\jd{The ``replace'' part is not clear, and in fact ties back to the very question we're asking; which measure is more fundamental and thus likely to be conserved as dynamics progress?}
 Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.
@@
 During the study period, growth rates of both the Delta and Omicron variants changed slowly, and therefore our constant-growth-rate assumption provides a reasonable approximation for their dynamics across two weeks.
-However, such assumption might not be applicable when growth rates are changing rapidly (e.g., due to an introduction of stingent control measures) or if the sampling window is too wide.
+However, this assumption might be problematic when growth rates are changing rapidly (e.g., due to an introduction of stringent control measures) or if the sampling window is too wide.
 Extending our framework to account for time-varying growth rates is relatively simple when inferring the forward incubation-period distribution from the corresponding backward distribution---we can simply replace $r$ with $r(t)$ in \eref{forward} because the backward incubation-period distribution is a weighted average of the forward incubation-period distributions and the number of individuals in each cohort (i.e., individuals who were infected at the same time).
-However, such extensions are more complicated for linking generation- and serial-interval distributions because the forward serial-interval distribution also depends on the cohort reproduction number---for example, if a certain cohort of infectors had higher reproduction number (e.g., because they were infected before control measures were observed), we are more likely to observe transmission that started from this cohort (see \citep{park2021forward} for mathematical details).
-Assuming a constant growth rate allows us to simplify this complexity because the reproduction number also stays constant during the exponential phase.
+However, such extensions are more complicated for linking generation- and serial-interval distributions because the forward serial-interval distribution also depends on the cohort reproduction number---for example, if a certain cohort of infectors had higher reproduction number (e.g., because they were infected before control measures were observed), we are more likely to observe transmission from this cohort (see \citep{park2021forward} for more details).
+Assuming exponential growth allows us to avoid this complexity.
 Extending our framework to account for time-varying growth rates can provide more accurate tools for inferring epidemiological delay distributions.
@@
 We do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
-In general, epidemic growth rates may differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
-We are not able to perform stratified analyses because individiaul-level information was not publicly available.
+In general, epidemic growth rates may differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences can affect estimates of epidemiological delay distributions, including the incubation-period and generation-interval distributions.
+We are not able to perform stratified analyses because individual-level information was not publicly available.
 Therefore, while we estimate unclear differences in incubation-period distributions between Delta and Omicron infections, controlling for other covariates, such as age and immune history, may help better characterize differences in Delta and Omicron infections.
 
-Finally, there are several sources of biases in serial interval data that we did not consider.
+Finally, there are several sources of biases in serial-interval data that we did not consider.
 For example, the direction of transmission is particularly difficult to establish for SARS-CoV-2 due to pre-symptomatic transmission.

### 2022-06-07 omicron-generation.tex
oops!

@@
 
-Serial interval data are taken from \cite{backer2021omicron}.
+Serial-interval data are taken from \cite{backer2021omicron}.
 Infector-infectee pairs were identified through contact tracing, and their symptom onset dates were reported through a national surveillance database.
@@
 Finally, there are several sources of biases in serial-interval data that we did not consider.
-For example, the direction of transmission is particularly difficult to establish for SARS-CoV-2 due to pre-symptomatic transmission.
+For example, the direction of transmission is difficult to establish for SARS-CoV-2 due to pre-symptomatic transmission.
 Other sources of information, such as exposure history and positive test results, can help resolve uncertainties but are imperfect.
-Serial interval data also depend on the ability of infected individuals to accurately recall when their symptoms started.
+Serial-interval data also depend on the ability of infected individuals to accurately recall when their symptoms started.
 Future studies may explore how these biases affect the inference of generation intervals from serial intervals.

### 2022-06-07 omicron-generation.tex
commit to pull

--- 2022-06-07 omicron-generation.tex
@@
 For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
-Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron \jd{What?: 165} and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
+Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
 however, the data are not publicly available with the original article. 
@@
 
-In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
-\jd{I understand why we show the differences, but I think we're doing calculations with the separate growth rates, not the differences and should talk less about differences here.}
+In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, their growth rates---into account.
 To estimate the growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases infected with each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
@@
 We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
-More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a penalized cubic spline fitted with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
-\jd{I am not sure if it's good to have the s() statement without the rest of the gam() statement. Should we leave it out?}
+More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a penalized cubic spline fitted with restricted maximum likelihood (specified as \texttt{gam(log(cases)$\sim$s(time, bs="cs"), method="REML")} using the \texttt{MGCV} \texttt{R} package):
 We use Gaussian likelihoods to fit to logged cases in part for convenience, and in part because the inferred numbers of cases infected with each variant are not whole numbers.
@@
 In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
-(B) Generation intervals for the cohort of infectors who are infected at the same time will have unbiased incubation periods.
-\jd{Awkward; in what sense do GIs ``have'' incubation periods?
-maybe: are not biased by dynamical effects on incubation periods?}
+(B) Generation intervals for the cohort of infectors who are infected at the same time are not biased by dynamical effects on incubation periods.
 \label{fig:diagram}

### 2022-06-07 omicron-generation.tex
almost done

@@
 Yinon Bar-on\textsuperscript{4},
-Joshua Weitz\textsuperscript{5,6,7},
+Joshua S. Weitz\textsuperscript{5,6,7},
 Sebastian Funk\textsuperscript{3,8}, 
@@
 \\
-\textbf{2} Division of International Epidemiology and Population Studies, Fogarty International Center, 20 National Institutes of Health, Bethesda, MD, USA
+\textbf{2} Division of International Epidemiology and Population Studies, Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
 \\
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
-\jd{Need to be a bit more explicit about how we ``match'' 4-fold while estimating 2.6.}
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previoulsy assumed generation-interval dystributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
 
@@
 On the other hand, the growth-rate difference is easier to estimate in real time, and is also more directly relevant to short-term dynamics. 
-For example, when two strains have the same $\RR>1$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
-\jd{The ``replace'' part is not clear, and in fact ties back to the very question we're asking; which measure is more fundamental and thus likely to be conserved as dynamics progress?}
+For example, when two strains have the same $\RR>1$, the one with shorter generation intervals will grow faster and become dominant as long as $\RR>1$;
+however, when $\RR$ is reduced below 1 (either due to intervention or susceptible depletion), the one with longer generation interval will grow faster.
+These transmission advantages are captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
 Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.
@@
 
-We do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
+We do not account for individual-level heterogeneity, such as age, vaccination status, or previous exposure history.
 In general, epidemic growth rates may differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences can affect estimates of epidemiological delay distributions, including the incubation-period and generation-interval distributions.
@@
 We thank Ron Milo for providing helpful comments on the manuscript.
-This project has received funding from European Union's Horizon 2020 research and innovation programme - project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (JW and JB, grant agreement number 101003688).
+JSW acknowledges support from the Army Research Office W911NF1910384 and the Ile de France region via the Chaires Blaise Pascal program.
+JW and JB has received funding from European Union's Horizon 2020 research and innovation programme -- project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688).
 The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S.

### 2022-06-08 omicron-generation.tex
compromising!

@@
 Instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
-\jd{Is this weird now that Backer is an author? I wouldn't like it if I were a reviewer.}
 
@@
 
-\jd{See above about whether we have the data or not.}
 We also model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
@@
 
-Assuming that the incidence of infection is changing exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged through time \citep{park2021forward}.
-\jd{Not really. It changes when susceptible numbers decrease rapidly. Can we just say ``relatively constant''? Should we explain about susceptibles?}
+Assuming that the incidence of infection will continue to change exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged through time \citep{park2021forward}.
 Then, we can focus on the forward serial-interval distribution at $t=0$, which in turn allows us to reparameterize the incubation-period and generation-interval distributions in terms of the infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$.
@@
 The joint distribution is parameterized by log-scale means, $\mu_I$ and $\mu_G$, log-scale variances, $\sigma_I^2$ and $\sigma_G^2$, and the log-scale correlation coefficient $\rho$.
-\jd{Changed because log mean sounds like the log of the mean, which could be misleading; people often say mean log, but then the rest of the sentence gets awkward.}
 Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log-scale mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log-scale variance of $\sigma_G^2 (1 - \rho^2)$.
@@
 Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
-\jd{We should think about this, too -- is it really just the forward if it's \Rt?}
-\swp{Not sure what you mean... during exponential growth, forward = intrinsic...}
-\jd{I mean, should we just say forward instead of intrinsic? And then skip the ``approximate'' just below?}
 \begin{equation}
@@
 
-\jd{How do we decide things are unclear? Overlapping CIs is generally too conservative.}
 Similar pictures arise for between-household transmission pairs, but the differences in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
@@
 This study presents a method for accounting for dynamical biases when inferring incubation-period distributions based on epidemic growth rates.
-\jd{I'm a bit worried about the ff.~s.
-Backward and forward intervals are measured at the same time, it's all about how they are collated (and what info is available.
-This relates a bit to the point about not having access to Backer's data.}
-In practice, it is easier to measure the backward incubation-period distribution because we can directly observe symptom onset. 
-Therefore, the observed incubation-period distributions are generally expected to be biased, and similar kinds of corrections will be necessary to accurately estimate the incubation-period distribution.
+Observed incubation-period distributions based on symptom-based cohorts are generally expected to be biased, and similar kinds of corrections will be necessary to accurately estimate the incubation-period distribution.
 We note that making these kinds of corrections will also depend on data availability, model complexity, and other epidemiological covariates affecting incubation periods, such as vaccine statuses. 

### 2022-06-22 omicron-generation.tex
done i think

@@
 
-Estimating the differences in the incubation-period, serial-interval, and generation-interval distributions of SARS-CoV-2 variants is critical to understanding their transmission dynamics and outbreak control.
+Estimating the differences in the incubation-period, serial-interval, and generation-interval distributions of SARS-CoV-2 variants is critical to understanding their transmission and control.
 However, the impact of epidemic dynamics is often neglected in estimating the timing of infection and transmission---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have been infected recently.
@@
 Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
+
+\section*{Significance}
+
+Characterizing changes in the effective transmissibility over the course of infection---also known as the generation-interval distribution---is essential to comparing emerging SARS-CoV-2 variants.
+Recent studies suggest that individuals infected with the Omicron variant develop symptoms earlier and transmit faster than those infected with the Delta variant.
+However, these studies typically neglect the impact of population-level dynamics on their inference: 
+when an epidemic is growing rapidly, we are more likely to find individuals who were infected recently, which biases us toward observing earlier transmission events.
+When we account for this dynamical bias, we find that Omicron infections from the Netherlands at the end of December 2021 had similar incubation periods, but shorter generation intervals, compared to Delta infections from the same period.
+Shorter generation intervals of the Omicron variant may be attributed to its higher effective reproduction number, which can cause faster local susceptible depletion around the contact network.
 
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previoulsy assumed generation-interval dystributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previously assumed generation-interval distributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
 
@@
 JSW acknowledges support from the Army Research Office W911NF1910384 and the Ile de France region via the Chaires Blaise Pascal program.
-JW and JB has received funding from European Union's Horizon 2020 research and innovation programme -- project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688).
+JW and JB has received funding from European Union's Horizon 2020 research and innovation programme---project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688).
+S.F.
+was supported by the Wellcome Trust (grant no.
+210758/Z/18/Z).
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S.

### 2022-06-22 omicron-generation.tex
homing (xiangshan)

@@
 
-Characterizing changes in the effective transmissibility over the course of infection---also known as the generation-interval distribution---is essential to comparing emerging SARS-CoV-2 variants.
-Recent studies suggest that individuals infected with the Omicron variant develop symptoms earlier and transmit faster than those infected with the Delta variant.
-However, these studies typically neglect the impact of population-level dynamics on their inference: 
-when an epidemic is growing rapidly, we are more likely to find individuals who were infected recently, which biases us toward observing earlier transmission events.
-When we account for this dynamical bias, we find that Omicron infections from the Netherlands at the end of December 2021 had similar incubation periods, but shorter generation intervals, compared to Delta infections from the same period.
-Shorter generation intervals of the Omicron variant may be attributed to its higher effective reproduction number, which can cause faster local susceptible depletion around the contact network.
+Recent studies suggest that individuals infected with the Omicron variant develop symptoms earlier (shorter incubation period) and transmit faster (shorter generation interval) than those infected with the Delta variant.
+However, these studies typically neglect population-level effects: 
+when an epidemic is growing, a greater proportion of current cases were infected recently, biasing us toward observing faster transmission events.
+Accounting for this dynamical bias, we find that Omicron infections from the Netherlands at the end of December 2021 had similar incubation periods, but shorter generation intervals, compared to Delta infections from the same period.
+Shorter generation intervals of the Omicron variant might be due to its higher effective reproduction number, which can cause faster local susceptible depletion around the contact network.
 

### 2022-10-24 response.tex
need to write response

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\ensuremath{{\mathcal R}_{0}}\xspace}
\newcommand{\Rs}{\Rx{\mathrm{s}}}
\newcommand{\Rpool}{\ensuremath{{\mathcal R}_{\textrm{\tiny{pool}}}}\xspace}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript.

\revtext{Editor review}
Dear Mr. Park,

Thank you for submitting your manuscript Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2 for consideration by PNAS. We apologize for the delay in assessing your manuscript. After an evaluation of its suitability for PNAS, your manuscript was assigned to an editor with expertise in the subject area of your work, who examined the manuscript and identified 2 expert reviewers in the field to assess your work. These reviews, which appear below, were considered by the editor, who determined that the issues raised by the reviewers were of sufficient concern to merit rejection of the manuscript.

The reviewers were positive, however, about certain aspects of the work, an assessment with which the editor concurs; accordingly, we are willing to consider a resubmission that constructively addresses all of the concerns raised in the reviews. The revised manuscript must satisfy both the reviewers and the editor in order to be accepted. Please know, though, that we cannot guarantee acceptance because new criticisms could arise upon re-evaluation, and another round of resubmission is not allowed.

If you would like to submit a resubmission, please submit it as a new paper. In your cover letter and the Comments to Editorial Staff, please specify that the paper is a resubmission and include your original tracking number and a point-by-point response to the editor and reviewers' comments. Your resubmission will be given a new tracking number. If you have any questions, please contact our office.

Dear Colleagues:

Both reviewers, as well as I, like this report and the way it is written. Reviewer 2 recommended its acceptance. Reviewer 1 raised an issue you need to address, "WHY the omicron variant had such a short generation time in this setting...", and 
recommend
 the rejection of your report. I agree with this review about this issue.

Based on that reviewer's criticism, I am recommending what I call a soft rejection. Address this reviewer's criticism in detail and modify the manuscript accordingly. By "address," I am considering the possibility of 
your
 presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.

I look forward to reading your revised manuscript.

Best wishes for your well-being and that of all the people you care for whom you care,

Be well,

Bruce

\rev{Reviewer \#1}

\revtext{The paper commences with a very clear introduction to the problems in estimating the transmission advantage of a new variant - in particular, the role of dynamical bias. This is followed by a very reasonable proposal for how the problem may be addressed. Applying these methods to a re-analysis of data on within- and between- household spread of delta and omicron variants of SARS-CoV-2 suggests that while both had similar incubation-period distributions, the mean generation interval for omicron was shorter by 0.3-0.8 days than previously estimated (0.6 days). The original analysis had also reported a difference in mean incubation periods of 3.2 days for omicron versus 4.4 days for delta; the methods used in this paper suggest that both have a mean of 4.1 days. The consequences of discriminating between generation time intervals on the estimation of transmission advantage is shown in Fig 5, although it is not clear how this pertains to dynamical bias as both the original study and this one reported a difference between the variants in this respect.}

Dynamical biases pertain to the estimation of delay distributions

\revtext{The caveats of this study, and the problems of applying this method to other datasets, are very thoroughly outlined in the discussion.}

Thank you.

\revtext{Less attention is paid to the more important question of WHY the omicron variant had such a short generation time in this setting.}

\revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}

Not circular.

\revtext{Other factors "such as more stringent contact tracing measures against the Omicron variant in the Netherlands [4], faster within-host clearance of the Omicron variant [35], and viral kinetics of reinfection and breakthrough infections" are mentioned without sufficient explanation as to why they would yield such a result.}

Can provide more detail.

\revtext{The role of immune evasion - although mentioned in passing - has not been sufficiently accommodated either in the discussion or in the analysis itself, even though it presents itself as the most likely explanation for the success of omicron in almost every global setting.}

Yes and no.

\revtext{To address this properly, the authors will have to account for the different proportions of the population currently immune to infection by either variant and how the level of cross-immunity between them affects our observations of transmission advantage. The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}

\rev{Reviewer \#2}

\revtext{The significance statement is accurate. I find the paper definitely relevant for future epidemiological studies on COVID-19, and even important for some technical works. But neither the methods nor the results are likely to have a huge impact on the field, even less on other fields.}

\revtext{The manuscript is technically sound and very well written. The results are convincing and clearly explained in the manuscript, and the arguments presented make sense from an epidemiological perspective. This paper emphasizes once more the importance of accounting for the epidemic dynamic (or the differential dynamic between two variants, as in this case) when inferring fundamental epidemiologically relevant quantities that vary in time during the infection. I find it an important contribution to the literature, albeit a technical one.}

\end{document}

### 2022-11-02 response.tex
done editing?

@@
 
-\revtext{Editor review}
-Dear Mr. Park,
+\rev{Editor}
 
-Thank you for submitting your manuscript Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2 for consideration by PNAS.
-We apologize for the delay in assessing your manuscript.
-After an evaluation of its suitability for PNAS, your manuscript was assigned to an editor with expertise in the subject area of your work, who examined the manuscript and identified 2 expert reviewers in the field to assess your work.
-These reviews, which appear below, were considered by the editor, who determined that the issues raised by the reviewers were of sufficient concern to merit rejection of the manuscript.
+\revtext{Dear Colleagues:}
 
-The reviewers were positive, however, about certain aspects of the work, an assessment with which the editor concurs; accordingly, we are willing to consider a resubmission that constructively addresses all of the concerns raised in the reviews.
-The revised manuscript must satisfy both the reviewers and the editor in order to be accepted.
-Please know, though, that we cannot guarantee acceptance because new criticisms could arise upon re-evaluation, and another round of resubmission is not allowed.
-
-If you would like to submit a resubmission, please submit it as a new paper.
-In your cover letter and the Comments to Editorial Staff, please specify that the paper is a resubmission and include your original tracking number and a point-by-point response to the editor and reviewers' comments.
-Your resubmission will be given a new tracking number.
-If you have any questions, please contact our office.
-
-Dear Colleagues:
-
-Both reviewers, as well as I, like this report and the way it is written.
+\revtext{Both reviewers, as well as I, like this report and the way it is written.
 Reviewer 2 recommended its acceptance.
 Reviewer 1 raised an issue you need to address, "WHY the omicron variant had such a short generation time in this setting...", and
-recommend
- the rejection of your report. I agree with this review about this issue.
+recommend the rejection of your report. I agree with this review about this issue.}
 
-Based on that reviewer's criticism, I am recommending what I call a soft rejection.
+\revtext{Based on that reviewer's criticism, I am recommending what I call a soft rejection.
 Address this reviewer's criticism in detail and modify the manuscript accordingly.
-By "address," I am considering the possibility of
-your
- presenting a compelling argument for why this issue need not be considered in this report. Your response to the reviews and the revised manuscript will be sent to the second reviewer.
-
-I look forward to reading your revised manuscript.
-
-Best wishes for your well-being and that of all the people you care for whom you care,
-
-Be well,
-
-Bruce
+By "address," I am considering the possibility of your presenting a compelling argument for why this issue need not be considered in this report.
+Your response to the reviews and the revised manuscript will be sent to the second reviewer.}
 

### 2022-11-02 omicron-generation.tex
ack
done editing?

@@
 
-A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
+Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
 \cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
-Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
-We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
-Previous simulations showed that network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
+Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
+The realized generation intervals represent time between actual infection events and are different from the intrinsic generation intervals, which represent the changes in the infectiousness of an infected individuals over the course of their infections \citep{champredon2015intrinsic}.
+
+Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals of the Omicron variant.
+For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} can lead to faster recovery, which in turn can also shorten the intrinsic generation interval \citep{roberts2007model}.
+Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
+At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
+however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
+At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.
+
+Even if the Delta and Omicron variants were to have identical intrinsic generation-interval distributions, their realized generation-interval distributions can still differ.
+For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
+Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
+If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, can also lead to shorter realized generation intervals.
+However, this explanation is also less likely given similarities in the inferred incubation-period distributions.
+
+Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
+To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
+in this case, both variants have identical intrinsic generation-interval distributions.
+However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
+Previous simulations showed that such network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
 While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
-Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
-Other factors, such as more stringent contact tracing measures against the Omicron variant in the Netherlands \citep{backer2021omicron} faster within-host clearance of the Omicron variant \citep{hay2022viral}, and viral kinetics of reinfection and breakthrough infections, also likely contributed to shortening of generation intervals.
-If shorter generation intervals of the Omicron variant represents an increased proportion of presymptomatic transmission, control measures that target symptomatic individuals can become less effective.
 
-While our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
+Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-In addition, the ``intrinsic'' generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic (or ``unmitigated'') generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
+In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
 Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
@@
 We thank Ron Milo for providing helpful comments on the manuscript. 
+JD was supported by the Canadian Institutes of Health Research, the Natural Sciences and Engineering Research Council of Canada, and the Michael G. DeGroote Institute for Infectious Disease Research.
 JSW acknowledges support from the Army Research Office W911NF1910384 and the Ile de France region via the Chaires Blaise Pascal program. 
 JW and JB has received funding from European Union's Horizon 2020 research and innovation programme---project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688).
-S.F.
-was supported by the Wellcome Trust (grant no.
-210758/Z/18/Z).
+SF was supported by the Wellcome Trust (grant no. 210758/Z/18/Z). 
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript. 

### 2022-11-02 response.tex
response

@@
 
-Dynamical biases pertain to the estimation of delay distributions
+Dynamical biases pertain to the estimation of delay distributions, and not to the estimation of transmission advantage. 
+We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
+Here, what we show is that the estimates of transmission advantage depends on the assumptions about the generation-interval distributions. 
+For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:
+
+``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previously assumed generation-interval distributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.''
 
@@
 
+We have now added text to the discussion explaining different types of generation intervals (intrinsic vs realized) and why the realized generation interval of the Omicron variant might be shorter:
+
+``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
+\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
+Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
+The realized generation intervals represent time between actual infection events and are different from the intrinsic generation intervals, which represent the changes in the infectiousness of an infected individuals over the course of their infections \citep{champredon2015intrinsic}.
+
+Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals of the Omicron variant.
+For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} can lead to faster recovery, which in turn can also shorten the intrinsic generation interval \citep{roberts2007model}.
+Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
+At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
+however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
+At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.
+
+Even if the Delta and Omicron variants were to have identical intrinsic generation-interval distributions, their realized generation-interval distributions can still differ.
+For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
+Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
+If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, can also lead to shorter realized generation intervals.
+However, this explanation is also less likely given similarities in the inferred incubation-period distributions.
+
+Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
+To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
+in this case, both variants have identical intrinsic generation-interval distributions.
+However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
+Previous simulations showed that such network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
+While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
+
+Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
+In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
+In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
+Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
+and can therefore be expected to improve estimates of effective reproduction numbers.''
+
 \revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}
 
-Not circular.
+This explanation is not circular.
+As we explain in the Discussion, the success of the Omicron variant at the population-level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation interval.
+Instead, we now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:
+
+``To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
+in this case, both variants have identical intrinsic generation-interval distributions.
+However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.''
 
@@
 
-Can provide more detail.
+We now provide a more detailed explanation.
 
@@
 
-Yes and no.
+We now discuss the possible effects of immune evasion more thorougly.
+While we agree that the immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.
 
@@
 The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}
+
+As explained previously, the population-level success of the Omicron variant is almost a separate question from the estimation of their incubation-period and generation-interval distributions, which is the main goal of our analysis.
+We also now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:
+
+``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
+At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
+however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
+At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.''
 
@@
 
+Thank you for your review.
+
+\bibliography{omicron}
+
 \end{document}

### 2022-11-15 response.tex
Not really started yet (don't look)

@@
 We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
-Here, what we show is that the estimates of transmission advantage depends on the assumptions about the generation-interval distributions. 
+Here, what we show is that estimates of transmission advantage depend on assumptions about the generation-interval distributions. 
 For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:
@@
 ``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previously assumed generation-interval distributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.''
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to the assumptions and results of \cite{pearson2021bounding} for Gauteng province.''
 

### 2022-11-16 response.tex
Flatting

--- 2022-11-02 response.tex
@@
 
-Dynamical biases pertain to the estimation of delay distributions
+Dynamical biases pertain to the estimation of delay distributions, and not to the estimation of transmission advantage. 
+We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
+Here, what we show is that estimates of transmission advantage depend on assumptions about the generation-interval distributions. 
+For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:
+
+``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.''
 
@@
 
+We have now added text to the discussion explaining different types of generation intervals (intrinsic vs realized) and why the realized generation interval of the Omicron variant might be shorter:
+
+``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
+\jd{“tried to” is odd; can we drop it?}
+\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
+Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
+The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.
+
+Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
+For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
+\jd{Something about: A greater proportion of Omicron than Delta infections will be in people with pre-existing immunity (because these people are less likely to be infected by Delta).}
+Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
+At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
+\jd{Don't understand the following; we're rebutting a point that hasn't been clearly outlined. The whole ¶ I guess could be edited to combine points above and below this comment.}
+however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Delta variant.
+At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.
+
+Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
+For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, resulting in a shorter realized generation interval.
+Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
+If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
+However, this explanation is also less likely given similarities in the inferred incubation-period distributions.
+
+Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
+To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
+in this case, both variants have identical intrinsic generation-interval distributions.
+We note that this scenario does not match our model, which assumes constant force of infection, not a deterministic number of infections.
+However, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
+\jd{It bugs me how awkwardly this matches our actual model but I think it's effective communication. I added an awkward s.~above, which you can think about.}
+Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
+While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
+
+Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
+In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
+In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than current estimates, given existing levels of interventions, including vaccination, and pandemic awareness.
+Estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
+Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
+and can therefore be expected to improve estimates of effective reproduction numbers.''
+
 \revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}
 
-Not circular.
+We don't believe that this explanation is circular, and have tried to clarify our explication further.
+As we explain in the Discussion, the success of the Omicron variant at the population level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation interval.
+We now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:
+
+\jd{Reinsert this text after editing.}
 
@@
 
-Yes and no.
+We now discuss the possible effects of immune evasion more thorougly.
 
@@
 The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}
+
+While we agree that immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.
+
+We now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:
+
+``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
+At the population level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
+however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
+At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.''
 

### 2022-11-28 response.tex
done

@@
 Thank you for the chance to revise and resubmit our manuscript. 
+We now discuss potential mechanisms driving shorter generation intervals of the Omicron variant in detail.
+In doing so, we distinguish intrinsic generation intervals, which represent changes in infectiousness over the course of infection, from realized generation intervals, which represent time between actual transmission events---our estimates correspond to the latter quantity.
+We also discuss the uncertainties associated with possible effects of immunity on the generation-interval distributions of the Omicron and Delta variants.
+As the reviewer pointed out, it is true that immune evasiveness of the Omicron variant is the main explanation for the population-level success of Omicron---however, its success or immune evasion does not require or depend on changes in generation-interval distributions as we explain.
+Below please find our detailed responses to reviewers.
 
@@
 Your response to the reviews and the revised manuscript will be sent to the second reviewer.}
+
+Thank you very much.
+We have revised the manuscript to discuss the effect of immunity.
+In doing so, we have tried to distinguish its population-level effect (which allowed the Omicron variant to dominate) from the individual-level effect (changes in the generation-interval distribution).
+We note these two effects are separate: its success or immune evasion does not require or depend on changes in generation-interval distributions.
 
@@
 ``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
-\jd{“tried to” is odd; can we drop it?}
-\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
+\cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
 Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
@@
 For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
-\jd{Something about: A greater proportion of Omicron than Delta infections will be in people with pre-existing immunity (because these people are less likely to be infected by Delta).}
-Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
-At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
-\jd{Don't understand the following; we're rebutting a point that hasn't been clearly outlined. The whole ¶ I guess could be edited to combine points above and below this comment.}
-however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Delta variant.
-At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.
+Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
+In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
+this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
+However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
+some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
+We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
+Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
+Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.
 
 Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
-For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, resulting in a shorter realized generation interval.
+For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
 Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
@@
 Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
-To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
-in this case, both variants have identical intrinsic generation-interval distributions.
-We note that this scenario does not match our model, which assumes constant force of infection, not a deterministic number of infections.
-However, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
-\jd{It bugs me how awkwardly this matches our actual model but I think it's effective communication. I added an awkward s.~above, which you can think about.}
+To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
+We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simplest example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
+Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
 Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
@@
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than current estimates, given existing levels of interventions, including vaccination, and pandemic awareness.
-Estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
+In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
 Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
@@
 
-\jd{Reinsert this text after editing.}
+``Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
+To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
+We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simplest example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
+Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
+Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
+While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).''
 
@@
 While we agree that immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.
-
 We now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:
 
-``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
-At the population level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
-however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
-At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.''
+``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
+In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
+this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
+However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
+some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
+We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
+Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
+Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.''
 

### 2022-11-28 omicron-generation.tex
done

@@
 Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
-\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
+\cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
 Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
-The realized generation intervals represent time between actual infection events and are different from the intrinsic generation intervals, which represent the changes in the infectiousness of an infected individuals over the course of their infections \citep{champredon2015intrinsic}.
+The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.
 
-Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals of the Omicron variant.
-For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} can lead to faster recovery, which in turn can also shorten the intrinsic generation interval \citep{roberts2007model}.
-Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals), but the relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is more difficult to predict, especially given its high immune evasiveness.
-At the population-level, the immune evasiveness of the Omicron variant allowed it to rapidly replace the previously dominant Delta variant;
-however, we note that the population-level dominance of the Omicron variant does not require it to have shorter generation intervals than the Omicron variant.
-At the individual-level, some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-Therefore, we would expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections would be more strongly influenced by the intrinsic differences in their viral kinetics, rather than their differences in the ability to evade immunity.
+Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
+For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
+Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
+In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
+this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
+However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
+some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
+We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
+Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
+Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.
 
-Even if the Delta and Omicron variants were to have identical intrinsic generation-interval distributions, their realized generation-interval distributions can still differ.
+Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
 For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
 Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
-If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, can also lead to shorter realized generation intervals.
+If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
 However, this explanation is also less likely given similarities in the inferred incubation-period distributions.
 
-Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections.
-To better illustrate the network effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days;
-in this case, both variants have identical intrinsic generation-interval distributions.
-However, if both individuals were each living in a household consisting of four other people, an individual infected with variant A will infect one household member everyday; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
-Previous simulations showed that such network effects can have considerable impact on realized generation intervals even during the initial exponential growth phase, when susceptible depletion is negligible at the population level \citep{park2020inferring}.
+Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
+To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
+We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simplest example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
+Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
+Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
 While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
 
-Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but it is still uncertain how infectiousness profiles differ intrinsically between Omicron and Delta.
+Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
 In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the previously assumed generation-interval distributions and estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.
 

### 2022-11-28 response.tex
done

@@
 We also discuss the uncertainties associated with possible effects of immunity on the generation-interval distributions of the Omicron and Delta variants.
-As the reviewer pointed out, it is true that immune evasiveness of the Omicron variant is the main explanation for the population-level success of Omicron---however, its success or immune evasion does not require or depend on changes in generation-interval distributions as we explain.
+As the reviewer pointed out, it is true that immune evasiveness of the Omicron variant is the main explanation for the population-level success of Omicron---however, this mechanism does not require or depend on changes in generation-interval distributions as we explain.
+
 Below please find our detailed responses to reviewers.
@@
 some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
-We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
-Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
-Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
+We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
+immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.
 
-Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions, their realized generation-interval distributions could still differ.
+Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions (e.g., a lack of changes in viral kinetics or the effect of pre-existing immunity), their realized generation-interval distributions could still differ.
 For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
@@
 To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
-We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simplest example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
-Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, thereby having shorter realized generation intervals.
+We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
+Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
 Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
@@
 Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
-In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their true infectiousness profile may be smaller than the estimated differences in their realized generation-interval distributions.
-In addition, the intrinsic generation intervals of both Omicron and Delta variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
+In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their infectiousness profiles may be smaller than the estimated differences in their realized generation-interval distributions.
+In addition, the counter-factual initial intrinsic generation intervals (for an immunologically naive population) of both variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
 Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
@@
 We don't believe that this explanation is circular, and have tried to clarify our explication further.
-As we explain in the Discussion, the success of the Omicron variant at the population level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation interval.
+As we explain in the Discussion, the success of the Omicron variant at the population level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation intervals.
 We now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:
@@
 some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasiveness may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
-We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, are separate from possible changes in the generation-interval distribution that we have outlined above.
-Even if Delta and Omicron variants had identical intrinsic generation-interval distributions, immune evasiveness alone would allow the Omicron variant to become dominant.
-Therefore, we might expect the differences in the intrinsic generation-interval distributions of the Delta and Omicron infections to be more strongly influenced by the intrinsic differences in their viral kinetics.''
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
+We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
+immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.''
 

### 2022-11-30 response.tex
update
edits from various ppl

@@
 
-Thank you for the chance to revise and resubmit our manuscript. 
+Thank you for the chance to revise and resubmit our manuscript (previously 2022-12304). 
 We now discuss potential mechanisms driving shorter generation intervals of the Omicron variant in detail.
@@
 ``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.''
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province in South Africa.''
 
@@
 \cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
-Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
+Here, we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
 The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.
@@
 some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them (relative to reinfection with Delta, which would be less immune evasive).
+Their combined effects on the resulting generation-interval distribution are unclear.
 We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
@@
 For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
-Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
+Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have a small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
 If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
@@
 To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
-We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
-Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
+We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect.
+Under the same scenario, if each individual closely interacted with four other contacts, the individual infected with variant A will infect one person every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
 Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.

### 2022-11-30 omicron-generation.tex
edits from various ppl

@@
 some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
-We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
+In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them (relative to reinfection with Delta, which would be less immune evasive).
+Their combined effects on the resulting generation-interval distribution are unclear.
+We further note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
 immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.
@@
 For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.
+With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province in South Africa.
 

### 2022-12-06 omicron-generation.tex
update author

@@
 Joshua S. Weitz\textsuperscript{5,6,7},
-Sebastian Funk\textsuperscript{3,8}, 
-Bryan T. Grenfell\textsuperscript{1,9},
-Jantien A Backer\textsuperscript{10},
-Jacco Wallinga\textsuperscript{10,11},
+Sebastian Funk\textsuperscript{3}, 
+Bryan T. Grenfell\textsuperscript{1,8},
+Jantien A Backer\textsuperscript{9},
+Jacco Wallinga\textsuperscript{9,10},
 Cecile Viboud\textsuperscript{2},
-Jonathan Dushoff\textsuperscript{12,13,14}
+Jonathan Dushoff\textsuperscript{11,12,13}
 \\
@@
 \\
-\textbf{8} Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, Keppel Street, London, UK
+\textbf{8} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
 \\
-\textbf{9} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
+\textbf{9} Center for Infectious Disease Control, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
 \\
-\textbf{10} Center for Infectious Disease Control, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
+\textbf{10} Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
 \\
-\textbf{11} Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
+\textbf{11} Department of Biology, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{12} Department of Biology, McMaster University, Hamilton, ON, Canada
+\textbf{12} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
 \\
-\textbf{13} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
-\\
-\textbf{14} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
+\textbf{13} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
 \\

### 2022-12-26 omicron-generation.tex
forgot to submit...

--- 2022-11-30 omicron-generation.tex
@@
 Kaiyuan Sun\textsuperscript{2},
-Sam Abbott\textsuperscript{3},
-Ron Sender\textsuperscript{4},
-Yinon Bar-on\textsuperscript{4},
-Joshua S. Weitz\textsuperscript{5,6,7},
-Sebastian Funk\textsuperscript{3,8}, 
+Sam Abbott\textsuperscript{3,4},
+Ron Sender\textsuperscript{5},
+Yinon Bar-on\textsuperscript{5},
+Joshua S. Weitz\textsuperscript{6,7,8},
+Sebastian Funk\textsuperscript{3,4}, 
 Bryan T. Grenfell\textsuperscript{1,9},
@@
 \\
-\textbf{4} Department of Plant and Environmental Sciences, Weizmann Institute of Science, Rehovot, Israel
+\textbf{4} Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, Keppel Street, London, UK
 \\
-\textbf{5} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
+\textbf{5} Department of Plant and Environmental Sciences, Weizmann Institute of Science, Rehovot, Israel
 \\
-\textbf{6} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
+\textbf{6} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
 \\
-\textbf{7} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
+\textbf{7} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
 \\
-\textbf{8} Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, Keppel Street, London, UK
+\textbf{8} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
 \\

### 2023-03-28 response.tex
making changes

--- 2022-11-28 response.tex
@@
 
-\noindent Dear Editor:
-
-Thank you for the chance to revise and resubmit our manuscript. 
-We now discuss potential mechanisms driving shorter generation intervals of the Omicron variant in detail.
-In doing so, we distinguish intrinsic generation intervals, which represent changes in infectiousness over the course of infection, from realized generation intervals, which represent time between actual transmission events---our estimates correspond to the latter quantity.
-We also discuss the uncertainties associated with possible effects of immunity on the generation-interval distributions of the Omicron and Delta variants.
-As the reviewer pointed out, it is true that immune evasiveness of the Omicron variant is the main explanation for the population-level success of Omicron---however, this mechanism does not require or depend on changes in generation-interval distributions as we explain.
-
-Below please find our detailed responses to reviewers.
-
-\rev{Editor}
-
-\revtext{Dear Colleagues:}
-
-\revtext{Both reviewers, as well as I, like this report and the way it is written.
-Reviewer 2 recommended its acceptance.
-Reviewer 1 raised an issue you need to address, "WHY the omicron variant had such a short generation time in this setting...", and
-recommend the rejection of your report. I agree with this review about this issue.}
-
-\revtext{Based on that reviewer's criticism, I am recommending what I call a soft rejection.
-Address this reviewer's criticism in detail and modify the manuscript accordingly.
-By "address," I am considering the possibility of your presenting a compelling argument for why this issue need not be considered in this report.
-Your response to the reviews and the revised manuscript will be sent to the second reviewer.}
-
-Thank you very much.
-We have revised the manuscript to discuss the effect of immunity.
-In doing so, we have tried to distinguish its population-level effect (which allowed the Omicron variant to dominate) from the individual-level effect (changes in the generation-interval distribution).
-We note these two effects are separate: its success or immune evasion does not require or depend on changes in generation-interval distributions.
-
 \rev{Reviewer \#1}
 
-\revtext{The paper commences with a very clear introduction to the problems in estimating the transmission advantage of a new variant - in particular, the role of dynamical bias.
-This is followed by a very reasonable proposal for how the problem may be addressed.
-Applying these methods to a re-analysis of data on within- and between- household spread of delta and omicron variants of SARS-CoV-2 suggests that while both had similar incubation-period distributions, the mean generation interval for omicron was shorter by 0.3-0.8 days than previously estimated (0.6 days).
-The original analysis had also reported a difference in mean incubation periods of 3.2 days for omicron versus 4.4 days for delta; the methods used in this paper suggest that both have a mean of 4.1 days.
-The consequences of discriminating between generation time intervals on the estimation of transmission advantage is shown in Fig 5, although it is not clear how this pertains to dynamical bias as both the original study and this one reported a difference between the variants in this respect.}
+\revtext{I am happy with the revisions made to this paper.}
 
-Dynamical biases pertain to the estimation of delay distributions, and not to the estimation of transmission advantage. 
-We agree that both the original study and this one report that the Omicron variant has a higher reproduction number. 
-Here, what we show is that estimates of transmission advantage depend on assumptions about the generation-interval distributions. 
-For example, as we explain in the Discussion, assuming longer generation-interval distributions can lead to further biases in the estimates of transmission advantange:
+Thank you for your review.
 
-``For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province.''
+\rev{Reviewer \#2}
 
-\revtext{The caveats of this study, and the problems of applying this method to other datasets, are very thoroughly outlined in the discussion.}
+\revtext{I am still of the opinion that the results, while undoubtedly interesting and important for infectious disease epidemiologists, do not present such a general interest to deserve publication in PNAS.
+Of course, the general public interest in COVID-19 can justify its publication anyway.
+The methods are not especially novel, although they are very well presented.
+The results are quite important since they provide the first high-quality estimates for the generation time of Omicron, while all existing estimates are unfortunately heavily biased and therefore unreliable.}
+
+Thank you for your review.
+We now discuss how disease generation time is linked to generation time in population biology.
+We hope that this discussion provides a bit more general interest to a wider group of readers.
+
+\revtext{Most of the changes by the authors concern the issue of why Omicron would have a shorter generation intervals.
+I had no problem with their discussion of this topic even in the previous version (though it was quite compact and technical) and I am pleased with the extended discussion they provide on the issue.
+Their argument about the network effects is quite hypothetical and I agree that it may be the weakest part of the paper.
+However, there are excellent reasons to believe that the force of infection during the Omicron wave in Europe has been much more intense than at any point during the Delta wave.
+Therefore, I am inclined to regard this hypothesis as a very good working hypothesis and I agree with the way it is presented.}
 
@@
 
-\revtext{Less attention is paid to the more important question of WHY the omicron variant had such a short generation time in this setting.}
+\revtext{The power-law dependence that the authors assume between the typical generation time for a contact pair and the incubation period of the infectee is a quite curious assumption.
+It is basically a consequence of the simplifying assumption that the joint distribution is a bivariate lognormal.
+I understand that the authors have no joint data to explore the validity of this assumption, and it would be pointless to push for a more complicated analysis without the right data.
+However, worth pointing out that there are more detailed analyses of this relation using better data from transmission pairs [Ferretti 2020, Hart 2021] that seem to suggest different functional forms, although they mostly refer to the original strain from Wuhan.}
 
-We have now added text to the discussion explaining different types of generation intervals (intrinsic vs realized) and why the realized generation interval of the Omicron variant might be shorter:
+Done.
 
-``Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
-\cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
-Here, we we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
-The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.
+\revtext{When the authors write 
+"With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant" 
+it would be nice to point out the contact tracing studies that estimate the same ratio from secondary attack rates and obtain similar numbers (e.g.
+~3 in UK, see UKHSA/Test&Trace reports from December 2021).
+The agreement between these numbers and their result would strengthen their argument about the importance of using proper generation interval estimates to obtain reliable estimates of relative reproduction advantage.}
 
-Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
-For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
-Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
-In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
-this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
-However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
-some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
-We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
-immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.
+\revtext{Minor comment: some of the preprints cited in the manuscript have been published in journals. Hay 2022 for example is here: https://elifesciences.org/articles/81849}
 
-Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions (e.g., a lack of changes in viral kinetics or the effect of pre-existing immunity), their realized generation-interval distributions could still differ.
-For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
-Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
-If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
-However, this explanation is also less likely given similarities in the inferred incubation-period distributions.
-
-Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
-To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
-We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
-Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
-Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
-While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
-
-Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
-In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their infectiousness profiles may be smaller than the estimated differences in their realized generation-interval distributions.
-In addition, the counter-factual initial intrinsic generation intervals (for an immunologically naive population) of both variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
-Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
-and can therefore be expected to improve estimates of effective reproduction numbers.''
-
-\revtext{The authors discuss the possibility of a "network effect" which is predicated on omicron having a higher Ro and therefore somewhat circular.}
-
-We don't believe that this explanation is circular, and have tried to clarify our explication further.
-As we explain in the Discussion, the success of the Omicron variant at the population level is due to immune evasion (also pointed out by reviwer 1), and this doesn't require shorter generation intervals.
-We now provide a concrete example for the network effect, explaining how higher reproduction number can lead to shorter \emph{realized} generation intervals even when two variants have identical \emph{intrinsic} generation intervals:
-
-``Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
-To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
-We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect without requiring new simulations or an extensive review of existing literature.
-Under the same scenario, if each individual were living in a household containing four other people, the individual infected with variant A will infect one household member every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
-Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
-While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).''
-
-\revtext{Other factors "such as more stringent contact tracing measures against the Omicron variant in the Netherlands [4], faster within-host clearance of the Omicron variant [35], and viral kinetics of reinfection and breakthrough infections" are mentioned without sufficient explanation as to why they would yield such a result.}
-
-We now provide a more detailed explanation.
-
-\revtext{The role of immune evasion - although mentioned in passing - has not been sufficiently accommodated either in the discussion or in the analysis itself, even though it presents itself as the most likely explanation for the success of omicron in almost every global setting.}
-
-We now discuss the possible effects of immune evasion more thorougly.
-
-\revtext{To address this properly, the authors will have to account for the different proportions of the population currently immune to infection by either variant and how the level of cross-immunity between them affects our observations of transmission advantage.
-The authors should at least justify why they have not taken such an obvious feature of this multi-strain system into account.}
-
-While we agree that immune evasion is the most likely explanation for the success of omicron in almost every global setting, its success or immune evasion does not require or depend on changes in generation-interval distributions.
-We now provide a more thorough explanation for how pre-existing immunity and immune evasiveness of the Omicron variant might shape the generation-interval distribution:
-
-``Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
-In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
-this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
-However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
-some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them---their combined effects on the resulting generation-interval distribution are unclear.
-We note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
-immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.''
-
-\rev{Reviewer \#2}
-
-\revtext{The significance statement is accurate.
-I find the paper definitely relevant for future epidemiological studies on COVID-19, and even important for some technical works.
-But neither the methods nor the results are likely to have a huge impact on the field, even less on other fields.}
-
-\revtext{The manuscript is technically sound and very well written.
-The results are convincing and clearly explained in the manuscript, and the arguments presented make sense from an epidemiological perspective.
-This paper emphasizes once more the importance of accounting for the epidemic dynamic (or the differential dynamic between two variants, as in this case) when inferring fundamental epidemiologically relevant quantities that vary in time during the infection.
-I find it an important contribution to the literature, albeit a technical one.}
-
-Thank you for your review.
-
-\bibliography{omicron}
+Fixed.
 

### 2023-03-28 omicron-generation.tex
making changes

@@
 
-Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the COVID-19 pandemic \citep{park2021roles}.
+Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the COVID-19 pandemic \citep{park2022importance}.
 Transmission advantages of invading variants are typically characterized by the ratios of reproduction numbers, $\Rinv/\Rres$, and the differences in growth rates, $\rinv-\rres$.
@@
 
+We assume that incubation periods and generation intervals follow a bivariate lognormal distribution---in doing so, we implicitly assume that two intervals have a power law dependence.
+Instead, previous studies showed that mechanistic models better capture the relationship between incubation period and generation interval \citep{ferretti2020timing,hart2021}. 
+Although we did not have sufficient data to test our assumptions, \cite{sender2021unmitigated} showed that a bivariate lognormal distribution can also capture the relationship between incubation period and generation interval well.
+Future studies should compare different functional forms to better understand their relationships.
+
 We do not account for individual-level heterogeneity, such as age, vaccination status, or previous exposure history.
@@
 These conclusions are relevant for other emerging and endemic pathogens in general.
+
+The study of disease generation time has broader implications for estimating generation time of a species in population biology.
+Historically, three ways of estimating of generation time have been discussed \citep{leslie1966intrinsic}.
+First, the mean age of reproduction corresponds to the forward distribution and is therefore insensitive to the underlying growth rate.
+Second, the mean age of mothers for a cohort of offspring that are born at the same year corresponds to the backward distribution, and therefore decreases as the growth rate increases.
+Finally, the ratio between log reproduction number and log multiplicative growth rate (equivalent to $r$ in epidemiology) corresponds to the intrinsic generation time with an additional assumption that the generation time is fixed.
+The link between serial and generation intervals further provides a theoretical basis for understanding generation time for a stage-structured population---the time between symptom onset is analogous to measuring generation time via time between successive recuirtment events for a given life stage \citep{lebreton2005age,bienvenu2015new}.
+Comparing generation-time definitions across two fields can provide valuable insight in understanding population dynamics.
 

### 2023-03-28 omicron-generation.tex
homing (台大)

--- 2022-12-26 omicron-generation.tex
@@
 
+We assume that incubation periods and generation intervals follow a bivariate lognormal distribution---thus, we implicitly assume a power-law relationship between them. 
+This was done because data were limited, and previous work \cite{sender2021unmitigated} has shown that this approach can also capture important aspects of the relationship.
+In fact, however, previous studies have shown that mechanistic models capture this relationship better \citep{ferretti2020timing,hart2021}. 
+When sufficient data are available, more mechanistic approaches may therefore be more suitable.
+
 We do not account for individual-level heterogeneity, such as age, vaccination status, or previous exposure history.
@@
 These conclusions are relevant for other emerging and endemic pathogens in general.
+
+\jd{Done quickly, needs cleaning (particularly cites).}
+This study also has potential implications for studying generation times in population biology \citep{leslie1966intrinsic}.
+Mean age of reproduction for a cohort of mothers corresponds to the forward generation distribution, while mean age of mothers for a cohort of offspring corresponds to the backward distribution.
+Both could therefore be affected by dynamical considerations.
+Similarly, the connection between generation times based at different recruitment stages is analogous to that between serial and generation intervals \citep{lebreton2005age,bienvenu2015new}.
+Comparing definitions and methods across these fields could provide valuable future insights.
 

### 2023-03-31 response.tex
homing (台大)
done

@@
 We now discuss how disease generation time is linked to generation time in population biology.
-We hope that this discussion provides a bit more general interest to a wider group of readers.
+We hope that this discussion provides a bit more general interest to a wider group of readers:
+
+``This study also has potential implications for studying generation times in population biology \citep{leslie1966intrinsic,mertz1971mathematical,caswell2009stage,steiner2014generation}.
+For example, mean age of reproduction for a cohort of mothers corresponds to the forward generation interval, while mean age of mothers for a cohort of offspring corresponds to the backward interval.
+Their difference are therefore likely affected by dynamical biases \citep{jonasson2022unifying}.
+Similarly, the connection between generation times based at different recruitment stages is analogous to that between serial and generation intervals \citep{lebreton2005age,bienvenu2015new,giaimo2023generation}.
+Comparing definitions and methods across these fields could provide valuable future insights.''
 
@@
 
-Done.
+We have added the following paragraph to the Discussion:
+
+``We assume that incubation periods and generation intervals follow a bivariate lognormal distribution---thus, we implicitly assume a power-law relationship between them. 
+This was done because data were limited, and previous work \cite{sender2021unmitigated} has shown that this approach can also capture important aspects of the relationship.
+In fact, however, previous studies have shown that mechanistic models capture this relationship better \citep{ferretti2020timing,hart2021}. 
+When sufficient data are available, more mechanistic approaches may therefore be more suitable.''
 
@@
 it would be nice to point out the contact tracing studies that estimate the same ratio from secondary attack rates and obtain similar numbers (e.g.
-~3 in UK, see UKHSA/Test&Trace reports from December 2021).
+~3 in UK, see UKHSA/Test\&Trace reports from December 2021).
 The agreement between these numbers and their result would strengthen their argument about the importance of using proper generation interval estimates to obtain reliable estimates of relative reproduction advantage.}
+
+We have added the following sentence:
+
+``Overall, our revised estimates are more consistent with contact tracing studies that reported 1.5--2.2 fold increase in secondary attack rates in household and non-household settings \citep{allen2022comparative,del2022secondary}.''
 
@@
 
+\bibliography{omicron.bib}
+
 \end{document}

### 2023-03-31 omicron-generation.tex
done

@@
 With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province in South Africa.
+Overall, our revised estimates are more consistent with contact tracing studies that reported 1.5--2.2 fold increase in secondary attack rates in household and non-household settings \citep{allen2022comparative,del2022secondary}.
 
@@
 
-\jd{Done quickly, needs cleaning (particularly cites).}
-This study also has potential implications for studying generation times in population biology \citep{leslie1966intrinsic}.
-Mean age of reproduction for a cohort of mothers corresponds to the forward generation distribution, while mean age of mothers for a cohort of offspring corresponds to the backward distribution.
-Both could therefore be affected by dynamical considerations.
-Similarly, the connection between generation times based at different recruitment stages is analogous to that between serial and generation intervals \citep{lebreton2005age,bienvenu2015new}.
+This study also has potential implications for studying generation times in population biology \citep{leslie1966intrinsic,mertz1971mathematical,caswell2009stage,steiner2014generation}.
+For example, mean age of reproduction for a cohort of mothers corresponds to the forward generation interval, while mean age of mothers for a cohort of offspring corresponds to the backward interval.
+Their difference are therefore likely affected by dynamical biases \citep{jonasson2022unifying}.
+Similarly, the connection between generation times based at different recruitment stages is analogous to that between serial and generation intervals \citep{lebreton2005age,bienvenu2015new,giaimo2023generation}.
 Comparing definitions and methods across these fields could provide valuable future insights.

### 2023-04-10 omicron-generation.tex
done

@@
 With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province in South Africa.
-Overall, our revised estimates are more consistent with contact tracing studies that reported 1.5--2.2 fold increase in secondary attack rates in household and non-household settings \citep{allen2022comparative,del2022secondary}.
+Overall, our revised estimates are more consistent with contact tracing studies that reported a 1.5--2.2 fold increase in secondary attack rates across household and non-household settings of Omicron relative to Delta variants \citep{allen2022comparative,del2022secondary}.
 
@@
 
-\section*{Supplementary Materials}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_compare_rho.ggp.pdf}
-\caption{
-\textbf{Sensitivity of the estimates of the mean generation interval to the assumed values of the correlation coefficient of the lognormal distribution.}
-Lines and shaded regions represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
-For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 (13--19 December) and 51 (20--26 December) of 2021.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_compare_stratified.ggp.pdf}
-\caption{
-\textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
-Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-household transmission) and the week of infectors' symptom onset (week 50, 13--19 December 2021, vs week 51, 20--26 December 2021,).
-Lines represent maximum likelihood estimates.
-Shaded regions represent the corresponding 95\% confidence intervals.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_reproduction_advantage_between.ggp.pdf}
-\caption{
-\textbf{Estimated time-varying reproduction number advantages of the Omicron variant using between-household generation-interval distributions.}
-(A) Estimated instantaneous reproduction numbers and their ratios over time while accounting for differences in the generation-interval distributions.
-(B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
-The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
-The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the forward generation-interval distributions for between-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
-Purple lines represent the ratio between the effective reproduction numbers of the Delta and Omicron variants.
-Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
-}
-\end{figure}
-
-\pagebreak
-
 \bibliography{omicron.bib}

### 2023-04-10 supp.tex
done

--- 2023-03-28 omicron-generation.tex
@@
 \documentclass[12pt]{article}
-% \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
-\usepackage[top=1in,footskip=1in]{geometry}
+\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
 
@@
 \usepackage{lineno}
-\linenumbers
 
@@
 \usepackage{natbib}
-\bibliographystyle{unsrt}
+\bibliographystyle{chicago}
 \date{\today}
@@
 
+\renewcommand{\theequation}{S.\arabic{equation}}
+
 \newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
+\newcommand{\Ro}{\Rx{0}}
+\newcommand{\Rc}{\Rx{\mathrm{c}}}
+\newcommand{\Rs}{\Rx{\mathrm{s}}}
 \newcommand{\RR}{\ensuremath{{\mathcal R}}\xspace}
-\newcommand{\Rres}{\Rx{\mathrm{res}}}
-\newcommand{\Rinv}{\Rx{\mathrm{inv}}}
 \newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
-\newcommand{\Rt}{\ensuremath{{\mathcal R}(t)}\xspace}
+\newcommand{\Rintrinsic}{\ensuremath{{\mathcal R}_{\textrm{\tiny intrinsic}}}\xspace}
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
@@
 \newcommand{\dsigma}{\dd{\sigma}}
-
-\newcommand{\rx}[1]{\ensuremath{{r}_{#1}}\xspace} 
-\newcommand{\rres}{\rx{\mathrm{res}}}
-\newcommand{\rinv}{\rx{\mathrm{inv}}}
 
@@
 \newcommand{\gdist}{g} %% generation-interval distribution
-\newcommand{\idist}{\ell} %% incubation-period distribution
+\newcommand{\idist}{\ell} %% incubation period distribution
 
@@
 
+\renewcommand{\thesection}{S\arabic{section}}
+
 \begin{document}
@@
 	\textbf\newline{
-		Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2
+		Supplementary Materials---Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2
 	}
@@
 
-\section*{Abstract}
-
-Estimating the differences in the incubation-period, serial-interval, and generation-interval distributions of SARS-CoV-2 variants is critical to understanding their transmission and control.
-However, the impact of epidemic dynamics is often neglected in estimating the timing of infection and transmission---for example, when an epidemic is growing exponentially, a cohort of infected individuals who developed symptoms at the same time are more likely to have been infected recently.
-Here, we re-analyze incubation-period and serial-interval data describing transmissions of the Delta and Omicron variants from the Netherlands at the end of December 2021.
-Previous analysis of the same data set reported shorter mean observed incubation period (3.2 days vs 4.4 days) and serial interval (3.5 days vs 4.1 days) for the Omicron variant, but the number of infections caused by the Delta variant decreased during this period as the number of Omicron infections increased.
-When we account for growth-rate differences of two variants during the study period, we estimate similar mean incubation periods (3.8--4.5 days) for both variants but a shorter mean generation interval for the Omicron variant (3.0 days; 95\% CI: 2.7--3.2 days) than for the Delta variant (3.8 days; 95\% CI: 3.7--4.0 days).
-We further note that the differences in estimated generation intervals may be driven by the ``network effect''---higher effective transmissibility of the Omicron variant can cause faster susceptible depletion among contact networks, which in turn prevents late transmission (therefore shortening realized generation intervals).
-Using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage of the Omicron variant.
-
-\section*{Significance}
-
-Recent studies suggest that individuals infected with the Omicron variant develop symptoms earlier (shorter incubation period) and transmit faster (shorter generation interval) than those infected with the Delta variant.
-However, these studies typically neglect population-level effects: 
-when an epidemic is growing, a greater proportion of current cases were infected recently, biasing us toward observing faster transmission events.
-Accounting for this dynamical bias, we find that Omicron infections from the Netherlands at the end of December 2021 had similar incubation periods, but shorter generation intervals, compared to Delta infections from the same period.
-Shorter generation intervals of the Omicron variant might be due to its higher effective reproduction number, which can cause faster local susceptible depletion around the contact network.
-
+Disclaimer: The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S.
+Department of Health and Human Services.
 \end{flushleft}
-
-\pagebreak
-
-\section{Introduction}
-
-Estimating transmission advantages of new SARS-CoV-2 variants is critical to predicting and controlling the course of the COVID-19 pandemic \citep{park2022importance}.
-Transmission advantages of invading variants are typically characterized by the ratios of reproduction numbers, $\Rinv/\Rres$, and the differences in growth rates, $\rinv-\rres$.
-These quantities are linked by the generation-interval distributions of the resident and invading variants.
-For example, an invading variant with shorter generation intervals---defined as the time between infection of the infector and the infectee---will exhibit faster epidemic growth ($\rinv > \rres > 0$) even if their reproduction numbers are identical ($\Rinv = \Rres > 1$).
-
-Estimating the generation-interval distribution is challenging, in part due to difficulties in observing actual infection events.
-Many researchers primarily focus on comparisons of other transmission intervals, such as the time between symptom onsets (also referred to as serial intervals) or between testing events \citep{abbott2022test} of the infector and the infectee.
-Each of these transmission-interval distributions can be subject to dynamical effects, which can cause transmission-interval distributions to systematically differ from the corresponding generation-interval distribution.
-
-For example, when the epidemic is growing, there will be more recent infections, and we are therefore more likely to observe recently infected individuals among a cohort of infectors who developed symptoms at the same time.
-In this case, their incubation periods will be shorter, on average, than those of their infectees, causing the mean serial interval to be longer than the mean generation interval \citep{park2021forward}.
-We refer to such effects of growth rate on expected intervals as ``dynamical bias''
-Because of dynamical bias, observed differences in transmission-interval distributions between variants are not necessarily equivalent to differences in the underlying generation-interval distributions when their growth rates differ.
-
-Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the B.1.617.2 (Delta) and B.1.1.529 (Omicron) variants from the Netherlands between 13 and 26 December 2021.
-The study found shorter mean within-household serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
-during this period, the incidence of Omicron cases were increasing, whereas the incidence of Delta cases were decreasing.
-We take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
-We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.
-
-\section{Methods}
-
-\subsection{Data}
-
-We analyze time series of reported COVID-19 cases (\url{https://data.rivm.nl/covid-19/}) and proportions of SARS-CoV-2 variants detected (\url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}) from the Netherlands between 29 November 2021 and 30 January 2022.
-Data sets are publicly available on the National Institute for Public Health and the Environment (RIVM) website.
-
-Serial-interval data are taken from \cite{backer2021omicron}.
-Infector-infectee pairs were identified through contact tracing, and their symptom onset dates were reported through a national surveillance database.
-Serial intervals were then calculated by taking the difference between symptom onset dates of the infector and the infectee.
-In order to ensure independence between serial intervals, one infectee was chosen at random for each infector in the original analysis.
-See original article for additional details of data collection.
-
-Publicly available data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates, symptom onset dates, or age.
-The original article presented serial-interval estimates stratified by the vaccination status in supplementary materials, but stratified data are not publicly available; we rely on publicly available data to keep the analysis simple and to focus on the qualitative effects of dynamical biases.
-The aggregated data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (week 50, 13--19 December 2021, and week 51, 20--26 December 2021), and the type of transmission (within- or between-household transmission).
-In the main text, we combine data from weeks 50 and 51 of 2021 (13--26 December) and present a stratified analysis in Supplementary Material.
-For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
-Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
-however, the data are not publicly available with the original article. 
-Instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.
-
-\subsection{Estimating epidemic growth rates}
-
-In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, their growth rates---into account.
-To estimate the growth rates of the Delta and Omicron variants, we first estimate the number of COVID-19 cases infected with each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week.
-We note that the proportion of Delta and Omicron variants detected is reported with the date of sampling, whereas the case data are reported with the date of report, meaning that there is some delay between the two data sets (typically around 2 days).
-For simplicity, we do not account for this delay in our growth-rate estimates; 
-instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
-We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making uncertainty due to sample size small;
-we note however that this estimate is also sensitive to  the assumption that sampling is random.
-
-We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
-More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a penalized cubic spline fitted with restricted maximum likelihood (specified as \texttt{gam(log(cases)$\sim$s(time, bs="cs"), method="REML")} using the \texttt{MGCV} \texttt{R} package):
-We use Gaussian likelihoods to fit to logged cases in part for convenience, and in part because the inferred numbers of cases infected with each variant are not whole numbers.
-In principle, it might be preferable to explicitly model the process of sampling for strain testing, and then using negative-binomial likelihoods for case numbers \citep{ver2007quasi}, but our main purpose here is simply to roughly estimate growth rates with reasonable uncertainties.
-Finally, we take the derivative of the predicted logged numbers of cases infected with each variant to obtain time-varying growth rate estimates.
-
-To obtain confidence intervals on the estimated time-varying growth rates, we generate 1000 parameter sets by resampling spline coefficients from a multivariate normal distribution using the estimated variance-covariance matrices.
-We calculate time-varying growth rates from each parameter set and use equi-tailed quantiles to generate 95\% confidence limits. 
-We note that this method of calculating confidence intervals gives point-wise confidence intervals, meaning that the confidence intervals give 95\% coverage for the set of estimates at each time point;
-these intervals are narrower than simultaneous confidence intervals, which give 95\% coverage for the set of estimated time series across the whole time period \citep{simpson2017}.
-
-\subsection{Estimating forward incubation-period distributions from backward incubation-period distributions}
-
-The incubation-period distributions from 513 individuals (consisting of 258 Omicron and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022, were previously reported in \cite{backer2021omicron}.
-These data cover a wider time period than the serial-interval data.
-\cite{backer2021omicron} used the methods of \cite{backer2020incubation}, which estimates incubation period by inferring distributions of time of infection for each individual from their known exposure dates.
-In particular, the methods of \cite{backer2020incubation} assume that the infection time is uniformly distributed across exposure dates and compares the inferred infection time to a known symptom-onset time to calculate the incubation period for each individual.
-Even if this method can accurately estimate the infection time, and therefore the incubation period, of each individual, dynamical biases can still affect this sort of cohort-based estimation of incubation period.
-
-More specifically, incubation periods (and other epidemiological delays) can be measured in two ways: forward and backward \citep{park2021forward}.
-Forward incubation periods are measured from a cohort of individuals who were infected at the same time. 
-We expect the forward incubation-period distribution $f_I(\tau)$ to remain relatively constant over the course of an epidemic of one given variant, although biases can arise in \emph{observing} incubation periods, based on public or medical awareness of the disease.
-Backward incubation periods are measured from a cohort of individuals who developed symptoms at the same time. 
-The backward incubation-period distribution is sensitive to epidemic dynamics:
-the difference between the forward and backward distribution arises because forward incubation periods look forward from the reference point towards symptom development, which is an individual-level process, while backward incubation periods look backwards towards an infection event, which requires an interaction with an infectious individual.
-
-In particular, when incidence of infection is growing exponentially, we are more likely to observe backward incubation periods that are shorter than the corresponding forward incubation periods because there will be relatively more individuals who were infected recently.
-Assuming that incidence of infection is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
-\begin{equation}
-b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx},
-\label{eq:backward}
-\end{equation}
-where the denominator is a normalization constant so that $b_I(\tau)$ integrates to 1.
-Therefore, the backward incubation-period distribution $b_I(\tau)$ gives a biased estimate of the corresponding forward distribution $f_I(\tau)$.
-The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution.
-
-Assuming a constant growth rate $r$, the corresponding forward incubation-period distributions can be calculated by inverting \eref{backward}, taking into account that $f_I$ is a probability distribution and therefore needs to be normalised to integrate to 1:
-\begin{equation}
-f_I(\tau) = \frac{\exp(r\tau) b_I(\tau) }{\int_0^\infty \exp(rx) b_I(x)\dx}.
-\label{eq:forward}
-\end{equation}
-Since incubation-period data are not provided, we are not able to fit \eref{forward} directly; 
-instead we take the backward incubation-period distributions $b_I(x)$ estimated by \cite{backer2021omicron}, which was originally assumed to follow a Weibull distribution, and apply \eref{forward}.
-In particular, \cite{backer2021omicron} estimated the scale and shape parameters of the Weibull distribution to be 4.93 (95\% CI: 4.51--5.37) and 1.83 (95\% CI: 1.59--2.08), respectively, for the Delta cases, and 3.60 (95\% CI: 3.23--3.98) and 1.50 ((95\% CI: 1.32--1.70), respectively, for Omicron cases.
-
-We also model the backward incubation-period distribution $b_I(\tau)$ using a Weibull distribution based on the assumptions of \cite{backer2021omicron}.
-To account for uncertainties in the original parameter estimates, we rely on a sampling scheme, similar to the one we used for the growth rate analysis (in Section 2.2).
-First, we approximate the previously inferred posterior distributions of the shape and scale parameters of the Weibull distribution using a lognormal distribution---we parameterize the lognormal distribution such that (i) its median matches the median of the posterior distributions and (ii) the probability that a random variable following the specified lognormal distribution falls between the lower and upper credible limits is 95\% \citep{park2020reconciling}.
-We draw 1000 samples of the shape and scale parameters (for the backward distribution $b_I(\tau)$) from the specified lognormal distributions and estimate the corresponding forward distribution using \eref{forward}.
-We take 95\% equi-tailed quantiles to obtain 95\% confidence intervals.
-We repeat the analysis across plausible ranges of $r$ for the Delta and Omicron variants separately (discussed later).
-
-\subsection{Estimating forward generation-interval distributions from forward serial-interval distributions}
-
-\begin{figure}[!tp]
-\includegraphics[width=\textwidth]{figure/figure_diagram.ggp.pdf}
-\caption{
-\textbf{Schematic diagrams of serial and generation intervals from symptom- and infection-based infector cohorts.}
-(A) Forward serial intervals are measured from the cohort of infectors who develop symptoms at the same time. 
-In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
-(B) Generation intervals for the cohort of infectors who are infected at the same time are not biased by dynamical effects on incubation periods.
-\label{fig:diagram}
-}
-\end{figure}
-
-Dynamical biases in the serial-interval distributions are more complex because the serial interval depends on the incubation periods of the infector and the infectee as well as the generation interval between them (\fref{diagram}).
-For example, \cite{backer2021omicron} measured the forward serial-interval distributions from cohorts of infectors who developed symptoms during the same week.
-In this case, the forward serial interval $\tau_s$ can be expressed in the form \citep{park2021forward}:
-\begin{equation}
-\tau_s =  - \tau_{i, 1} + \tau_{g, \mathrm{symp}} + \tau_{i, 2},
-\label{eq:serial}
-\end{equation}
-where $\tau_{i, 1}$ represents the backward incubation period of the infector (because all infectors developed symptoms at the same time), and $\tau_{i, 2},$ represents the forward incubation period of the infectee.
-Here, $\tau_{g, \mathrm{symp}}$ represents the generation interval between the infector and the infectee; we use the subscript $\mathrm{symp}$ to indicate that these generation intervals are measured from infectors who developed symptoms at the same time.
-
-The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased (compared to the generation-interval distribution for an infection-based cohort) because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and are therefore likely to transmit earlier (\fref{diagram}A).
-This generation-interval distribution for a symptom-based cohort depends on the backward incubation-period distribution:
-\begin{align}
-f_{G, \mathrm{symptom}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) b_I(x) \dx,
-\end{align}
-where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$, and $b_I(x)$ represents the backward incubation-period distribution.
-Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time, \fref{diagram}B):
-\begin{align}
-f_{G, \mathrm{inf}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) f_I(x) \dx.
-\label{eq:meangen}
-\end{align}
-Previous analyses of serial-interval distributions typically assumed that the incubation periods and generation intervals are independent \citep{ganyani2020estimating}; in this case, the generation-interval distribution for the symptom-based and infection-based cohorts are identical.
-
-In summary, when an epidemic is growing exponentially, there are two opposing effects affecting the relationship between the mean forward serial and generation interval.
-First, infectors who developed symptoms at the same time are more likely to have shorter (backward) incubation periods than the corresponding forward incubation periods of their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
-Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}}] > \mathbb E[\tau_{g, \mathrm{symp}}]$ due to correlations between incubation periods and generation intervals.
-Therefore, the difference between the mean forward serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean forward serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
-Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed the authors to conclude that the mean generation and serial intervals are identical.
-For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time, \fref{diagram}B), and drop the subscript $\mathrm{inf}$.
-
-Assuming that the incidence of infection will continue to change exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged through time \citep{park2021forward}.
-Then, we can focus on the forward serial-interval distribution at $t=0$, which in turn allows us to reparameterize the incubation-period and generation-interval distributions in terms of the infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$.
-Under this parameterization, for a given length of a serial interval $\tau$, we can rewrite the incubation period of the infector as $-\alpha_1$; the generation interval as $\alpha_2 - \alpha_1$; and the incubation period of the infectee as $\tau - \alpha_2$.
-Then, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be expressed in terms of three distributions (\eref{serial}): the backward incubation-period distribution of the infector $b_I(-\alpha_1)$ (taken from \eref{backward}), the forward generation-interval distribution conditional on a known value of the incubation period $x$, $f_{G|I}(\alpha_2-\alpha_1|-\alpha_1)$, and the forward incubation-period distribution of the infectee $f_I(\tau - \alpha_2)$.
-Integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ and rewriting the backward incubation-period distribution $b_I(-\alpha_1)$ in terms of the forward distribution, we obtain \citep{park2021forward}:
-\begin{equation}
-f_S(\tau) = \frac{1}{\phi} \int_{-\infty}^0\int_{\alpha_1}^\tau \exp(r \alpha_1) f_{G|I}(\alpha_2 - \alpha_1|- \alpha_1) f_I(- \alpha_1) f_I(\tau - \alpha_2) \dd \alpha_2 \dd \alpha_1,
-\label{eq:forwardserial}
-\end{equation}
-where $\phi$ is a normalization constant chosen so that $\int f_S(x) \dx = 1$.
-As discussed earlier, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
-As we show in Results, the exponential growth rate changes over the study period, including weeks 50 and 51 (13--26 December 2021);
-for illustrative purposes, we choose representative values of $r$ that for Delta and Omicron during this period and also explore across plausible ranges of $r$ (see below).
-
-While the derivation of the forward serial-interval distribution \eref{forwardserial} may be complex, its implementation is simple.
-The main difference between our model and previous models that neglect dynamical effects \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation} is the exponential growth term $\exp(r \alpha_1)$ and the normalization term $\phi$---it is relatively straightforward to include these terms in existing models of serial intervals.
-\cite{ferretti2020quantifying, ferretti2020timing} also included this term in their analyses of serial-interval data, but only accounted for the epidemic growth effect (and not the decay effect).
-
-We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
-The joint distribution is parameterized by log-scale means, $\mu_I$ and $\mu_G$, log-scale variances, $\sigma_I^2$ and $\sigma_G^2$, and the log-scale correlation coefficient $\rho$.
-Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log-scale mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log-scale variance of $\sigma_G^2 (1 - \rho^2)$.
-For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
-We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation (also known as the method of moments); we note that we are unable to directly fit a lognormal distribution to the forward incubation-period distribution because we are relying on existing estimates rather than raw data.
-Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
-We then calculate the mean forward generation interval using \eref{meangen}.
-The 95\% confidence intervals are calculated by taking the estimated variance-covariance matrix for our mean and standard deviation parameters and calculating the corresponding variance-covariance for the overall mean using Taylor expansion---this method is also known as the Delta method \citep{oehlert1992note}.
-We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.
-In Supplementary Material, we explore how assumptions about $\rho$ affect inferences of the generation-interval distribution.
-
-\subsection{Estimating instantaneous reproduction numbers}
-
-We use our estimates of the generation-interval distributions to infer instantaneous reproduction numbers \Rt\ of the Delta and Omicron variant, as well as the ratio between the two reproduction numbers.
-Estimating the instantaneous reproduction number---defined as the average number of secondary infections that a primary case will generate if epidemiological conditions remain constant \citep{fraser2007estimating}---requires the intrinsic generation-interval distribution $g(\tau)$:
-\begin{equation}
-\Rt = \frac{i(t)}{\int_0^\infty i(t-x) g(x) \dx},
-\end{equation}
-where $i(t)$ represents incidence of infection.
-Here, we approximate the intrinsic generation-interval distribution with the forward generation-interval that we estimate for weeks 50 and 51 of 2021 (13--26 December)---when the epidemic is growing or decaying exponentially, we expect the forward generation-interval to be a good proxy for the intrinsic generation-interval distribution \citep{champredon2015intrinsic, park2020inferring}.
-Incidence of infection is approximated by shifting the smoothed case trajectories by one week to account for reporting delays.
-This method of approximating incidence of infection assumes a fixed delay between infection and case reporting;
-in practice, deconvolution is required to accurately estimate the incidence of infection \citep{goldstein2009reconstructing}.
-Case reports are also sensitive to changes in testing behavior, and therefore our estimates of $\Rt$ must be interpreted with care.
-Confidence intervals are calculated by sampling parameters of the smoothed case trajectories as well as the generation-interval distributions from multivariate normal distributions and repeating the analysis 1000 times.
-
-\section{Results}
-
-\fref{epidemic} summarizes the epidemiological context in the Netherlands during the study period.
-The first known Omicron case in the Netherlands was sampled on 19 November 2021 \citep{backer2021omicron}, during a period when COVID-19 incidence was decreasing (\fref{epidemic}A).
-As the Omicron variant continued to spread and increase in proportion (\fref{epidemic}B), the number of COVID-19 cases started to increase (\fref{epidemic}A).
-Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
-The number of COVID-19 cases infected with the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022 (\fref{epidemic}D).
-The number of COVID-19 cases infected with the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
-These changes in growth rates coincide with the introduction of lockdown on 19 December 2021 \citep{netherlandslockdown} and its relaxation beginning 15 January 2022 \citep{netherlandsrelax,netherlandsopen}.
-We note that the growth-rate difference between the Delta and Omicron variants decreased over time.
-Hereafter, we use $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as representative growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
-We then evaluate the growth-rate effects across $r=-0.1/\mathrm{day}$--$0.0/\mathrm{day}$ for the Delta variant and $r=0.1/\mathrm{day}$--$0.2/\mathrm{day}$ for the Omicron variant as a sensitivity analysis.
-
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{figure/figure_epidemic.ggp.pdf}
-\caption{
-\textbf{Epidemic dynamics of the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
-(A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
-The solid line represents the 7-day moving average.
-Data are publicly available at \url{https://data.rivm.nl/covid-19/}.
-(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available at \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
-(C) Weekly numbers of COVID-19 cases infected with the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
-Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
-(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth-rate differences (purple).
-Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
-Growth rates are estimated by taking the derivative of the generalized additive model estimates of logged number of cases.
-\label{fig:epidemic}
-}
-\end{figure}
-
-Previous analysis of a cohort of individuals who developed symptoms between 1 December 2021 and 2 January 2022 found longer mean (backward) incubation period for the Delta variant than for the Omicron variant \citep{backer2021omicron} (\fref{incubation}A).
-However, when we account for growth-rate differences and re-estimate the forward incubation periods, we find that both variants have similar incubation-period distributions with a mean of 4.1 days (95\% CI: 3.8--4.4 days) for the Delta variant and 4.2 days (95\% CI: 3.6--4.9 days) for the Omicron variant \fref{incubation}B).
-In this case, the difference between the mean backward and forward incubation periods correspond to $-22\%$ and $7\%$ bias for the Omicron and Delta variants, respectively.
-Although the exact estimate of the mean forward incubation periods of both variants are sensitive to the assumed growth/decay rates, we find similar means across a plausible ranges of growth rates (\fref{incubation}C--D).
-For example, the mean forward incubation period of the Delta variant changes from 3.8 days (95\% CI: 3.5--4.1 days) to 4.4 days (95\% CI: 4.0--4.8 days) as we change the assumed values of $r$ from $-0.1/\mathrm{days}$ to $0.0/\mathrm{days}$ (\fref{incubation}C),
-while the mean forward incubation period of the Omicron variant changes from 3.8 days (95\% CI: 3.4--4.4 days) to 4.5 days (95\% CI: 3.9--5.5 days) as we change the assumed values of $r$ from $0.1/\mathrm{days}$ to $0.2/\mathrm{days}$ (\fref{incubation}D).
-Wider confidence intervals for the Omicron variant are driven by greater uncertainties from the dynamical correction, which is larger for Omicron because of higher absolute growth rates.
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_incubation.ggp.pdf}
-\caption{
-\textbf{Observed and corrected differences in incubation-period distributions of Delta and Omicron variants.}
-(A) Posterior median estimates of the observed (backward) incubation periods of the Delta (black) and Omicron (orange) variants by \cite{backer2021omicron}.
-(B) Forward incubation-period distributions assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta (black) and Omicron (orange) variants, respectively.
-(C--D) Corrected estimates of the mean forward incubation-period for different assumptions about the growth rates of the Delta (C) and Omicron variants (D).
-Lines represent median estimates.
-Shaded regions represent the corresponding 95\% confidence intervals.
-\label{fig:incubation}
-}
-\end{figure}
-
-We can use these estimates of the forward incubation-period distributions to estimate the forward generation-interval distributions.
-For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13--26 December, 2021).
-For within-household transmission pairs (\fref{serial}A), the Omicron variant has shorter mean serial interval (3.1 days; 95\% CI: 2.9--3.3 days) than that of the Delta variant (3.7 days; 95\% CI: 3.5--3.8 days).
-When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a slightly larger difference (\fref{serial}B):
-3.0 days (95\% CI: 2.7--3.2 days) for the Omicron variant and 3.8 days (95\% CI: 3.7--4.0 days) for the Delta variant.
-Our estimate of this difference in these mean generation intervals is robust across plausible ranges of assumptions about the growth rates of the variants (\fref{serial}C--D).
-Assuming lower values of the correlation $\rho$ between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
-In particular, the generation-interval estimates of the Omicron variant are more sensitive to the assumed values of $\rho$ due to faster changes in incidence of infection---for example, changing $\rho$ from 0.85 to 0.5 changes the mean generation-interval estimates for the Omicron variant from 3.1 days (95\% CI: 2.8--3.3 days) to 2.7 days (95\% CI: 2.5--2.9 days).
-We explore a wide range of $\rho$ to consider the possibility that our original assumption ($\rho=0.75$) may under or over-estimate the true $\rho$.
-
-Similar pictures arise for between-household transmission pairs, but the differences in mean serial intervals are unclear (\fref{serial}E): 3.0 days (95\% CI: 2.7--3.3 days) for the Omicron variant and 3.3 days (95\% CI: 3.0 days--3.6 days) for the Delta variant.
-Consistent with the original study, which also reported shorter mean serial intervals for between-household pairs \citep{backer2021omicron}, we estimate shorter mean generation intervals for between-household Delta pairs.
-While the difference in mean generation intervals is larger, there is greater uncertainty in their mean estimates (\fref{serial}F): 2.9 days (95\% CI: 2.5--3.3 days) for the Omicron variant and 3.5 days (95\% CI: 3.2--3.8 days) for the Delta variant.
-Once again, these patterns are robust across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants (\fref{serial}G--H).
-
-In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (13--19 December 2021 and 20--26 December 2021).
-While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all strata, except for within-household transmission pairs during week 50 (13--19 December 2021).
-We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant;
-this decrease in the mean generation interval is likely associated with the lockdown.
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_compare_main.ggp.pdf}
-\caption{
-\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants.}
-(A, E) Observed and fitted forward serial-interval distributions for within-household (A) and between-household (E) transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
-Serial intervals are calculated for infectors who developed symptoms on weeks 50 and 51 (13--26 December, 2021).
-Points represent the observed data.
-Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant.
-(B, F) Estimated forward generation-interval distributions for within-household (B) and between-household (F) transmission pairs in the Netherlands.
-(C, D, G, H) Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta (C, G) and Omicron variants (G, H) for within-household (C, D) and between-household (G, H) transmission pairs.
-Lines represent maximum likelihood estimates.
-Shaded regions represent the corresponding 95\% confidence intervals.
-\label{fig:serial}
-}
-\end{figure}
-
-Accounting for differences in the generation-interval distributions, we estimate that the instantaneous reproduction number of the Omicron variant decreased from 1.73 (95\% CI: 1.59--1.89) to 1.14 (95\% CI: 1.00--1.32) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
-On the other hand, the instantaneous reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
-We estimate the reproduction advantage (i.e., the ratio between the instantaneous reproduction numbers of the Omicron and Delta variants) stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
-However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore the reproduction advantage (\fref{reproduction}B).
-In this case, the reproduction advantage decreases from 2.38 (95\% CI: 2.13--2.67) to 1.43 (95\% CI: 1.17--1.75), corresponding to roughly 4--13\% bias.
-Using between-household generation intervals also gives similar conclusions about changes and biases in the reproduction number estimates (Supplementary Figure S3).
-
-In both cases, the decrease in the reproduction advantage coincides with the decrease in the reproduction number of the Omicron variant, implying that epidemiological changes driving the dynamic had larger effects on the transmission of the Omicron variant than on the transmission of Delta variant;
-a larger reduction in the reproduction number of the Omicron variant also caused its growth rate to decrease faster, causing changes in the observed growth-rate difference shown earlier (\fref{epidemic}D).
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{figure/figure_reproduction_advantage.ggp.pdf}
-\caption{
-\textbf{Estimated instantaneous reproduction number advantages of the Omicron variant.}
-(A) Estimated instantaneous reproduction numbers and their ratios over time while accounting for differences in the generation-interval distributions.
-(B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
-The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
-The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the forward generation-interval distributions for within-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
-Purple lines represent the ratio between the effective reproduction numbers of the Delta and Omicron variants.
-Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
-\label{fig:reproduction}
-}
-\end{figure}
-
-\section{Discussion}
-
-We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands in late 2021 and early 2022.
-The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
-Accounting for differences in epidemic growth rates, however, we find similar incubation-period distributions for both variants but a shorter (by 0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
-Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January.
-Improving generation-interval estimates by taking dynamical effects into account may improve understanding of epidemic dynamics and control measures.
-
-The generation-interval distribution describes changes in the individual-level transmission dynamics over the course of infection and therefore provides crucial information for epidemic control.
-A few studies have estimated the generation-interval distributions of SARS-CoV-2 infections from serial-interval data, but most of them neglect the effects of epidemic growth rates \citep{ganyani2020estimating,he2020temporal,zhao2021estimating,hart2022generation}---these practices can be largely attributed to historical work that concluded that serial and generation intervals have the same means based on the assumption that infectors and infectees have identical incubation-period distributions \citep{svensson2007note,britton2019estimation,lehtinen2021relationship}.
-We build on newer work \citep{park2021forward}, which demonstrated theoretically that forward serial-interval distributions depend on epidemic growth rates, and further confirm that estimates of the forward generation-interval distributions are indeed sensitive to epidemic growth rates.
-These effects are also pertinent to epidemiological inferences of past events from a cohort of infected individuals who experienced a later event at the same time---this includes inferences of other delay distributions, such as incubation-period distributions, as well as viral load trajectories \citep{hay2021estimating}.
-Our sensitivity analysis also shows that the assumptions about the correlation between incubation periods and generation intervals can also have important effects on the estimates of the generation-interval distributions (Supplementary Figure S1).
-
-This study presents a method for accounting for dynamical biases when inferring incubation-period distributions based on epidemic growth rates.
-Observed incubation-period distributions based on symptom-based cohorts are generally expected to be biased, and similar kinds of corrections will be necessary to accurately estimate the incubation-period distribution.
-We note that making these kinds of corrections will also depend on data availability, model complexity, and other epidemiological covariates affecting incubation periods, such as vaccine statuses. 
-Accounting for different sources of biases is critical to accurately estimating incubation-period distributions (and other epidemiological distributions alike) but will necessarily increase uncertainties in the estimates.
-On the other hand, it is still possible to characterize the forward incubation-period distributions without making growth-rate-based corrections through a careful cohorting of individuals with similar infection times when detailed information about infection time is available---we were not able to explore this in our analysis because we relied on publicly available information, which do not contain individual-level information, such as exposure or symptom onset dates.
-
-A few studies have suggested that the incubation period of the Omicron variant may be shorter than that of the Delta variant.
-The median estimates of the Omicron incubation period typically range between 3--4 days, consistent with earlier findings of \citep{backer2021omicron}. 
-However, these data were collected when the number of Omicron infections was growing rapidly \citep{jansen2021investigation,song2022serial}, suggesting that they may have been subject to similar biases.
-On the other hand, incubation-period estimates based on individuals who were exposed from the same event are likely more reliable (because they look forward in time):
-\cite{brandal2021outbreak} estimated the median incubation period of the Omicron variant to be 3 days among those who attended the same holiday party ($n=117$) on 26 November 2021 in Norway. 
-However, we cannot rule out the possibility that some of these attendees were infected prior to the party given that some individuals had COVID-like symptoms prior to the party with at least 96 of the attendees sharing offices; neglecting these factors can lead to underestimation of the mean incubation period.
-Systematic comparisons of data collection methods and epidemiological contexts are needed to properly assess the differences in incubation period distributions of the Delta and Omicron variants.
-
-Some studies have also estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
-\cite{ito2022estimating,selby2022generation} estimated the generation-interval distributions of the Omicron variant but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
-Here, we estimate a shorter mean \emph{realized} generation interval for the Omicron variant.
-The realized generation intervals represent time between actual infection events, and are different from the intrinsic generation intervals, which reflect the average profile of infectiousness of infected individuals over the course of their infections \citep{champredon2015intrinsic}.
-
-Shorter realized generation intervals of the Omicron variant may be driven, in part, by shorter intrinsic generation intervals.
-For example, faster within-host clearance of the Omicron variant \citep{hay2022viral} may lead to faster recovery, which could in turn shorten the intrinsic generation interval \citep{roberts2007model}.
-Pre-existing immunity has also typically been associated with faster recovery \citep{kissler2021viral} (and therefore, shorter generation intervals).
-In our analysis, a greater proportion of Omicron than Delta infections would have been in people with pre-existing immunity (because these people are less likely to be infected by Delta);
-this allowed the Omicron variant to rapidly replace the previously dominant Delta variant at the population level.
-However, the individual-level relationship between immunity and the intrinsic generation-interval distributions of the Omicron variant is difficult to predict given its high immune evasiveness:
-some individuals may have been able to elicit a strong immune response, thereby recovering faster, but many other individuals likely experienced a full course of infection.
-In other words, pre-existing immunity may shorten intrinsic generation intervals, whereas immune evasion may lengthen them (relative to reinfection with Delta, which would be less immune evasive).
-Their combined effects on the resulting generation-interval distribution are unclear.
-We further note that the population-level effects of immune evasiveness, which allowed the Omicron variant to invade rapidly, do not depend on these possible changes in the generation-interval distribution:
-immune evasion could have allowed Omicron to replace Delta even without differences in intrinsic generation-interval distributions.
-
-Even if the Delta and Omicron variants had identical intrinsic generation-interval distributions (e.g., a lack of changes in viral kinetics or the effect of pre-existing immunity), their realized generation-interval distributions could still differ.
-For example, if there are more stringent contact tracing measures against the Omicron variant, individuals who are infected with the Omicron variant would transmit for a shorter amount of time, thereby resulting in a shorter realized generation interval.
-Although this was the case in the Netherlands \citep{backer2021omicron}, targeted control measures against a specific variant are likely to have a small impact on the overall generation-interval distribution given that variants are usually identified for a small fraction of total infections.
-If Delta and Omicron variants have different degrees of symptomaticity or different incubation periods, behavioral changes after symptom onset, such as self-isolation, could also lead to shorter realized generation intervals.
-However, this explanation is also less likely given similarities in the inferred incubation-period distributions.
-
-Instead, we tentatively hypothesize that the differences in realized generation intervals may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction number of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from being realized.
-To illustrate this effect, consider a scenario in which an individual infected with variant A can infect one person per day for four days whereas another individual infected with variant B can infect two people per day for four days---in this case, both variants have identical intrinsic generation-interval distributions.
-We note that this scenario does not match our model, which assumes time-varying force of infection, not a deterministic number of infections; nonetheless, this scenario provides a simple example for explaining the network effect.
-Under the same scenario, if each individual closely interacted with four other contacts, the individual infected with variant A will infect one person every day; in contrast, an individual infected with variant B will infect two people per day and no longer transmit (effectively) after the first two days, leading to shorter realized generation intervals.
-Previous simulations showed that such network effects can have considerable impact on realized generation intervals \citep{park2020inferring}.
-While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
-
-Our study indicates that the Omicron variant has a shorter mean realized generation interval than that of the Delta variant, but the difference between the intrinsic infectiousness profiles remains unclear.
-In particular, similarities in the incubation-period distributions of the Delta and Omicron variants suggest that the differences in their infectiousness profiles may be smaller than the estimated differences in their realized generation-interval distributions.
-In addition, the counter-factual initial intrinsic generation intervals (for an immunologically naive population) of both variants are likely longer than what we estimate given existing levels of interventions, including vaccination, and pandemic awareness---estimating initial intrinsic generation-interval distributions of SARS-CoV-2 variants is expected to be a difficult problem as it requires data from times when awareness levels were low \citep{sender2021unmitigated}.
-Nonetheless, estimates of realized generation-interval distributions describe current epidemic dynamics, implicitly accounting for intervention and behavioral effects,
-and can therefore be expected to improve estimates of effective reproduction numbers.
-
-Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
-In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
-More generally, the bias in inferring the reproduction advantage of an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
-For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant (\textgreater\ 4-fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
-With our generation-interval estimates, we estimate a 2.6-fold reproduction advantage for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the assumptions and results of \cite{pearson2021bounding} for Gauteng province in South Africa.
-
-We considered two ways of measuring transmission advantages: growth-rate differences and reproduction advantage.
-Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
-On the other hand, the growth-rate difference is easier to estimate in real time, and is also more directly relevant to short-term dynamics. 
-For example, when two strains have the same $\RR>1$, the one with shorter generation intervals will grow faster and become dominant as long as $\RR>1$;
-however, when $\RR$ is reduced below 1 (either due to intervention or susceptible depletion), the one with longer generation interval will grow faster.
-These transmission advantages are captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
-Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.
-
-There are several limitations to our analysis.
-First, we primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
-In doing so, we implicitly assume that the delay between infection and reports is fixed.
-However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
-While this limitation does not affect our generation-interval estimates, our inferences of the transmission advantages of the Omicron variant should be interpreted with care.
-
-We assume a constant growth rate for each variant throughout our analysis.
-During the study period, growth rates of both the Delta and Omicron variants changed slowly, and therefore our constant-growth-rate assumption provides a reasonable approximation for their dynamics across two weeks.
-However, this assumption might be problematic when growth rates are changing rapidly (e.g., due to an introduction of stringent control measures) or if the sampling window is too wide.
-Extending our framework to account for time-varying growth rates is relatively simple when inferring the forward incubation-period distribution from the corresponding backward distribution---we can simply replace $r$ with $r(t)$ in \eref{forward} because the backward incubation-period distribution is a weighted average of the forward incubation-period distributions and the number of individuals in each cohort (i.e., individuals who were infected at the same time).
-However, such extensions are more complicated for linking generation- and serial-interval distributions because the forward serial-interval distribution also depends on the cohort reproduction number---for example, if a certain cohort of infectors had higher reproduction number (e.g., because they were infected before control measures were observed), we are more likely to observe transmission from this cohort (see \citep{park2021forward} for more details).
-Assuming exponential growth allows us to avoid this complexity.
-Extending our framework to account for time-varying growth rates can provide more accurate tools for inferring epidemiological delay distributions.
-
-We assume that incubation periods and generation intervals follow a bivariate lognormal distribution---thus, we implicitly assume a power-law relationship between them. 
-This was done because data were limited, and previous work \cite{sender2021unmitigated} has shown that this approach can also capture important aspects of the relationship.
-In fact, however, previous studies have shown that mechanistic models capture this relationship better \citep{ferretti2020timing,hart2021}. 
-When sufficient data are available, more mechanistic approaches may therefore be more suitable.
-
-We do not account for individual-level heterogeneity, such as age, vaccination status, or previous exposure history.
-In general, epidemic growth rates may differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences can affect estimates of epidemiological delay distributions, including the incubation-period and generation-interval distributions.
-We are not able to perform stratified analyses because individual-level information was not publicly available.
-Therefore, while we estimate unclear differences in incubation-period distributions between Delta and Omicron infections, controlling for other covariates, such as age and immune history, may help better characterize differences in Delta and Omicron infections.
-
-Finally, there are several sources of biases in serial-interval data that we did not consider.
-For example, the direction of transmission is difficult to establish for SARS-CoV-2 due to pre-symptomatic transmission.
-Other sources of information, such as exposure history and positive test results, can help resolve uncertainties but are imperfect.
-Serial-interval data also depend on the ability of infected individuals to accurately recall when their symptoms started.
-Future studies may explore how these biases affect the inference of generation intervals from serial intervals.
-While comparisons of incubation-period and serial-interval distributions can shed insight on pathogen dynamics, both distributions typically do not account for the dynamics of asymptomatic infections; neglecting these differences can further bias estimates of transmissibility of a pathogen \citep{park2020time}.
-
-Monitoring changes in key epidemiological parameters is critical to understanding the evolution of SARS-CoV-2 and predicting its future dynamics \citep{kraemer2021monitoring}.
-Our study synthesizes a previously developed theoretical framework on serial- and generation-interval distributions and presents methodological advances in monitoring epidemiological parameters.
-Similar efforts will be critical to improve estimates of the infectiousness profiles of future SARS-CoV-2 variants, especially among asymptomatically infected individuals.
-These conclusions are relevant for other emerging and endemic pathogens in general.
-
-\jd{Done quickly, needs cleaning (particularly cites).}
-This study also has potential implications for studying generation times in population biology \citep{leslie1966intrinsic}.
-Mean age of reproduction for a cohort of mothers corresponds to the forward generation distribution, while mean age of mothers for a cohort of offspring corresponds to the backward distribution.
-Both could therefore be affected by dynamical considerations.
-Similarly, the connection between generation times based at different recruitment stages is analogous to that between serial and generation intervals \citep{lebreton2005age,bienvenu2015new}.
-Comparing definitions and methods across these fields could provide valuable future insights.
-
-\section*{Data availability}
-
-All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/omicron-generation}).
-
-\section*{Acknowledgements}
-
-We thank Ron Milo for providing helpful comments on the manuscript. 
-JD was supported by the Canadian Institutes of Health Research, the Natural Sciences and Engineering Research Council of Canada, and the Michael G. DeGroote Institute for Infectious Disease Research.
-JSW acknowledges support from the Army Research Office W911NF1910384 and the Ile de France region via the Chaires Blaise Pascal program. 
-JW and JB has received funding from European Union's Horizon 2020 research and innovation programme---project EpiPose, Epidemic intelligence to Minimize COVID-19’s Public Health, Societal and Economical Impact (grant agreement number 101003688).
-SF was supported by the Wellcome Trust (grant no. 210758/Z/18/Z). 
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript. 
-The findings and conclusions in this study are those of the authors and do not necessarily represent the official position of the funding agencies, the National Institutes of Health, or the U.S.
-Department of Health and Human Services.
 
@@
 (B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
-The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
+The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week.
 The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the forward generation-interval distributions for between-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
@@
 
-\pagebreak
-
-\bibliography{omicron.bib}
-
 \end{document}
